### 2017-05-29 main.tex
initial commit

\PassOptionsToPackage{utf8}{inputenc}
\documentclass{bioinfo}
\copyrightyear{2017} \pubyear{2017}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Original Article}

\begin{document}
\firstpage{1}

\subtitle{Genome Analysis}

\title[Dryhic]{Dryhic: Increasing Reproducibility of Hi-C Samples with Abnormal Karyotypes}

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$, François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb 1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina Cuartero\,$^{\text{\sfb 1,2}}$ Guillaume Fillion\,$^{\text{\sfb 1,2}}$ and Miguel Beato\,$^{\text{\sfb 1,2}}$}

\address{$^{\text{\sf 1}}$Gene Regulation, Stem Cells and Cancer Program, Centre for Genomic Regulation (CRG), The
	Barcelona Institute of Science and Technology (BIST), Barcelona, Spain \\
	$^{\text{\sf 2}}$Universitat Pompeu Fabra (UPF), Barcelona, Spain}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{
	\textbf{Motivation:} Several strategies have been proposed to remove biases from Hi-C experiments. However, samples with abnormal karyotypes (such as cancer cell lines) present challenges not completely addressed by existing methods. We introduce an alternative designed to model the total number of contacts per genomic loci, thus taking into account possible chromosomal abnormalities.\\
	\textbf{Results:} The proposed method increases reproducibility among replicas of Hi-C samples with abnormal karyotype and obtained using different protocols. The implementation is fast and scales well in terms of computing resources for resolutions up to 1 Kbp.\\	
	\textbf{Availability:} The strategy proposed is implemented as an R package available at \href{http://www.github.com/qenvio/dryhic}{http://www.github.com/qenvio/dryhic}.\\	
	\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
	\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}	online.}
\maketitle

\section{Introduction}

High throughput chromosomal conformation capture (HiC) experiments allow us to explore the spatial organization of chromatin in cells \citep{lieberman2009comprehensive}. Contrary to other 3C techniques, HiC is unbiased in the sense that all possible pairwise interactions between restriction fragments might be interrogated. However, this does not guarantee absence of bias. For instance, genome features as CG content, restriction enzyme site distribution and sequence uniqueness (related to mappability) have been described to modify the result in such experiments \citep{yaffe2011probabilistic}.

Several strategies have been proposed to remove biases in HiC experiments \citep{schmitt2016genome}, either modeling explicitly a set of genomic features \citep{yaffe2011probabilistic,hu2012hicnorm} or using algorithms to implicitly correct for unknown biases \citep{imakaev2012iterative}, the latter often being preferred due to its fast implementation. These methods are suitable for normal cells but they were not designed for cells with abnormal karyotypes. Although there have been some proposals that consider aneuploidy \citep{wu2016computational}, to our knowledge, there is a lack of methodology dealing with aberrant karyotypes (common in cancer cell lines) that present abnormalities more complex than entire chromosome duplications.

Here we propose \texttt{oned}, an alternative designed to model total number of contacts per genomic loci, thus taking into account possible chromosomal abnormalities. It explicitly models the contribution of known biases via a generalized additive model (GAM). The resulting data present better overall reproducibility between replicas and across different protocols. Besides, the implementation is fast and scales up to 1 Kbp resolution with reasonable computing resources.

%\enlargethispage{12pt}

\begin{methods}

\section{Methods}

\subsection{Model}

A common representation for the data yielded in a HiC experiment is a contact matrix, obtained slicing the genome in contiguous bins of fixed length (resolution) and computing the number of contacts between each pair of bins, corresponding to a cell in the contact matrix ($x_{ij}$).

We assumed that the total number of contacts per bin ($t_{i}$) could be approximated by a negative binomial distribution with every known genomic feature contributing independently to its mean ($\lambda_i$). Given that this relationship might not be linear (Figure \ref{fig:totals}A), we allowed a smooth representation using thin plate penalized regression splines \citep{wood2003thin} in a generalized additive model \citep{wood2011fast}. The model can be parametrized as follows:
$$
t_i = \sum_{i}{x_{ij}}
$$
$$
t_i \sim NB(\lambda_i, \theta) \\
$$
$$
log(\lambda_i) \propto \sum_{K}{f_k(z_k)}
$$
where $x_{ij}$ are the number of contacts between bins $i$ and $j$, and there is an additive term for each $z_k$ genomic feature. The smooth functions $\{f_k(z_k)\}$ are estimated jointly with the negative binomial dispersion parameter $\theta$ using R \citep{coreteam2014r} \texttt{mgcv} package \citep{wood2011fast}.

Once the model of the total counts has been estimated, we can further rescale  $\{\lambda_i\}$ to obtain a correction vector $\{\lambda_i'\}$ that can be used to compute the corrected counts ($c_{ij}$) or, alternatively, it can be included in downstream analysis:
$$
\lambda_i' = \frac{\lambda_i}{\sum_j^n{\lambda_j}/n}
$$
$$
c_{ij} = \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}
$$

Following  previous works \citep{yaffe2011probabilistic,hu2012hicnorm} we included mappability, CG content and number of restriction sites (comparable to effective length) as genomic features, but the model and implementation allow easy extension with different genomic features.

\begin{figure}[!tpb]
	\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
	\caption{
		A Non linear relationship between a genomic feature (number of restriction sites per bin) and the total number of contacts per bin for a T47D sample. B Total number of contacts per bin in chromosome 17 of a T47D sample for raw data (in brown, noisier) and corrected data (in blue, preciser).
	}\label{fig:totals}
\end{figure}

\subsection{Data sources}

As an environment to test the bias correction, we gather a set of published \citep{ledily2014distinct,encode2012integrated,rao20143d,stadhouders2017transcription,lin2012global,dixon2012topological} and unpublished data from HiC experiments of different cell types and organisms. The details can be found in Table \ref{tab:samples}. Briefly, we used several experiments comprising T47D breast cancer cell lines (7 samples), K562 leukemia cell lines (4 samples), both with aberrant karyotypes and mouse primary B (6 samples )and ES (7 samples) cells, both with normal karyotypes. Experiments used different protocols (\textit{i.e.} original HiC \citep{lieberman2009comprehensive} and in-situ HiC \citep{rao20143d}), different restriction enzymes (DpnII, HindIII, MboI and NcoI) and produced in distinct laboratories.

We also make use of array-based copy-number segmentation of the two cell lines obtained from COSMIC database \citep{forbes2010cosmic}.

\subsection{Data processing}

All data were processed though a in-house pipeline based on TADBit \citep{serra2016structural}. For more details, see methods of \cite{stadhouders2017transcription}. The pipeline is available, with minor modifications, in \cite{quilez2017managing} material. Briefly, after fastq quality control, paired reads were mapped against the corresponding reference genome (hg38 and mm10) taking into account the restriction enzyme motifs. Non-informative contacts were removed applying the following TADBit \citep{serra2016structural} filters: Self-circle, dangling-ends, error, extra dangling-end, duplicated and random-breaks. Then, we developed the routines contained in the \texttt{dryhic} R package to efficiently create sparse representations of contact matrices (at 100 Kb resolution) and further apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections. \texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp} \citep{wu2016computational} were used to get \textit{lgf} and \textit{caib} corrections respectively.

\subsection{Total number of contacts}

In non-diploid samples, we expect the coverage to be directly related to the number of copies (a duplicated region will show, on average, twice the number of contacts). Given that our model is based on total number of contacts, which in turn is directly related to the number of copies, we reasoned that a preliminary test would be to check if the corrected number of contacts per bin ($ct_i = \sum_{i}{c_{ij}}$) reflects the copy number (as measured by an independent technique such as array-based copy-number segmentation).

\subsection{Sample comparison}

In order to test performance, we relied on pair-wise comparisons of contact matrices. However, there is little consensus on how to compare two HiC matrices. The simplest metric is the Spearman correlation applied to cis (intra-chromosomal) contacts up to a given distance (\textit{i.e.} 5 Mb). Though, the decay in the number of contacts with increasing genomic distance is the main driver of contact abundance and it could mask real differences between samples. This distance decay can be taken into account, for instance, computing the expected number of interactions given the genomic distance between any pair of loci. Thus, we can compare matrices in terms of observed over expected contacts, via the Pearson correlation (again focusing on a given distance range). Alternatively, we can compute the correlation per distance stratum and then obtain a stratum-adjusted correlation coefficient (SCC) as defined in \cite{yang2017hicrep}. Additionally, \cite{yan2017hicspector} proposed a reproducibility score based on the correlation between a set of the Laplacian matrix's eigenvectors (corresponding to the smallest eigenvalues) for the two experiments.

We defined three sets of samples: One with all T47D experiments plus two K562 samples, another with all K562 experiments plus two T47D samples and a third one comprising all mm10 samples (see Table \ref{tab:samples} for details). Given a set of experiments and a metric, we first computed all pair-wise combinations between experiments and then classified the comparisons according to the characteristics of each pair (\textit{i.e.} same / different cell types, same / different protocols, etc ...).

To benchmark our method (\textbf{\textit{oned}}), we also test the performance of other alternatives. Three of them were based on the same implicit iterative correction \citep{imakaev2012iterative}: One computing one iteration (\textbf{\textit{vanilla}}), another iterating until convergence (\textbf{\textit{ice}}) and a third one using the chromosome adjusted iterative correction (\textbf{\textit{caib}}) \citep{wu2016computational}. The fourth one was inspired on the explicit bias correction originally presented by \cite{yaffe2011probabilistic}, but based on the local genomic features (\textbf{\textit{lgf}}) approach introduced by \cite{hu2012hicnorm} as implemented in \cite{servant2012hitc}.

Besides, we compared the raw matrices and used these comparisons as a baseline to present the gain / loss of correlation relative to the uncorrected data, thus allowing us to verify the effect of the bias correction procedure. The differences relative to the raw comparisons were estimated using a linear mixed model with the method as fixed effect and the chromosome as a random effect, and using the \texttt{lmer} function of the \texttt{lme4} package \citep{bates2015lme4}.

In an attempt to summarize the results, we reasoned that the correlation between unrelated samples (\textit{i.e} different cell types) should be lower than between related samples (\textit{i.e} same cell types), so we could re-interpret each correlation metric as a classifier score. Thus, we computed the receiver operating characteristic (ROC) curves and corresponding area under the curve (AUC) using the \texttt{ROCR} package \citep{sing2005rocr}.

\end{methods}

\section{Results}

\subsection{Copy number}

Figure \ref{fig:totals}B shows the total number of corrected contacts along chromosme 17 of a T47D sample. Our correction approach presents a more stable and precise estimation of the copy number. Quantitatively (Figure \ref{fig:copy_number}), Pearson correlation with independent copy number data is high for both T47D and K562 cell lines (averages of 0.87 and 0.78 respectively), not only compared with raw data (averages of 0.74 and 0.63 respectively) but also with any of the alternative methods (averages ranging 0.39-0.68 and 0.30-0.53 respectively). Similar results were obtained using Spearman correlation (Supplementary Figure 2).

\begin{figure}
	\centerline{\includegraphics[width=.45\textwidth]{img/copy_number_figure2.pdf}}
	\caption{
		Pearson correlation between total number of contacts per bin and independent copy number estimation (COSMIC) for each of the methods compared. Left panel T47D breast cancer cell line, right panel K562 leukemia cell line. The new proposal (in blue) outperforms the rest of alternatives.
	}\label{fig:copy_number}
\end{figure}

\subsection{Samples with aberrant karyotypes}

Focusing on the Pearson correlation of observed over expected counts, methods based on the iterative correction (mainly \textit{ice} and \textit{caib} but also \textit{vanilla}) present the undesirable feature of higher correlations between unrelated samples than using raw data (top left panels of Figures \ref{fig:aberrant}A and \ref{fig:aberrant}C, Supplementary Figure 3). Among the comparisons on cells of the same type, \textit{oned} is the only method outperforming uncorrected data when comparing experiments using different protocols (top right panel of Figure \ref{fig:aberrant}A) and it is among the top performant when comparing experiments using the same protocol (rest of panels of Figures \ref{fig:aberrant}A and \ref{fig:aberrant}C). Consequently, our proposal is the one that better discriminates unrelated samples form similar ones (as it can be seen in the ROC curves and AUCs of Figures \ref{fig:aberrant}B and D). Complementary analysis using other metrics (Supplementary Figures 3, 4 and 5) yielded similar results. Notably, Spearman correlation of contacts presents the worst performance for all scenarios and seems to be a poor choice as a metric to compare HiC matrices.

\begin{figure}
	\centerline{\includegraphics[width=.50\textwidth]{img/correlation_aberrant_figure3.pdf}}
	\caption{
		Results of the comparison between samples with aberrant karyotype. Left panels: Average difference between each correction method (Y axis) and the raw data (brown dashed line) with 95\% CI for T47D (A) and K562 (C) sets. Right panels: Corresponding ROC curves and AUCs for T47D (B) and K562 (D) sets. All results in this figure are based on Pearson correlations between the observed over expected counts. Our method increases reproducibility of similar samples while being able to discriminate different ones.
	}\label{fig:aberrant}
\end{figure}

\subsection{Samples with normal karyotypes}

Regarding cells with diploid genomes, we confirmed that our bias correction strategy also achieves good results in terms separation between unrelated and similar samples (Figure \ref{fig:normal} and Supplementary Figures 6 and 7), increasing the reproducibility across different experimental protocols for similar primary B and ES mouse cell samples. Again, Spearman correlation of contacts revealed as the worst metric.

\begin{figure}
	\centerline{\includegraphics[width=.50\textwidth]{img/correlation_normal_figure4.pdf}}
	\caption{
		Results of the comparison between samples with normal karyotype. A Average difference between each correction method (Y axis) and the raw data (brown dashed line) with 95\% CI for the mm10 sets. B Corresponding ROC curves and AUCs. The introduced strategy do not under perform compared to the alternatives.
	}\label{fig:normal}
\end{figure}

\subsection{Computing resources}

One of the main reasons for the broad usage of the iterative correction \citep{imakaev2012iterative} is the easy and fast implementation, moreover compared to existing explicit methods \citep{servant2012hitc}. Our proposal uses only the total number of contacts per bin instead the full matrix for the model estimation. Thus, it achieves similar computing times to the fastest approaches (Figure \ref{fig:times} A). Yet it scales with increasing resolutions (decreassing bin size) (Figure \ref{fig:times} B).

\begin{figure}
	\centerline{\includegraphics[width=.35\textwidth]{img/times_sina_figure5.pdf}}
	\caption{
		Computing time (Y axis, log scale) of all the bias methods used. Each dot corresponds to one sample. The only method faster than ours under performs in all sample comparisons.
	}\label{fig:times}
\end{figure}

\section{Discussion and Conclusions}

\section*{Acknowledgements}

\section*{Funding}

This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}

\bibliographystyle{natbib}
\bibliography{references}

\end{document}

### 2017-05-31 main.tex
Inproving manuscript

@@
 
-Several strategies have been proposed to remove biases in HiC experiments \citep{schmitt2016genome}, either modeling explicitly a set of genomic features \citep{yaffe2011probabilistic,hu2012hicnorm} or using algorithms to implicitly correct for unknown biases \citep{imakaev2012iterative}, the latter often being preferred due to its fast implementation.
-These methods are suitable for normal cells but they were not designed for cells with abnormal karyotypes.
+Several strategies have been proposed to remove biases in HiC experiments \citep{schmitt2016genome}, either modeling explicitly an a-priori set of known genomic features as potential sources of bias \citep{yaffe2011probabilistic,hu2012hicnorm} or using algorithms to implicitly correct for unknown biases \citep{imakaev2012iterative}.
+Notably, the latter has a broader usage due to its fast implementation.
+Both types of methods are suitable for normal cells but they were not designed for cells with abnormal karyotypes.
 Although there have been some proposals that consider aneuploidy \citep{wu2016computational}, to our knowledge, there is a lack of methodology dealing with aberrant karyotypes (common in cancer cell lines) that present abnormalities more complex than entire chromosome duplications.
 
-Here we propose \texttt{oned}, an alternative designed to model total number of contacts per genomic loci, thus taking into account possible chromosomal abnormalities.
+Here we propose \textit{oned}, an alternative designed to model total number of contacts per genomic loci, thus taking into account possible chromosomal abnormalities.
 It explicitly models the contribution of known biases via a generalized additive model (GAM).
-The resulting data present better overall reproducibility between replicas and across different protocols.
+The resulting data present better overall reproducibility between replicas and across different protocols, specially, but not restricted to, in the presence of aberrant karyotypes.
 Besides, the implementation is fast and scales up to 1 Kbp resolution with reasonable computing resources.
@@
 
-A common representation for the data yielded in a HiC experiment is a contact matrix, obtained slicing the genome in contiguous bins of fixed length (resolution) and computing the number of contacts between each pair of bins, corresponding to a cell in the contact matrix ($x_{ij}$).
+A common representation for the data yielded in a HiC experiment is a contact matrix, obtained slicing the genome in contiguous bins of fixed length (resolution) and computing the number of contacts between each pair of bins, corresponding to a cell in the contact matrix ($x_{ij}$), providing a quantification of the interaction between the tow loci.
 
-We assumed that the total number of contacts per bin ($t_{i}$) could be approximated by a negative binomial distribution with every known genomic feature contributing independently to its mean ($\lambda_i$).
-Given that this relationship might not be linear (Figure \ref{fig:totals}A), we allowed a smooth representation using thin plate penalized regression splines \citep{wood2003thin} in a generalized additive model \citep{wood2011fast}.
+Our proposal is based on the overall contacts registered for each bin, thus reducing the matrix to one dimension (hence the name).
+We assumed that the total number of contacts per bin ($t_{i}$) could be approximated by a negative binomial distribution with every potentially biasing genomic feature (known a priori) contributing independently to its mean ($\lambda_i$).
+Given that this relationship might not be linear (as can be seen in Figure \ref{fig:totals}A), we allowed a smooth representation using thin plate penalized regression splines \citep{wood2003thin} in a generalized additive model \citep{wood2011fast}.
 The model can be parametrized as follows:
@@
 where $x_{ij}$ are the number of contacts between bins $i$ and $j$, and there is an additive term for each $z_k$ genomic feature.
-The smooth functions $\{f_k(z_k)\}$ are estimated jointly with the negative binomial dispersion parameter $\theta$ using R \citep{coreteam2014r} \texttt{mgcv} package \citep{wood2011fast}.
+The smooth functions $\{f_k(z_k)\}$ are estimated jointly with the negative binomial dispersion parameter $\theta$ using R \citep{coreteam2014r} \texttt{mgcv} package \citep{wood2011fast}.%should we parametrize in terms of gamma-poisson instead negative binomial???
 
@@
 
-Following  previous works \citep{yaffe2011probabilistic,hu2012hicnorm} we included mappability, CG content and number of restriction sites (comparable to effective length) as genomic features, but the model and implementation allow easy extension with different genomic features.
+Following  previous works \citep{yaffe2011probabilistic,hu2012hicnorm} we included mappability, CG content and number of restriction sites (comparable to effective length) as genomic features, but the model and implementation allow easy extension with any available genomic feature.
 
@@
 		A Non linear relationship between a genomic feature (number of restriction sites per bin) and the total number of contacts per bin for a T47D sample.
-B Total number of contacts per bin in chromosome 17 of a T47D sample for raw data (in brown, noisier) and corrected data (in blue, preciser).
+B Total number of contacts per bin in chromosome 17 of a T47D sample.
+Raw (in brown, noisier) and corrected (in blue, more precise) data are depicted.
 	}\label{fig:totals}
@@
 The details can be found in Table \ref{tab:samples}.
-Briefly, we used several experiments comprising T47D breast cancer cell lines (7 samples), K562 leukemia cell lines (4 samples), both with aberrant karyotypes and mouse primary B (6 samples )and ES (7 samples) cells, both with normal karyotypes.
-Experiments used different protocols (\textit{i.e.} original HiC \citep{lieberman2009comprehensive} and in-situ HiC \citep{rao20143d}), different restriction enzymes (DpnII, HindIII, MboI and NcoI) and produced in distinct laboratories.
+Briefly, we used several experiments comprising T47D breast cancer cell lines (7 samples), K562 leukemia cell lines (4 samples), both with aberrant karyotypes, and mouse primary B (6 samples ) and ES (7 samples) cells, both with normal diploid karyotypes.
+The experiments used different protocols (\textit{i.e.} original HiC \citep{lieberman2009comprehensive} and in-situ HiC \citep{rao20143d}), different restriction enzymes (DpnII, HindIII, MboI and NcoI) and they were produced in distinct laboratories.
 
@@
 
-All data were processed though a in-house pipeline based on TADBit \citep{serra2016structural}.
-For more details, see methods of \cite{stadhouders2017transcription}.
-The pipeline is available, with minor modifications, in \cite{quilez2017managing} material.
+All data were processed though an in-house pipeline based on TADBit \citep{serra2016structural}.
 Briefly, after fastq quality control, paired reads were mapped against the corresponding reference genome (hg38 and mm10) taking into account the restriction enzyme motifs.
 Non-informative contacts were removed applying the following TADBit \citep{serra2016structural} filters: Self-circle, dangling-ends, error, extra dangling-end, duplicated and random-breaks.
-Then, we developed the routines contained in the \texttt{dryhic} R package to efficiently create sparse representations of contact matrices (at 100 Kb resolution) and further apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections.
+For more details, see methods of \cite{stadhouders2017transcription}.
+In addition, the pipeline is available, with minor modifications, in \cite{quilez2017managing} material.
+
+We developed the routines contained in the \texttt{dryhic} R package to efficiently create sparse representations of contact matrices and further apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections.
 \texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp} \citep{wu2016computational} were used to get \textit{lgf} and \textit{caib} corrections respectively.
+All results are based on 100 Kbp resolution, although no major differences were found using different ones (data not shown).
 
-\subsection{Total number of contacts}
+\subsection{Total number of contacts as copy number proxy}
 
-In non-diploid samples, we expect the coverage to be directly related to the number of copies (a duplicated region will show, on average, twice the number of contacts).
-Given that our model is based on total number of contacts, which in turn is directly related to the number of copies, we reasoned that a preliminary test would be to check if the corrected number of contacts per bin ($ct_i = \sum_{i}{c_{ij}}$) reflects the copy number (as measured by an independent technique such as array-based copy-number segmentation).
+In euploid samples, we expect the coverage of a HiC experiment to be directly related to the number of copies (a duplicated region will show, on average, twice the number of contacts).
+An ilustrative example can be seen in Figure \ref{fig:copy_number}B.
+Given that our model is based on total number of contacts, which in turn is directly related to the number of copies, we reasoned that a preliminary test would be to check if the corrected number of contacts per bin ($ct_i = \sum_{i}{c_{ij}}$) reflects the copy number (as measured by an independent technique such as array-based copy-number segmentation) better than the alternatives.
 
@@
 Alternatively, we can compute the correlation per distance stratum and then obtain a stratum-adjusted correlation coefficient (SCC) as defined in \cite{yang2017hicrep}.
-Additionally, \cite{yan2017hicspector} proposed a reproducibility score based on the correlation between a set of the Laplacian matrix's eigenvectors (corresponding to the smallest eigenvalues) for the two experiments.
+Additionally, \cite{yan2017hicspector} proposed a reproducibility score based on the correlation between a set of the Laplacian matrix's eigenvectors (corresponding to the smallest eigenvalues) of the two experiments.
 
@@
 To benchmark our method (\textbf{\textit{oned}}), we also test the performance of other alternatives.
-Three of them were based on the same implicit iterative correction \citep{imakaev2012iterative}: One computing one iteration (\textbf{\textit{vanilla}}), another iterating until convergence (\textbf{\textit{ice}}) and a third one using the chromosome adjusted iterative correction (\textbf{\textit{caib}}) \citep{wu2016computational}.
+Three of them were based on the same implicit iterative correction \citep{imakaev2012iterative}: A first alternative computing one iteration (\textbf{\textit{vanilla}}), a second one iterating until convergence (\textbf{\textit{ice}}) and a third one using the chromosome adjusted iterative correction (\textbf{\textit{caib}}) \citep{wu2016computational}.
 The fourth one was inspired on the explicit bias correction originally presented by \cite{yaffe2011probabilistic}, but based on the local genomic features (\textbf{\textit{lgf}}) approach introduced by \cite{hu2012hicnorm} as implemented in \cite{servant2012hitc}.
@@
 Besides, we compared the raw matrices and used these comparisons as a baseline to present the gain / loss of correlation relative to the uncorrected data, thus allowing us to verify the effect of the bias correction procedure.
-The differences relative to the raw comparisons were estimated using a linear mixed model with the method as fixed effect and the chromosome as a random effect, and using the \texttt{lmer} function of the \texttt{lme4} package \citep{bates2015lme4}.
+The differences relative to the raw comparisons were estimated using a linear mixed model with the method as fixed effect and the chromosome as a random effect, and using the \texttt{lmer} function of the \texttt{lme4} R package \citep{bates2015lme4}.
 
-In an attempt to summarize the results, we reasoned that the correlation between unrelated samples (\textit{i.e} different cell types) should be lower than between related samples (\textit{i.e} same cell types), so we could re-interpret each correlation metric as a classifier score.
-Thus, we computed the receiver operating characteristic (ROC) curves and corresponding area under the curve (AUC) using the \texttt{ROCR} package \citep{sing2005rocr}.
+In an attempt to further summarize the results, we reasoned that the correlation between unrelated samples (\textit{i.e} different cell types) should be lower than the correlation between related samples (\textit{i.e} same cell types), so we could re-interpret each correlation metric as a classifier score.
+Thus, we computed the receiver operating characteristic (ROC) curves and corresponding area under the curve (AUC) using the R \texttt{ROCR} package \citep{sing2005rocr}.
 
@@
 Focusing on the Pearson correlation of observed over expected counts, methods based on the iterative correction (mainly \textit{ice} and \textit{caib} but also \textit{vanilla}) present the undesirable feature of higher correlations between unrelated samples than using raw data (top left panels of Figures \ref{fig:aberrant}A and \ref{fig:aberrant}C, Supplementary Figure 3).
-Among the comparisons on cells of the same type, \textit{oned} is the only method outperforming uncorrected data when comparing experiments using different protocols (top right panel of Figure \ref{fig:aberrant}A) and it is among the top performant when comparing experiments using the same protocol (rest of panels of Figures \ref{fig:aberrant}A and \ref{fig:aberrant}C).
+Among the comparisons of cells with the same type, \textit{oned} is the only method outperforming uncorrected data when comparing experiments using different protocols (top right panel of Figure \ref{fig:aberrant}A) and it is among the top performant when comparing experiments using the same protocol (rest of the panels in Figures \ref{fig:aberrant}A and \ref{fig:aberrant}C).
 Consequently, our proposal is the one that better discriminates unrelated samples form similar ones (as it can be seen in the ROC curves and AUCs of Figures \ref{fig:aberrant}B and D).
+
 Complementary analysis using other metrics (Supplementary Figures 3, 4 and 5) yielded similar results.
@@
 		Results of the comparison between samples with aberrant karyotype.
-Left panels: Average difference between each correction method (Y axis) and the raw data (brown dashed line) with 95\% CI for T47D (A) and K562 (C) sets.
+Left panels: Average difference and 95\% CI between each correction method (Y axis) and the raw data (brown dashed line) for T47D (A) and K562 (C) sets.
 Right panels: Corresponding ROC curves and AUCs for T47D (B) and K562 (D) sets.
@@
 
-Regarding cells with diploid genomes, we confirmed that our bias correction strategy also achieves good results in terms separation between unrelated and similar samples (Figure \ref{fig:normal} and Supplementary Figures 6 and 7), increasing the reproducibility across different experimental protocols for similar primary B and ES mouse cell samples.
+Regarding cells with diploid genomes, we confirmed that our bias correction strategy also achieves good results in terms separation between unrelated and similar samples (Figure \ref{fig:normal} and Supplementary Figures 6 and 7), increasing the reproducibility across different experimental protocols for primary B and ES mouse cell samples.
 Again, Spearman correlation of contacts revealed as the worst metric.
@@
 		Results of the comparison between samples with normal karyotype.
-A Average difference between each correction method (Y axis) and the raw data (brown dashed line) with 95\% CI for the mm10 sets.
+A Average difference and 95\% CI between each correction method (Y axis) and the raw data (brown dashed line) for the mm10 sets.
 B Corresponding ROC curves and AUCs.
-The introduced strategy do not under perform compared to the alternatives.
+The introduced strategy does not under perform compared to the alternatives.
 	}\label{fig:normal}
@@
 Our proposal uses only the total number of contacts per bin instead the full matrix for the model estimation.
-Thus, it achieves similar computing times to the fastest approaches (Figure \ref{fig:times} A).
-Yet it scales with increasing resolutions (decreassing bin size) (Figure \ref{fig:times} B).
+Thus, it achieves similar computing times that the fastest approaches (Figure \ref{fig:times} A).
+Yet it scales with increasing resolutions (decreasing bin size) (Figure \ref{fig:times} B).
 
@@
 	\caption{
-		Computing time (Y axis, log scale) of all the bias methods used. Each dot corresponds to one sample. The only method faster than ours under performs in all sample comparisons.
+		Computing time (Y axis, log scale) of all the bias correction methods used. Each dot corresponds to one sample. The only method faster than ours under performs in all sample comparisons.
 	}\label{fig:times}
@@
 \section*{Acknowledgements}
+
+We would like to thank all members of the four labs involved in the 4DGenome Synergy project for the fruitful discussions during project meetings.
+EV wants to acknowledge specially the members of Miguel Beato's lab for their insights during lab meetings.
 

### 2017-06-01 main.tex
Wrote the abstract.

@@
 
-\title[Dryhic]{Dryhic: Increasing Reproducibility of Hi-C Samples with Abnormal Karyotypes}
+\title[Dryhic]{OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes}
 
-\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$, François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb 1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina Cuartero\,$^{\text{\sfb 1,2}}$ Guillaume Fillion\,$^{\text{\sfb 1,2}}$ and Miguel Beato\,$^{\text{\sfb 1,2}}$}
+\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$, François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb 1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina Cuartero\,$^{\text{\sfb 1,2}}$,Miguel Beato\,$^{\text{\sfb 1,2}}$ and Guillaume Fillion\,$^{\text{\sfb 1,2}}$}
 
@@
 \abstract{
-	\textbf{Motivation:} Several strategies have been proposed to remove biases from Hi-C experiments.
-However, samples with abnormal karyotypes (such as cancer cell lines) present challenges not completely addressed by existing methods.
-We introduce an alternative designed to model the total number of contacts per genomic loci, thus taking into account possible chromosomal abnormalities.\\
-	\textbf{Results:} The proposed method increases reproducibility among replicas of Hi-C samples with abnormal karyotype and obtained using different protocols.
-The implementation is fast and scales well in terms of computing resources for resolutions up to 1 Kbp.\\
-	\textbf{Availability:} The strategy proposed is implemented as an R package available at \href{http://www.github.com/qenvio/dryhic}{http://www.github.com/qenvio/dryhic}.\\	
+  \textbf{Motivation:} The three-dimensional conformation of genomes is an
+essential component of their biological activity. The advent of the Hi-C
+technology enabled an unprecedented advance in our understanding of genome
+structures. However, Hi-C is subject to systematic biases that can
+compromise downstream analyses. Several strategies have been proposed to
+remove those biases, but none of them addresses the common issue of
+abnormal karyotypes. Many experiments are performed in cancer cell lines,
+which typically harbor large-scale copy number variations that create
+visible defects on the raw Hi-C maps. The consequences of these widespread
+artifacts on the normalized maps have not been investigated so far.\\
+	\textbf{Results:}
+We observed that current normalization methods perform badly in the
+presence of large-scale copy number variations, obscuring biological
+viarations and enhancing batch effects. To address this issue, we
+developed an alternative appoach designed to take into account chromosomal
+abnormalities. The method, called OneD, increases reproducibility among
+replicates of Hi-C samples with abnormal karyotype, thus significantly
+outperforming previous methods.  In the absence of chromosomal
+aberrantions, OneD fared equally well as state-of-the-art methods, making
+it a safe choice for Hi-C normalization. OneD is fast and scales well in
+terms of computing resources for resolutions up to 1 kbp.\\	
+	\textbf{Availability:} OneD is implemented as an R package available at \href{http://www.github.com/qenvio/dryhic}{http://www.github.com/qenvio/dryhic}.\\	
 	\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
@@
 
-This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}
+This work was supported by the Spanish Ministry of Economy and
+Competitiveness, ‘Centro de Excelencia Severo Ochoa 2013-2017’,
+SEV-2012-0208, ERC Synergy Grant 609989 (GF) and...
+\vspace*{-12pt}
 

### 2017-06-01 main.tex
Wrote the first few paragraphs of the introduction.

@@
 
-High throughput chromosomal conformation capture (HiC) experiments allow us to explore the spatial organization of chromatin in cells \citep{lieberman2009comprehensive}.
-Contrary to other 3C techniques, HiC is unbiased in the sense that all possible pairwise interactions between restriction fragments might be interrogated.
-However, this does not guarantee absence of bias.
-For instance, genome features as CG content, restriction enzyme site distribution and sequence uniqueness (related to mappability) have been described to modify the result in such experiments \citep{yaffe2011probabilistic}.
+One of the crown achievements of modern biology was to realize that
+genomes have an underlying three-dimensional structure contributing to
+their activity (CITE). In mammals, this organization plays a key role in
+guiding enhancer-promoter contacts (CITE), in V(D)J recombination (CITE)
+and in X chromosome inactivation (CITE). One of the most significant
+breakthroughs towards this insight was the development of the high
+throughput chromosomal conformation capture (Hi-C) technology, assaying
+chromosomal contacts at a genome-wide scale
+\citep{lieberman2009comprehensive}. Exploring the spatial organization of
+chromatin inside the cell is now common and Hi-C is part of the standard
+molecular biology toolbox (CITE).
 
-Several strategies have been proposed to remove biases in HiC experiments \citep{schmitt2016genome}, either modeling explicitly an a-priori set of known genomic features as potential sources of bias \citep{yaffe2011probabilistic,hu2012hicnorm} or using algorithms to implicitly correct for unknown biases \citep{imakaev2012iterative}.
-Notably, the latter has a broader usage due to its fast implementation.
-Both types of methods are suitable for normal cells but they were not designed for cells with abnormal karyotypes.
-Although there have been some proposals that consider aneuploidy \citep{wu2016computational}, to our knowledge, there is a lack of methodology dealing with aberrant karyotypes (common in cancer cell lines) that present abnormalities more complex than entire chromosome duplications.
+Contrary to the precursor technologies 3C, 4C and 5C \citep{de2012decade},
+Hi-C interrogates all possible pairwise interactions between restriction
+fragments. However, this does not guarantee the absence of bias. For
+instance, local genome features such as the G+C content, the availability
+of restriction enzyme sites and the mappability of the sequencing reads
+have been shown to modify the results \citep{yaffe2011probabilistic}. In
+addition, the general experimental biases such as batch effects and
+protocol variation also apply. It is thus important to normalize
+Hi-C data in order to remove biases and artifacts, so that they are not
+confused with biological signal.
 
-Here we propose \textit{oned}, an alternative designed to model total number of contacts per genomic loci, thus taking into account possible chromosomal abnormalities.
-It explicitly models the contribution of known biases via a generalized additive model (GAM).
-The resulting data present better overall reproducibility between replicas and across different protocols, specially, but not restricted to, in the presence of aberrant karyotypes.
-Besides, the implementation is fast and scales up to 1 Kbp resolution with reasonable computing resources.
+Several methods have been proposed to remove biases in Hi-C experiments
+\citep{schmitt2016genome}. The first strategy is to model biases
+explicitly from a defined set of local genomic features, such as the G+C
+content and the availability of restriction sites. This approach is used
+in the method of \cite{yaffe2011probabilistic} and in Hicnorm by
+\cite{hu2012hicnorm}. The second strategy is to implicitly correct for
+unknown biases by enforcing some regularity on the data. This approach is
+used in the ICE method of \cite{imakaev2012iterative}, where the total
+amount of contacts is imposed to be equal for every bin. ICE is currently
+the most popular method, which is in part due to the speed of the
+available implementation.
+
+Both strategies are suitable for cells with a normal karyotype, but they
+were not designed for cell types with karyotypic aberrations.
+
+Although there have been some proposals that consider aneuploidy
+\citep{wu2016computational}, to our knowledge, there is a lack of
+methodology dealing with aberrant karyotypes (common in cancer cell lines)
+that present abnormalities more complex than entire chromosome
+duplications.
+
+Here we propose \textit{oned}, an alternative designed to model total
+number of contacts per genomic loci, thus taking into account possible
+chromosomal abnormalities. It explicitly models the contribution of known
+biases via a generalized additive model (GAM). The resulting data present
+better overall reproducibility between replicas and across different
+protocols, specially, but not restricted to, in the presence of aberrant
+karyotypes. Besides, the implementation is fast and scales up to 1 Kbp
+resolution with reasonable computing resources.
 

### 2017-06-02 main.tex
Wrote the introduction.

@@
 visible defects on the raw Hi-C maps. The consequences of these widespread
-artifacts on the normalized maps have not been investigated so far.\\
+artifacts on the normalized maps are mostly unexplored.\\
 	\textbf{Results:}
@@
 genomes have an underlying three-dimensional structure contributing to
-their activity (CITE). In mammals, this organization plays a key role in
-guiding enhancer-promoter contacts (CITE), in V(D)J recombination (CITE)
-and in X chromosome inactivation (CITE). One of the most significant
-breakthroughs towards this insight was the development of the high
-throughput chromosomal conformation capture (Hi-C) technology, assaying
-chromosomal contacts at a genome-wide scale
-\citep{lieberman2009comprehensive}. Exploring the spatial organization of
-chromatin inside the cell is now common and Hi-C is part of the standard
-molecular biology toolbox (CITE).
+their activity \citep{rowley2016three}. In mammals, this organization
+plays a key role in guiding enhancer-promoter contacts (CITE), in V(D)J
+recombination (CITE) and in X chromosome inactivation (CITE). One of the
+most significant breakthroughs towards this insight was the development of
+the high throughput chromosomal conformation capture technology (Hi-C),
+assaying chromosomal contacts at a genome-wide scale
+\citep{lieberman2009comprehensive}. Nowadays, exploring the spatial
+organization of chromatin has become a priority in many fields and
+Hi-C has become part of the standard molecular biology toolbox (CITE).
 
@@
 Hi-C interrogates all possible pairwise interactions between restriction
-fragments. However, this does not guarantee the absence of bias. For
-instance, local genome features such as the G+C content, the availability
-of restriction enzyme sites and the mappability of the sequencing reads
-have been shown to modify the results \citep{yaffe2011probabilistic}. In
-addition, the general experimental biases such as batch effects and
-protocol variation also apply. It is thus important to normalize
-Hi-C data in order to remove biases and artifacts, so that they are not
-confused with biological signal.
+fragments. However, this does not guarantee that the method has no bias.
+On the contrary, local genome features such as the G+C content, the
+availability of restriction enzyme sites and the mappability of the
+sequencing reads have been shown to impact the results
+\citep{yaffe2011probabilistic}. In addition, the general experimental
+biases such as batch effects and protocol variation also apply. It is thus
+important to normalize Hi-C data in order to remove biases and artifacts,
+so that they are not confused with biological signal.
 
@@
 in the method of \cite{yaffe2011probabilistic} and in Hicnorm by
-\cite{hu2012hicnorm}. The second strategy is to implicitly correct for
+\cite{hu2012hicnorm}. The second strategy is to implicitly correct
 unknown biases by enforcing some regularity on the data. This approach is
 used in the ICE method of \cite{imakaev2012iterative}, where the total
-amount of contacts is imposed to be equal for every bin. ICE is currently
-the most popular method, which is in part due to the speed of the
-available implementation.
+amount of contacts of every bin is imposed to be the same. ICE is
+currently the most popular method, due in part to its speed.
 
-Both strategies are suitable for cells with a normal karyotype, but they
-were not designed for cell types with karyotypic aberrations.
+Neither of these strategies were designed with regard for cell types with
+karyotypic aberrations, most common in cancer cells in culture.  Hi-C is
+very sensitive to aneuploidy, copy number variations and translocations.
+Actually, these aberrations have so much influence on the outcome that the
+artifacts can be used to re-assemble the target genome
+\citep{korbel2013genome}.  So far the only attempt to address the issue
+was the correction method caICB by \cite{wu2016computational}. However,
+caICB applies a chromosome-wide correction, effectively excluding the
+numerous cases of partial aneuploidy and regional copy number variations.
 
-Although there have been some proposals that consider aneuploidy
-\citep{wu2016computational}, to our knowledge, there is a lack of
-methodology dealing with aberrant karyotypes (common in cancer cell lines)
-that present abnormalities more complex than entire chromosome
-duplications.
-
-Here we propose \textit{oned}, an alternative designed to model total
-number of contacts per genomic loci, thus taking into account possible
-chromosomal abnormalities. It explicitly models the contribution of known
-biases via a generalized additive model (GAM). The resulting data present
-better overall reproducibility between replicas and across different
-protocols, specially, but not restricted to, in the presence of aberrant
-karyotypes. Besides, the implementation is fast and scales up to 1 Kbp
-resolution with reasonable computing resources.
+Here we propose \textit{OneD}, a method to correct local chromosomal
+abnormalities. OneD explicitly models the contribution of known biases via
+a generalized additive model. The normalized data is more reproducible
+between replicates and across different protocols. Importantly, OneD is
+also applicable when cells have a normal karyotype, where it performs as
+well as the best normalization methods. Finally, the implementation is
+faster than ICE and scales up to 1 kbp resolution with reasonable
+computing resources.
 

### 2017-06-02 main.tex
Wrote the methods section.

@@
 
-A common representation for the data yielded in a HiC experiment is a contact matrix, obtained slicing the genome in contiguous bins of fixed length (resolution) and computing the number of contacts between each pair of bins, corresponding to a cell in the contact matrix ($x_{ij}$), providing a quantification of the interaction between the tow loci.
+The most common representation of Hi-C data is a contact matrix, obtained
+by slicing the genome in $n$ consecutive bins of fixed size (the
+resolution) and computing the number of contacts between each pair of
+bins. The values are stored in the cells of the contact matrix ($x_{ij}$),
+quantifying of the interaction between the tow loci at positions $i$ and
+$j$.
 
-Our proposal is based on the overall contacts registered for each bin, thus reducing the matrix to one dimension (hence the name).
-We assumed that the total number of contacts per bin ($t_{i}$) could be approximated by a negative binomial distribution with every potentially biasing genomic feature (known a priori) contributing independently to its mean ($\lambda_i$).
-Given that this relationship might not be linear (as can be seen in Figure \ref{fig:totals}A), we allowed a smooth representation using thin plate penalized regression splines \citep{wood2003thin} in a generalized additive model \citep{wood2011fast}.
-The model can be parametrized as follows:
-$$
-t_i = \sum_{i}{x_{ij}}
-$$
-$$
-t_i \sim NB(\lambda_i, \theta) \\
-$$
-$$
-log(\lambda_i) \propto \sum_{K}{f_k(z_k)}
-$$
-where $x_{ij}$ are the number of contacts between bins $i$ and $j$, and there is an additive term for each $z_k$ genomic feature.
-The smooth functions $\{f_k(z_k)\}$ are estimated jointly with the negative binomial dispersion parameter $\theta$ using R \citep{coreteam2014r} \texttt{mgcv} package \citep{wood2011fast}.%should we parametrize in terms of gamma-poisson instead negative binomial???
+Our approach is to model the tally of contacts for each bin, thus reducing
+the matrix to a one dimension score (hence the name OneD). We assume that
+the total number of contacts per bin ($t_{i}$) can be approximated by a
+negative binomial distribution. This choice is sensible because the amount
+of contacts is a discrete variable and because the negative binomial
+distribution allows for overdispersion. We further assume that the
+explicit sources of bias have independent contributions to the mean of the
+distribution for a given bin ($\lambda_i$).
 
-Once the model of the total counts has been estimated, we can further rescale  $\{\lambda_i\}$ to obtain a correction vector $\{\lambda_i'\}$ that can be used to compute the corrected counts ($c_{ij}$) or, alternatively, it can be included in downstream analysis:
-$$
-\lambda_i' = \frac{\lambda_i}{\sum_j^n{\lambda_j}/n}
-$$
-$$
-c_{ij} = \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}
-$$
+Given that this relationship might not be linear (see for instance
+Figure~\ref{fig:totals}A), we allowed a smooth representation
+using thin plate penalized regression splines \citep{wood2003thin} in a
+generalized additive model \citep{wood2011fast}. The model can be
+parametrized as
 
-Following  previous works \citep{yaffe2011probabilistic,hu2012hicnorm} we included mappability, CG content and number of restriction sites (comparable to effective length) as genomic features, but the model and implementation allow easy extension with any available genomic feature.
+\begin{align*}
+t_i &= \sum_j^n{x_{ij}} \sim  NB(\lambda_i, \theta), \\
+\log(\lambda_i) &\propto \sum_{k}{f_k(z_k)},
+\end{align*}
+
+\noindent
+where $x_{ij}$ is the raw number of contacts between bins $i$ and $j$, and
+$z_k$ is the additive bias of genomic feature $k$. The smooth functions
+$\{f_k(\cdot)\}$ are estimated jointly with the negative binomial
+dispersion parameter $\theta$ using the \texttt{mgcv} package
+\citep{wood2011fast} of the R software \citep{coreteam2014r}.
+
+Once the parameters of the model are estimated, we rescale the estimated
+means $\{\lambda_i\}$ to obtain a correction vector $\{\lambda_i'\}$ that
+can be used to compute the corrected counts ($\hat{x}_{ij}$).
+
+\begin{align*}
+\lambda_i' &= \frac{\lambda_i}{\sum_j^n{\lambda_j}/n} \\
+\hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}.
+\end{align*}
+
+In line with previous methods
+\citep{yaffe2011probabilistic,hu2012hicnorm}, the features
+used to fit the model are the local G+C content, read mappability and the 
+content and number of restriction sites. The model and the implementation
+can be easily extended with any user-provided genomic feature.
 
@@
 \centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
-	\caption{
-		A Non linear relationship between a genomic feature (number of restriction sites per bin) and the total number of contacts per bin for a T47D sample.
-B Total number of contacts per bin in chromosome 17 of a T47D sample.
-Raw (in brown, noisier) and corrected (in blue, more precise) data are depicted.
-	}\label{fig:totals}
+\caption{A. Non linear relationship between the number of restriction
+sites per bin and the total number of contacts per bin in T47D. B. Total
+number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
+corrected (blue) signals are shown. The region that corresponds to the
+first 20 Mbp of the chromosome is present in single copy, explaining that
+the signal is approximately half compared to the rest of the chromosome.
+Also notice that the raw signal is noisier than the corrected one.}
+\label{fig:totals}
 \end{figure}
@@
 
-As an environment to test the bias correction, we gather a set of published \citep{ledily2014distinct,encode2012integrated,rao20143d,stadhouders2017transcription,lin2012global,dixon2012topological} and unpublished data from HiC experiments of different cell types and organisms.
-The details can be found in Table \ref{tab:samples}.
-Briefly, we used several experiments comprising T47D breast cancer cell lines (7 samples), K562 leukemia cell lines (4 samples), both with aberrant karyotypes, and mouse primary B (6 samples ) and ES (7 samples) cells, both with normal diploid karyotypes.
-The experiments used different protocols (\textit{i.e.} original HiC \citep{lieberman2009comprehensive} and in-situ HiC \citep{rao20143d}), different restriction enzymes (DpnII, HindIII, MboI and NcoI) and they were produced in distinct laboratories.
+To test the correction of biases, we gathered a set of
+published \citep{ledily2014distinct, encode2012integrated, rao20143d,
+stadhouders2017transcription, lin2012global, dixon2012topological} and
+unpublished Hi-C data of different cell types and organisms. The details
+can be found in Table \ref{tab:samples}.
 
-We also make use of array-based copy-number segmentation of the two cell lines obtained from COSMIC database \citep{forbes2010cosmic}.
+We used several experiments comprising T47D breast cancer cell
+lines (7 samples), K562 leukemia cell lines (4 samples), both with
+aberrant karyotypes, and mouse primary B (6 samples ) and ES (7 samples)
+cells, both with normal diploid karyotypes. The experiments were carried
+out in different laboratories, following either the original Hi-C protocol
+\citep{lieberman2009comprehensive} or the newer \textit{in situ} version
+\citep{rao20143d}, and using different restriction enzymes (DpnII,
+HindIII, MboI and NcoI).
+
+We also used array-based copy-number segmentation of the two cell
+lines obtained from COSMIC database \citep{forbes2010cosmic}.
 
@@
 
-All data were processed though an in-house pipeline based on TADBit \citep{serra2016structural}.
-Briefly, after fastq quality control, paired reads were mapped against the corresponding reference genome (hg38 and mm10) taking into account the restriction enzyme motifs.
-Non-informative contacts were removed applying the following TADBit \citep{serra2016structural} filters: Self-circle, dangling-ends, error, extra dangling-end, duplicated and random-breaks.
-For more details, see methods of \cite{stadhouders2017transcription}.
-In addition, the pipeline is available, with minor modifications, in \cite{quilez2017managing} material.
+All data were processed though a pipeline based on TADBit
+\citep{serra2016structural}. Briefly, after controlling the quality of
+fastq files, paired-end reads were mapped in the corresponding reference
+genome (hg38 and mm10) taking into account the restriction enzyme site.
+Non-informative contacts were removed applying the following TADBit
+filters: self-circle, dangling-ends, error, extra dangling-end, duplicated
+and random-breaks. For more details, see the methods section of
+\cite{stadhouders2017transcription}. In addition, the pipeline is
+available from the supplementary material published by
+\cite{quilez2017managing}.
 
-We developed the routines contained in the \texttt{dryhic} R package to efficiently create sparse representations of contact matrices and further apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections.
-\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp} \citep{wu2016computational} were used to get \textit{lgf} and \textit{caib} corrections respectively.
-All results are based on 100 Kbp resolution, although no major differences were found using different ones (data not shown).
+We developed the routines contained in the \texttt{oned} R package to
+efficiently create sparse representations of contact matrices and further
+apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections.
+\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
+\citep{wu2016computational} were used to carry out the \textit{lgf} and
+\textit{caib} corrections respectively. All the results are based on a
+resolution of 100 kbp, but we found no major differences for different
+values (not shown).
 
-\subsection{Total number of contacts as copy number proxy}
+\subsection{Comparison of Hi-C matrices}
 
-In euploid samples, we expect the coverage of a HiC experiment to be directly related to the number of copies (a duplicated region will show, on average, twice the number of contacts).
-An ilustrative example can be seen in Figure \ref{fig:copy_number}B.
-Given that our model is based on total number of contacts, which in turn is directly related to the number of copies, we reasoned that a preliminary test would be to check if the corrected number of contacts per bin ($ct_i = \sum_{i}{c_{ij}}$) reflects the copy number (as measured by an independent technique such as array-based copy-number segmentation) better than the alternatives.
+There is no universally accepted standard to compare Hi-C matrices. The
+simplest metric is the Spearman correlation applied to intra-chromosomal
+contacts up to a given distance (5 Mb in what follows). The second is
+to compare the observed over expected contacts via the Pearson correlation
+up to a given distance range. Compared to the first, this metric gives
+more weight to changes occuring away from the diagonal. The third option
+is to compute a correlation per distance stratum and then obtain a
+stratum-adjusted correlation coefficient (SCC) as defined in
+\cite{yang2017hicrep}.  Finally, \cite{yan2017hicspector} proposed a
+reproducibility score based on the correlation between the eigenvectors of
+the Laplacian matrix associated with the smallest eigenvalues. This
+approach borrows the concepts of spectral clustering
+\citep{von2007tutorial} and amounts to comparing high level features of
+the matrix.
 
-\subsection{Sample comparison}
+We defined three data sets to measure experimental reproducibility after
+normalization: The first contained the samples from T47D plus two samples
+from K562, the second contained the samples from K562 plus two samples
+from T47D, the third contained all the mouse samples (see Table
+\ref{tab:samples} for details). Given a set of experiments and a metric,
+we first computed all pairwise combinations between experiments and then
+classified the comparisons according to the characteristics of each pair
+(cell type, protocol, batch, and treatment).
 
-In order to test performance, we relied on pair-wise comparisons of contact matrices.
-However, there is little consensus on how to compare two HiC matrices.
-The simplest metric is the Spearman correlation applied to cis (intra-chromosomal) contacts up to a given distance (\textit{i.e.} 5 Mb).
-Though, the decay in the number of contacts with increasing genomic distance is the main driver of contact abundance and it could mask real differences between samples.
-This distance decay can be taken into account, for instance, computing the expected number of interactions given the genomic distance between any pair of loci.
-Thus, we can compare matrices in terms of observed over expected contacts, via the Pearson correlation (again focusing on a given distance range).
-Alternatively, we can compute the correlation per distance stratum and then obtain a stratum-adjusted correlation coefficient (SCC) as defined in \cite{yang2017hicrep}.
-Additionally, \cite{yan2017hicspector} proposed a reproducibility score based on the correlation between a set of the Laplacian matrix's eigenvectors (corresponding to the smallest eigenvalues) of the two experiments.
+We benchmark \textbf{\textit{oned}}, against methods
+\textbf{\textit{vanilla}}, \textbf{\textit{ice}}
+\citep{imakaev2012iterative}, \textbf{\textit{caib}}
+\citep{wu2016computational} and \textbf{\textit{lgf}}
+\citep{hu2012hicnorm, servant2012hitc}. The first three methods correct
+biases implicitly, whereas the fourth method does it explicitly.
 
-We defined three sets of samples: One with all T47D experiments plus two K562 samples, another with all K562 experiments plus two T47D samples and a third one comprising all mm10 samples (see Table \ref{tab:samples} for details).
-Given a set of experiments and a metric, we first computed all pair-wise combinations between experiments and then classified the comparisons according to the characteristics of each pair (\textit{i.e.} same / different cell types, same / different protocols, etc ...).
-
-To benchmark our method (\textbf{\textit{oned}}), we also test the performance of other alternatives.
-Three of them were based on the same implicit iterative correction \citep{imakaev2012iterative}: A first alternative computing one iteration (\textbf{\textit{vanilla}}), a second one iterating until convergence (\textbf{\textit{ice}}) and a third one using the chromosome adjusted iterative correction (\textbf{\textit{caib}}) \citep{wu2016computational}.
-The fourth one was inspired on the explicit bias correction originally presented by \cite{yaffe2011probabilistic}, but based on the local genomic features (\textbf{\textit{lgf}}) approach introduced by \cite{hu2012hicnorm} as implemented in \cite{servant2012hitc}.
-
-Besides, we compared the raw matrices and used these comparisons as a baseline to present the gain / loss of correlation relative to the uncorrected data, thus allowing us to verify the effect of the bias correction procedure.
-The differences relative to the raw comparisons were estimated using a linear mixed model with the method as fixed effect and the chromosome as a random effect, and using the \texttt{lmer} function of the \texttt{lme4} R package \citep{bates2015lme4}.
-
-In an attempt to further summarize the results, we reasoned that the correlation between unrelated samples (\textit{i.e} different cell types) should be lower than the correlation between related samples (\textit{i.e} same cell types), so we could re-interpret each correlation metric as a classifier score.
-Thus, we computed the receiver operating characteristic (ROC) curves and corresponding area under the curve (AUC) using the R \texttt{ROCR} package \citep{sing2005rocr}.
+To measure the gain or loss of similarity upon normalization, we compared
+raw matrices to obtain a baseline. The differences with this baseline were
+estimated using a linear mixed model fitted with the \texttt{lmer}
+function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
+fixed effect was the normalization method and the random effect was the
+chromosome. Receiver operating characteristic (ROC) curves were then
+computed using \texttt{ROCR} package \citep{sing2005rocr}.
 
 \end{methods}
+
+%% --------------- Results --------------- %%
 
@@
 
-Figure \ref{fig:totals}B shows the total number of corrected contacts along chromosme 17 of a T47D sample.
-Our correction approach presents a more stable and precise estimation of the copy number.
-Quantitatively (Figure \ref{fig:copy_number}), Pearson correlation with independent copy number data is high for both T47D and K562 cell lines (averages of 0.87 and 0.78 respectively), not only compared with raw data (averages of 0.74 and 0.63 respectively) but also with any of the alternative methods (averages ranging 0.39-0.68 and 0.30-0.53 respectively).
-Similar results were obtained using Spearman correlation (Supplementary Figure 2).
+In aneuploid samples, the local coverage of the Hi-C experiment is
+proportional to the number of copies of the region (a duplicated region in
+a diploid genome will have on average a 50\% increase in the number of
+contacts). See Figure \ref{fig:copy_number}B for a typical example.
+Given that our model is based on the total number of contacts, which in
+turn is related to the number of copies, we reasoned that a
+preliminary test would be to check if the corrected number of contacts per
+bin ($\hat{t}_i = \sum_{i}{\hat{x}_{ij}}$) reflects the copy number (as
+measured by an independent technique such as array-based copy-number
+segmentation) better than the alternatives.
+
+Figure \ref{fig:totals}B shows the total number of corrected contacts
+along chromosme 17 of a T47D sample. Our correction approach presents a
+more stable and precise estimation of the copy number. Quantitatively
+(Figure \ref{fig:copy_number}), Pearson correlation with independent copy
+number data is high for both T47D and K562 cell lines (averages of 0.87
+and 0.78 respectively), not only compared with raw data (averages of 0.74
+and 0.63 respectively) but also with any of the alternative methods
+(averages ranging 0.39-0.68 and 0.30-0.53 respectively). Similar results
+were obtained using Spearman correlation (Supplementary Figure 2).
 
 \begin{figure}
-	\centerline{\includegraphics[width=.45\textwidth]{img/copy_number_figure2.pdf}}
+	\centerline{\includegraphics[width=.45\textwidth]
+{img/copy_number_figure2.pdf}}
 	\caption{
-		Pearson correlation between total number of contacts per bin and independent copy number estimation (COSMIC) for each of the methods compared.
-Left panel T47D breast cancer cell line, right panel K562 leukemia cell line.
-The new proposal (in blue) outperforms the rest of alternatives.
-	}\label{fig:copy_number}
+    Pearson correlation between total number of contacts per bin and
+independent copy number estimation (COSMIC) for each of the methods
+compared. Left panel T47D breast cancer cell line, right panel K562
+leukemia cell line. The new proposal (in blue) outperforms the rest of
+alternatives.  }\label{fig:copy_number}
 \end{figure}
@@
 
-Focusing on the Pearson correlation of observed over expected counts, methods based on the iterative correction (mainly \textit{ice} and \textit{caib} but also \textit{vanilla}) present the undesirable feature of higher correlations between unrelated samples than using raw data (top left panels of Figures \ref{fig:aberrant}A and \ref{fig:aberrant}C, Supplementary Figure 3).
-Among the comparisons of cells with the same type, \textit{oned} is the only method outperforming uncorrected data when comparing experiments using different protocols (top right panel of Figure \ref{fig:aberrant}A) and it is among the top performant when comparing experiments using the same protocol (rest of the panels in Figures \ref{fig:aberrant}A and \ref{fig:aberrant}C).
-Consequently, our proposal is the one that better discriminates unrelated samples form similar ones (as it can be seen in the ROC curves and AUCs of Figures \ref{fig:aberrant}B and D).
+Focusing on the Pearson correlation of observed over expected counts,
+methods based on the iterative correction (mainly \textit{ice} and
+\textit{caib} but also \textit{vanilla}) present the undesirable feature
+of higher correlations between unrelated samples than using raw data (top
+left panels of Figures \ref{fig:aberrant}A and \ref{fig:aberrant}C,
+Supplementary Figure 3). Among the comparisons of cells with the same
+type, \textit{oned} is the only method outperforming uncorrected data when
+comparing experiments using different protocols (top right panel of Figure
+\ref{fig:aberrant}A) and it is among the top performant when comparing
+experiments using the same protocol (rest of the panels in Figures
+\ref{fig:aberrant}A and \ref{fig:aberrant}C). Consequently, our proposal
+is the one that better discriminates unrelated samples form similar ones
+(as it can be seen in the ROC curves and AUCs of Figures
+\ref{fig:aberrant}B and D).
 
-Complementary analysis using other metrics (Supplementary Figures 3, 4 and 5) yielded similar results.
-Notably, Spearman correlation of contacts presents the worst performance for all scenarios and seems to be a poor choice as a metric to compare HiC matrices.
+**
+In an attempt to further summarize the results, we reasoned that the
+correlation between unrelated samples (\textit{i.e} different cell types)
+should be lower than the correlation between related samples (\textit{i.e}
+same cell types), so we could re-interpret each correlation metric as a
+classifier score.
+**
+
+Complementary analysis using other metrics (Supplementary Figures 3, 4 and
+5) yielded similar results. Notably, Spearman correlation of contacts
+presents the worst performance for all scenarios and seems to be a poor
+choice as a metric to compare HiC matrices.
 
@@
     Results of the comparison between samples with aberrant karyotype.
-Left panels: Average difference and 95\% CI between each correction method (Y axis) and the raw data (brown dashed line) for T47D (A) and K562 (C) sets.
-Right panels: Corresponding ROC curves and AUCs for T47D (B) and K562 (D) sets.
-All results in this figure are based on Pearson correlations between the observed over expected counts.
-Our method increases reproducibility of similar samples while being able to discriminate different ones.
-	}\label{fig:aberrant}
+Left panels: Average difference and 95\% CI between each correction method
+(Y axis) and the raw data (brown dashed line) for T47D (A) and K562 (C)
+sets. Right panels: Corresponding ROC curves and AUCs for T47D (B) and
+K562 (D) sets. All results in this figure are based on Pearson
+correlations between the observed over expected counts. Our method
+increases reproducibility of similar samples while being able to
+discriminate different ones.  }\label{fig:aberrant}
 \end{figure}
@@
 
-Regarding cells with diploid genomes, we confirmed that our bias correction strategy also achieves good results in terms separation between unrelated and similar samples (Figure \ref{fig:normal} and Supplementary Figures 6 and 7), increasing the reproducibility across different experimental protocols for primary B and ES mouse cell samples.
-Again, Spearman correlation of contacts revealed as the worst metric.
+Regarding cells with diploid genomes, we confirmed that our bias
+correction strategy also achieves good results in terms separation between
+unrelated and similar samples (Figure \ref{fig:normal} and Supplementary
+Figures 6 and 7), increasing the reproducibility across different
+experimental protocols for primary B and ES mouse cell samples. Again,
+Spearman correlation of contacts revealed as the worst metric.
 

### 2017-06-02 main.tex
Wrote the first paragraph of the results.

@@
 
-\title[Dryhic]{OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes}
+\title[OneD]{OneD: increasing reproducibility of Hi-C Samples with
+abnormal karyotypes}
 
-\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$, François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb 1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina Cuartero\,$^{\text{\sfb 1,2}}$,Miguel Beato\,$^{\text{\sfb 1,2}}$ and Guillaume Fillion\,$^{\text{\sfb 1,2}}$}
+\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
+François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
+1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
+Cuartero\,$^{\text{\sfb 1,2}}$,Miguel Beato\,$^{\text{\sfb 1,2}}$ and
+Guillaume Fillion\,$^{\text{\sfb 1,2}}$}
 
-\address{$^{\text{\sf 1}}$Gene Regulation, Stem Cells and Cancer Program, Centre for Genomic Regulation (CRG), The
-	Barcelona Institute of Science and Technology (BIST), Barcelona, Spain \\
+\address{$^{\text{\sf 1}}$Gene Regulation, Stem Cells and Cancer Program,
+Centre for Genomic Regulation (CRG), The Barcelona Institute of Science
+and Technology (BIST), Barcelona, Spain \\
 $^{\text{\sf 2}}$Universitat Pompeu Fabra (UPF), Barcelona, Spain}
@@
 terms of computing resources for resolutions up to 1 kbp.\\	
-	\textbf{Availability:} OneD is implemented as an R package available at \href{http://www.github.com/qenvio/dryhic}{http://www.github.com/qenvio/dryhic}.\\	
-	\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
-	\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}	online.}
+\textbf{Availability:} OneD is implemented as an R package available at
+\href{http://www.github.com/qenvio/dryhic}
+{http://www.github.com/qenvio/dryhic}.\\	
+textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
+textbf{Supplementary information:} Supplementary data are available at
+\textit{Bioinformatics}	online.}
+
 \maketitle
@@
 \subsection{Model}
+\label{sec:model}
 
@@
 
-\begin{align*}
+\begin{align}
+\notag
 \lambda_i' &= \frac{\lambda_i}{\sum_j^n{\lambda_j}/n} \\
+\label{eq:xhat}
 \hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}.
-\end{align*}
+\end{align}
 
@@
 can be easily extended with any user-provided genomic feature.
-
-\begin{figure}[!tpb]
-\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
-\caption{A. Non linear relationship between the number of restriction
-sites per bin and the total number of contacts per bin in T47D. B. Total
-number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
-corrected (blue) signals are shown. The region that corresponds to the
-first 20 Mbp of the chromosome is present in single copy, explaining that
-the signal is approximately half compared to the rest of the chromosome.
-Also notice that the raw signal is noisier than the corrected one.}
-\label{fig:totals}
-\end{figure}
 
@@
 lines obtained from COSMIC database \citep{forbes2010cosmic}.
+
+\begin{table*}
+\processtable{Hi-C data sets used in this study.
+\label{tab:samples}}
+{\begin{tabular}{ccccccc}
+  \toprule
+  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
+  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set} &
+  \textbf{Source} \\
+  \midrule
+dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
+  DpnII & CRG & T47D, K562 & NA \\
+b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
+  DpnII & CRG & T47D & NA \\
+dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
+  DpnII & CRG & T47D, K562 & NA \\
+HindIII\_T0 & T47D & dilution Hi-C &
+  HindIII & CRG & T47D & SRR1054341 \\
+NcoI\_T0    & T47D & dilution Hi-C &
+  NcoI      & CRG & T47D & SRR1054343 \\
+ENCLB758KFU & T47D & dilution Hi-C &
+  HindIII   & UMass & T47D & ENCLB758KFU \\
+ENCLB183QHG & T47D & dilution Hi-C &
+  HindIII   & UMass & T47D & ENCLB183QHG \\
+HIC069  & K562 & \textit{in situ} Hi-C &
+  MboI & Baylor & T47D, K562 &    SRR1658693 \\
+HIC070  & K562 & \textit{in situ} Hi-C &
+  MboI & Baylor & T47D, K562 &    SRR1658694 \\
+HIC071  & K562 & \textit{in situ} Hi-C &
+  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
+HIC074  & K562 & \textit{in situ} Hi-C &
+  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
+b7fa2d8db\_bfac48760 & B-cell  & \textit{in situ} Hi-C &
+  DpnII & CRG & mm10 & GSE96611 \\
+fc3e8b36a\_7bf1bf374 & ES-cell & \textit{in situ} Hi-C &
+  DpnII & CRG & mm10 & GSE96611 \\
+b7fa2d8db\_7284b867a & B-cell  & \textit{in situ} Hi-C &
+  DpnII & CRG & mm10 & GSE96611 \\
+fc3e8b36a\_38bfd1b33 & ES-cell & \textit{in situ} Hi-C &
+  DpnII & CRG & mm10 & GSE96611 \\
+b7fa2d8db\_58e812fc2 & B-cell  & \textit{in situ} Hi-C &
+  DpnII & CRG & mm10 & GSE96611 \\
+fc3e8b36a\_c990a254e & ES-cell & \textit{in situ} Hi-C &
+  DpnII & CRG & mm10 & GSE96611 \\
+b7fa2d8db\_73f11d923 & B-cell  & \textit{in situ} Hi-C &
+  DpnII & CRG & mm10 & GSE96611 \\
+fc3e8b36a\_4bf044f18 & ES-cell & \textit{in situ} Hi-C &
+  DpnII & CRG & mm10 & GSE96611 \\
+GSM987817 & B-cell  & dilution Hi-C &
+  HindIII & UCSC & mm10 & SRR543428-SRR543431 \\
+GSM987818 & B-cell  & dilution Hi-C &
+  HindIII & UCSC & mm10 & SRR543432-SRR543442 \\
+GSM862720 & ES-cell & dilution Hi-C &
+  HindIII & UCSC & mm10 & SRR443883-SRR443885 \\
+GSM862721 & ES-cell & dilution Hi-C &
+  HindIII & UCSC & mm10 & SRR400251-SRR400256 \\
+GSM862722 & ES-cell & dilution Hi-C &
+  NcoI    & UCSC & mm10 & SRR443886-SRR443888 \\
+  \botrule
+\end{tabular}}{}
+\end{table*}
 
@@
 simplest metric is the Spearman correlation applied to intra-chromosomal
-contacts up to a given distance (5 Mb in what follows). The second is
-to compare the observed over expected contacts via the Pearson correlation
-up to a given distance range. Compared to the first, this metric gives
-more weight to changes occuring away from the diagonal. The third option
-is to compute a correlation per distance stratum and then obtain a
-stratum-adjusted correlation coefficient (SCC) as defined in
-\cite{yang2017hicrep}.  Finally, \cite{yan2017hicspector} proposed a
-reproducibility score based on the correlation between the eigenvectors of
-the Laplacian matrix associated with the smallest eigenvalues. This
-approach borrows the concepts of spectral clustering
-\citep{von2007tutorial} and amounts to comparing high level features of
-the matrix.
+contacts up to a given distance (5 Mb in what follows). The second option
+is to measure the similarity of observed over expected contacts via the
+Pearson correlation up to a given distance range. Compared to the first,
+this metric gives more weight to changes occuring away from the diagonal.
+The third option is to compute a correlation per distance stratum and then
+obtain a stratum-adjusted correlation coefficient (SCC) as defined in
+\cite{yang2017hicrep}. Finally, the last option, proposed by
+\cite{yan2017hicspector} is to measure the Pearson correlation between the
+last eigenvectors of the Laplacian of the Hi-C matrix. This approach
+borrows the concepts of spectral clustering \citep{von2007tutorial} and
+amounts to comparing high level features of the matrix.
 
@@
 
-\subsection{Copy number}
+\subsection{Bias correction}
 
-In aneuploid samples, the local coverage of the Hi-C experiment is
-proportional to the number of copies of the region (a duplicated region in
-a diploid genome will have on average a 50\% increase in the number of
-contacts). See Figure \ref{fig:copy_number}B for a typical example.
-Given that our model is based on the total number of contacts, which in
-turn is related to the number of copies, we reasoned that a
-preliminary test would be to check if the corrected number of contacts per
-bin ($\hat{t}_i = \sum_{i}{\hat{x}_{ij}}$) reflects the copy number (as
-measured by an independent technique such as array-based copy-number
-segmentation) better than the alternatives.
+The principle of OneD is to model explicit Hi-C biases onto a single
+variable: the total amount of contacts for each bin of the matrix (see
+section~\ref{sec:model}). The reason for this choice of that the total
+amount of contacts is approximately proportional to the local copy number.
+For instance, a duplicated region in a diploid genome will show on average
+a 50\% increase in the number of contacts.
 
-Figure \ref{fig:totals}B shows the total number of corrected contacts
-along chromosme 17 of a T47D sample. Our correction approach presents a
-more stable and precise estimation of the copy number. Quantitatively
-(Figure \ref{fig:copy_number}), Pearson correlation with independent copy
-number data is high for both T47D and K562 cell lines (averages of 0.87
-and 0.78 respectively), not only compared with raw data (averages of 0.74
-and 0.63 respectively) but also with any of the alternative methods
-(averages ranging 0.39-0.68 and 0.30-0.53 respectively). Similar results
-were obtained using Spearman correlation (Supplementary Figure 2).
+\begin{figure}[!tp]
+\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
+\caption{A. Non linear relationship between the number of restriction
+sites per bin and the total number of contacts per bin in T47D. B. Total
+number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
+corrected (blue) signals are shown. The region that corresponds to the
+first 20 Mbp of the chromosome is present in single copy, explaining that
+the signal is approximately half compared to the rest of the chromosome.
+Also notice that the raw signal is noisier than the corrected one.}
+\label{fig:totals}
+\end{figure}
+
+Other biases will also affect the total amount of contacts, though not
+proportionately. Figure~\ref{fig:totals}A shows the relationship
+between the amount of contacts and the availability of retriction enzyme
+sites. Four clouds are visible. Each corresponds to a copy number
+between one and four. In all of them, the relationship flattens as the
+number of restriction sites increases. OneD thus fits a non-linear
+relationship between the total amount of contacts and the predictors of
+bias (by default the G+C content, the number of restriction sites and the
+mappability of the reads).
+
+The biases are estimated for each bin of the genome, and each cell of the
+matrix is corrected using equation (\ref{eq:xhat}). The corrected amount
+of contacts is still proportional to the copy number, but it is now less
+noisy. The correction thus represents a more stable and precise estimation
+of the copy number. Figure \ref{fig:totals}B shows the corrected number of
+contacts along chromosme 17 of a T47D sample. The corrected contacts have
+a narrower range than the raw contacts.
+
+We tested the validity of this approach against the Catalogue Of Somatic
+Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure
+\ref{fig:copy_number} shows the Pearson correlation between the corrected
+number of contacts and the copy number estimation for both T47D and K562
+cell lines. OneD achieves the highest correlations in both cases.
+Actually, it is the only method that increases the agreement with the
+external data. The other methods, including caICB, only make the
+normalized data more divergent than the raw counts.  Similar results were
+obtained using Spearman correlation (Supplementary Figure 2).
+
+Taken together, these results indicate that OneD is an efficient way to
+estimate the local copy number in order to correct for it.
 

### 2017-06-03 main.tex
Wrote the result section about aberrant karyotypes.

@@
 1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
-Cuartero\,$^{\text{\sfb 1,2}}$,Miguel Beato\,$^{\text{\sfb 1,2}}$ and
+Cuartero\,$^{\text{\sfb 1,2}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$ and
 Guillaume Fillion\,$^{\text{\sfb 1,2}}$}
@@
   \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
-  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set} &
+  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set(s)} &
   \textbf{Source} \\
@@
 \subsection{Comparison of Hi-C matrices}
+\label{sec:comp}
 
@@
 
-We benchmark \textbf{\textit{oned}}, against methods
-\textbf{\textit{vanilla}}, \textbf{\textit{ice}}
-\citep{imakaev2012iterative}, \textbf{\textit{caib}}
+We benchmarked \textbf{\textit{oned}}, against \textbf{\textit{vanilla}},
+\textbf{\textit{ice}} \citep{imakaev2012iterative}, \textbf{\textit{caib}}
 \citep{wu2016computational} and \textbf{\textit{lgf}}
@@
 
-\subsection{Bias correction}
+\subsection{Correction method}
 
-The principle of OneD is to model explicit Hi-C biases onto a single
+The principle of OneD is to explicitly model Hi-C biases on a single
 variable: the total amount of contacts for each bin of the matrix (see
-section~\ref{sec:model}). The reason for this choice of that the total
-amount of contacts is approximately proportional to the local copy number.
-For instance, a duplicated region in a diploid genome will show on average
-a 50\% increase in the number of contacts.
+\ref{sec:model}). The reason for this choice of that the total amount of
+contacts is approximately proportional to the local copy number. For
+instance, a duplicated region in a diploid genome will show on average a
+50\% increase in the number of contacts. Discontinuities of the amount of
+contacts thus correspond to changes of the copy number.
 
@@
 corrected (blue) signals are shown. The region that corresponds to the
-first 20 Mbp of the chromosome is present in single copy, explaining that
-the signal is approximately half compared to the rest of the chromosome.
-Also notice that the raw signal is noisier than the corrected one.}
+first 20 Mbp of the chromosome is present in single copy in this clone of
+T47D, explaining that the signal is approximately half compared to the
+rest of the chromosome.  Also notice that the raw signal is noisier than
+the corrected one.}
 \label{fig:totals}
@@
 
-Other biases will also affect the total amount of contacts, though not
-proportionately. Figure~\ref{fig:totals}A shows the relationship
-between the amount of contacts and the availability of retriction enzyme
-sites. Four clouds are visible. Each corresponds to a copy number
-between one and four. In all of them, the relationship flattens as the
-number of restriction sites increases. OneD thus fits a non-linear
-relationship between the total amount of contacts and the predictors of
+Other biases affect the total amount of contacts in a countinuous but not
+necessarily linear way. Figure~\ref{fig:totals}A shows the relationship
+between the amount of contacts and the number of retriction enzyme sites
+in T47D, a breast cancer cell line with aberrant and unstable karyotype.
+Four clouds are visible. Each corresponds to a copy number between one and
+four. In all of them, the relationship flattens as the number of
+restriction sites increases. For this reason, OneD fits a non-linear
+relationship between the total amount of contacts and the known sources of
 bias (by default the G+C content, the number of restriction sites and the
@@
 
-The biases are estimated for each bin of the genome, and each cell of the
-matrix is corrected using equation (\ref{eq:xhat}). The corrected amount
-of contacts is still proportional to the copy number, but it is now less
-noisy. The correction thus represents a more stable and precise estimation
-of the copy number. Figure \ref{fig:totals}B shows the corrected number of
-contacts along chromosme 17 of a T47D sample. The corrected contacts have
-a narrower range than the raw contacts.
+The biases are estimated genome-wide and each cell of the matrix is then
+corrected using equation (\ref{eq:xhat}). Note that the corrected amount
+of contacts is still proportional to the copy number. The correction thus
+represents a more stable and precise estimation of the copy number.
+Figure~\ref{fig:totals}B shows the corrected number of contacts along
+chromosme 17 of a T47D. OneD greatly reduces the wiggling of the total
+amount of contacts.
 
@@
 number of contacts and the copy number estimation for both T47D and K562
-cell lines. OneD achieves the highest correlations in both cases.
-Actually, it is the only method that increases the agreement with the
-external data. The other methods, including caICB, only make the
-normalized data more divergent than the raw counts.  Similar results were
-obtained using Spearman correlation (Supplementary Figure 2).
-
-Taken together, these results indicate that OneD is an efficient way to
-estimate the local copy number in order to correct for it.
+cell lines. Similar results were obtained using Spearman correlation
+(Supplementary Figure 2). All the methods except OneD decrease the
+agreement with the copy number because they partially correct this bias.
+In contrast, OneD enhances the conformity of the signal with the copy
+number. Not correcting for variable copy number at that stage may seem to
+defeat the purpose of a normalization method, but we will see below how
+this can lead to better performance.
 
@@
 
-\subsection{Samples with aberrant karyotypes}
+\subsection{Aberrant karyotypes}
 
-Focusing on the Pearson correlation of observed over expected counts,
-methods based on the iterative correction (mainly \textit{ice} and
-\textit{caib} but also \textit{vanilla}) present the undesirable feature
-of higher correlations between unrelated samples than using raw data (top
-left panels of Figures \ref{fig:aberrant}A and \ref{fig:aberrant}C,
-Supplementary Figure 3). Among the comparisons of cells with the same
-type, \textit{oned} is the only method outperforming uncorrected data when
-comparing experiments using different protocols (top right panel of Figure
-\ref{fig:aberrant}A) and it is among the top performant when comparing
-experiments using the same protocol (rest of the panels in Figures
-\ref{fig:aberrant}A and \ref{fig:aberrant}C). Consequently, our proposal
-is the one that better discriminates unrelated samples form similar ones
-(as it can be seen in the ROC curves and AUCs of Figures
-\ref{fig:aberrant}B and D).
+We first benchmarked the performance of OneD on biological samples with an
+aberrant karyotype. A good normalization method should increase the
+similarity between biological replicates by reducing irrelevant
+experimental variation, such as batch effects, laboratory of origin and
+protocol variations. Similarly, a good normalization should decrease the
+similarity between samples that received a different treatment in order to
+enhance the biological differences.
 
-**
-In an attempt to further summarize the results, we reasoned that the
-correlation between unrelated samples (\textit{i.e} different cell types)
-should be lower than the correlation between related samples (\textit{i.e}
-same cell types), so we could re-interpret each correlation metric as a
-classifier score.
-**
+We assembled two Hi-C data sets obtained from T47D and K562, two cancer
+cell lines with aberrant karyotypes. In each set we spiked two
+samples from the other cell line (see Table~\ref{tab:samples}) in order to
+introduce biological variability. We measured the Pearson correlation of
+observed over expected counts to compare matrices before and after
+normalization by different methods (see \ref{sec:comp}). This gave an
+indication of the impact of a given normalization method. The results are
+summarized in Figure~\ref{fig:aberrant}.
 
-Complementary analysis using other metrics (Supplementary Figures 3, 4 and
-5) yielded similar results. Notably, Spearman correlation of contacts
-presents the worst performance for all scenarios and seems to be a poor
-choice as a metric to compare HiC matrices.
+The \textit{caib} and \textit{ice} methods increased the similarity
+between the different cell lines (Figures~\ref{fig:aberrant}A and
+\ref{fig:aberrant}C and Supplementary Figure~3). This is an undesirable
+effect, as it obscures the biological variability. In the same vein, these
+methods decreased the similarity between samples that received the same
+treatment (Figures~\ref{fig:aberrant}A), suggesting that the normalization
+process is detrimental to the biological signal in these two cases. The
+method \textit{vanilla} followed the same trend but to a lesser extent,
+consistent with the fact that it consists of a single \textit{ice}
+iteration.
+
+OneD was the only method to increase the similarity between experiments
+carried out on the same material but with a different protocol
+(Figure~\ref{fig:aberrant}A). In these conditions, \textit{lgf} was
+decreasing the similarity. Taken together, the results show that OneD
+enhanced biological variation while decreasing experimental variation.
+
+An important application of normalization methods in experimental setups
+is to identify outliers. We thus investigated the capacity of the
+different methods to help identify the samples from the other cell type
+spiked in the data set. We interpreted the pairwise comparison scores as
+classifier scores and summarized the results with a ROC curve
+(Figures~\ref{fig:aberrant}B and D).
+
+All the methods, including the absence of normalizaton, succeeded in
+recognizing the T47D outliers among the K562 samples, but recognizing the
+K562 outliers among the T47D samples proved more challenging. OneD
+increased the discrimination power compared to the raw matrices, but all
+the other methods decreased it. Actually, they performed little better
+than random on this task.
+
+Using the other metrics described in \ref{sec:comp} yielded similar
+results (Supplementary Figures 3, 4 and 5). Note that Spearman correlation
+of contacts presented the worst performance for all scenarios, and it thus
+seems to be a poor choice as a metric to compare HiC matrices.
+
+Taken together, these results show that OneD enhances biological variation
+and reduces experimental noise on samples with an aberrant karyotype.
 
@@
 \caption{
-    Results of the comparison between samples with aberrant karyotype.
-Left panels: Average difference and 95\% CI between each correction method
-(Y axis) and the raw data (brown dashed line) for T47D (A) and K562 (C)
-sets. Right panels: Corresponding ROC curves and AUCs for T47D (B) and
-K562 (D) sets. All results in this figure are based on Pearson
-correlations between the observed over expected counts. Our method
-increases reproducibility of similar samples while being able to
-discriminate different ones.  }\label{fig:aberrant}
+Results of the comparison between samples with aberrant karyotype.  A and
+B. Average changes compared to raw on the T47D and K562 sets.  The bars
+represent 95\% confidence intervals centered on the mean difference of the
+correlation score between a given correction method and the raw data. The
+brown dashed line indicates the value of the average score on raw matrices
+(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
+the curve are indicated in the bottom right corner. The color code is the
+same as panels A and B. The brown lines correspond to raw matrices.  All
+results in this figure are based on Pearson correlations between the
+observed over expected counts.  Our method increases reproducibility of
+similar samples while being able to discriminate different ones.}
+\label{fig:aberrant}
 \end{figure}
 
-\subsection{Samples with normal karyotypes}
+\subsection{Normal karyotypes}
 

### 2017-06-03 main.tex
Finished the text.

@@
 \maketitle
+
+%% --------------- Introduction --------------- %%
 
@@
 results in this figure are based on Pearson correlations between the
-observed over expected counts.  Our method increases reproducibility of
-similar samples while being able to discriminate different ones.}
+observed over expected counts.}
 \label{fig:aberrant}
@@
 
-Regarding cells with diploid genomes, we confirmed that our bias
-correction strategy also achieves good results in terms separation between
-unrelated and similar samples (Figure \ref{fig:normal} and Supplementary
-Figures 6 and 7), increasing the reproducibility across different
-experimental protocols for primary B and ES mouse cell samples. Again,
-Spearman correlation of contacts revealed as the worst metric.
+Does the performance of OneD with aberrant karyotypes comes at the cost of
+decreased performance with normal karyotypes? To address this question, we
+assembled another data set comprised of normal mouse B cells and embryonic
+stem (ES) cells. The cell types were pooled in almost equal proportions
+(see Table~\ref{tab:samples}) and the same tests as above were performed.
+
+In these conditions, variations of the protocol had a strong effect on the
+impact of the differen normalization methods (Figure~\ref{fig:normal}A).
+For instance, \textit{caib} and \textit{ice} increased the similarity when
+the protocols were different, but decreased it when the protocols were the
+same. The effect was stronger when comparing identical cell types, but the
+same trend appeared when compaing different cell types, indicating that
+these methods may enhance or reduce biological variation, depending on the
+context. Once more, \textit{vanilla} followed the same trend as
+\textit{ice} to a lesser extent. The \textit{lgf} method increased the
+similarity when comparing the same cells with different experimental
+protocols, and had little to no effect in the other cases. This indicates
+that \textit{lgf} is a safe choice in this case.
+
+OneD decreased the similarity between different cell types when using the
+same protocol and increased it between identical cell types when using
+different protocols. In this other two cases, it had little effect. In
+summary, OneD never enhanced the experimental noise and it reduced it in
+one more case than \textit{lgf}.
+
+When interpreting the similarity scores as classification scores, we
+observed that all the methods could identify approximately 50\% of the
+pairs of biologicate pairs, after which their performance diverged
+(Figure~\ref{fig:normal}B). OneD achieved the highest area under the curve
+on this problem, even though it was not always above the others methods at
+every position of the graph.
+
+Using other metrics to compare matrices gave similar results
+(Supplementary Figures 6 and 7). OneD was always among the top scoring
+methods. In these conditions, Spearman correlation of contacts again
+appeared as as the worst comparison metric as it showed a lower
+performance for all the normalization methods.
+
+Taken togetheer, these results indicate that OneD performs as well as the
+best normalization methods even with normal karyotypes.
 
@@
 \caption{
-		Results of the comparison between samples with normal karyotype.
-A Average difference and 95\% CI between each correction method (Y axis) and the raw data (brown dashed line) for the mm10 sets.
-B Corresponding ROC curves and AUCs.
-The introduced strategy does not under perform compared to the alternatives.
-	}\label{fig:normal}
+Results of the comparison between samples with normal karyotype.  A.
+Average changes compared to raw on the mouse data set. The bars are as in
+B. ROC curves on the mouse data set. The legend is as in
+Figure~\ref{fig:aberrant}.}
+\label{fig:normal}
 \end{figure}
 
-\subsection{Computing resources}
+\subsection{Speed}
 
-One of the main reasons for the broad usage of the iterative correction \citep{imakaev2012iterative} is the easy and fast implementation, moreover compared to existing explicit methods \citep{servant2012hitc}.
-Our proposal uses only the total number of contacts per bin instead the full matrix for the model estimation.
-Thus, it achieves similar computing times that the fastest approaches (Figure \ref{fig:times} A).
-Yet it scales with increasing resolutions (decreasing bin size) (Figure \ref{fig:times} B).
+Finally, we compared the computational speed of the different
+normalization methods. One of the main reasons for the broad acceptance of
+\textit{vanilla} and \textit{ICE} as standards is the speed of the
+available implementation \citep{imakaev2012iterative}. This is even 
+more significant that current explicit methods \citep{servant2012hitc} are
+much slower in comparison.
+
+Unlike those, OneD corrects a single variable instead of the whole matrix.
+It uses only the total number of contacts per bin and thus estimates the
+model parameters much faster than previous explicit methods.
+Figure\ref{fig:times} shows the running times of the different methods on
+the samples described in Table~\ref{tab:samples}. The fastest method is
+\textit{vanilla} and the slowest is \textit{lgf}, with an over 100-fold
+span between the two.
+
+OneD is the second fastest method and it always ran in less than 1 min in
+the conditions of the benchmark. Interestingly, the running time of OneD
+is much more homogeneous than that of the other methods. The reason is
+that the size of the regression problem to be solved by the \texttt{mgcv}
+package is always the same at fixed reslution.
+
+Taken together, these results show that OneD is very competitive in terms
+of computational speed compared to existing methods.
 
@@
 \caption{
-		Computing time (Y axis, log scale) of all the bias correction methods used. Each dot corresponds to one sample. The only method faster than ours under performs in all sample comparisons.
-	}\label{fig:times}
+Computing time of the bias correction methods. Each dot corresponds to
+one sample. The only method faster than ours under performs in all sample
+comparisons. Note the logarithmic scale on the y-axis.}
+\label{fig:times}
 \end{figure}
+
+%% --------------- Discussion --------------- %%
 
@@
 
+Here we introduced OneD, a fast computational method to normalize Hi-C
+matrices. OneD was developed ground up to address the need to normalize
+data collected on samples with aberrant karyotypes, but it applies
+seemlessly to the case of normal karyotypes. We showed that OneD performs
+significantly better than other methods when the cells present karyotypic
+aberrations (Figure~\ref{fig:aberrant}), and that it performs equall well
+as the best methods otherwise (Figure~\ref{fig:normal}). We also showed
+that OneD is on average faster than \textit{ice}, which makes it very
+competitive from the point of view of computational speed.
+
+The originality of OneD lies in that it projects all the biases onto a
+single variable: the total amount of contact per bin. This allows greater
+running speed, while preserving a good performance on samples with
+karyotypic aberrations. One of the reasons that OneD is better able to
+highlight the biological distinctions between samples is that it does not
+correct the copy number but enhances it (Figures~\ref{fig:totals} and
+\ref{fig:copy_number}).
+
+This raises the question whether variations of the copy number constitute
+a biological signal or an artifact. Note that in the presence of
+translocations, deletions and large scale rearrangements, correcting the
+copy number is equally faulty as not correcting it. Indeed, in this case
+the reference genome on which the matrix is drawn is incorrect and
+changing the values of the pixels cannot account for this information.
+
+Depending on the intention of the user, the effect of the copy number
+should either be kept or removed. This is why OneD leaves does not perform
+the correction by default, but allows the user to obtain a
+diploid-equivalent Hi-C map computed from a hidden Marko model. The
+resulting matrices have a near constant amount of contacts per bin, but
+the artifacts caused by the misfit between the genome of the sample and
+the reference genome are still present (for instance, the artifacts caused
+by large scale inversions are not changed in any way).
+
+Overall, OneD constitutes a safe choice to normalize Hi-C matrices. If the
+karyotype of the sample is aberrant, it enhances the biological variation.
+If not, it performs at least equally well as other methods with an
+advantage in terms of computational speed.
+
 \section*{Acknowledgements}
 
-We would like to thank all members of the four labs involved in the 4DGenome Synergy project for the fruitful discussions during project meetings.
-EV wants to acknowledge specially the members of Miguel Beato's lab for their insights during lab meetings.
+We would like to thank the members of the 4DGenome Synergy project for the
+fruitful discussions during project meetings. EV wants to acknowledge
+the members of Miguel Beato's laboratory for their insights during lab
+meetings.
 
@@
 Competitiveness, ‘Centro de Excelencia Severo Ochoa 2013-2017’,
-SEV-2012-0208, ERC Synergy Grant 609989 (GF) and...
+SEV-2012-0208, ERC Synergy Grant 609989 and...
+
 \vspace*{-12pt}

### 2017-06-04 main.tex
Added benchmark info and HMM method.

@@
 can be easily extended with any user-provided genomic feature.
+
+\subsection{Copy number correction}
+
+Briefly, a hidden Markov model with emissions distributed as a Student's
+$t$ variable is fitted on the corrected total amount of contacts per bin.
+The model consists of 7 states that correspond to 1, 2, 3, 4, 5, 6 and 8
+copies of the target, for a total of 3 emission parameters (a single
+scaling parameter, a single standard deviation and a single degree of
+freedom for all the states) and 21 transition parameters.
+
+The model is fitted with the Baum-Welch algorithm \citep{baum1966} until
+convergence, following the detail given by \cite{filion2010systematic}.
+The Viterbi path is then computed and corresponds to the inferred copy
+number of each bin ($c_i$).
+
+A correction equal to the square root of the copy number is then applied
+to the whole matrix. More specifically, each cell is updated to
+
+\begin{equation*}
+\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}.
+\end{equation*}
 
@@
 \textit{vanilla} and \textit{ICE} as standards is the speed of the
-available implementation \citep{imakaev2012iterative}. This is even 
-more significant that current explicit methods \citep{servant2012hitc} are
-much slower in comparison.
+available implementation \citep{imakaev2012iterative}. This is even more
+significant that current explicit methods \citep{servant2012hitc} are much
+slower in comparison.
 
-Unlike those, OneD corrects a single variable instead of the whole matrix.
-It uses only the total number of contacts per bin and thus estimates the
-model parameters much faster than previous explicit methods.
-Figure\ref{fig:times} shows the running times of the different methods on
-the samples described in Table~\ref{tab:samples}. The fastest method is
+Unlike the other methods, OneD corrects a single variable instead of the
+whole matrix. It uses only the total number of contacts per bin and thus
+estimates the model parameters much faster than previous explicit methods.
+We measured the running time of the different tools on a 3.3~GHz machine
+with 62 GB~RAM, always using the default paramters. Figure\ref{fig:times}
+shows the running times of the different methods on the samples described
+in Table~\ref{tab:samples} at 100 kbp resolution. The fastest method is
 \textit{vanilla} and the slowest is \textit{lgf}, with an over 100-fold
@@
 OneD is the second fastest method and it always ran in less than 1 min in
-the conditions of the benchmark. Interestingly, the running time of OneD
-is much more homogeneous than that of the other methods. The reason is
-that the size of the regression problem to be solved by the \texttt{mgcv}
-package is always the same at fixed reslution.
+the conditions of the benchmark. In all cases, the memory footprint was
+less than 300 MB. Interestingly, the running time of OneD is much more
+homogeneous than that of the other methods. The reason is that the size of
+the regression problem to be solved by the \texttt{mgcv} package is always
+the same at fixed reslution.
 

### 2017-06-05 main.tex
Added figure 5B and fixed small typos

@@
 presence of large-scale copy number variations, obscuring biological
-viarations and enhancing batch effects. To address this issue, we
-developed an alternative appoach designed to take into account chromosomal
+variations and enhancing batch effects. To address this issue, we
+developed an alternative approach designed to take into account chromosomal
 abnormalities. The method, called OneD, increases reproducibility among
@@
 outperforming previous methods.  In the absence of chromosomal
-aberrantions, OneD fared equally well as state-of-the-art methods, making
+aberrations, OneD fared equally well as state-of-the-art methods, making
 it a safe choice for Hi-C normalization. OneD is fast and scales well in
@@
 {http://www.github.com/qenvio/dryhic}.\\	
-textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
-textbf{Supplementary information:} Supplementary data are available at
+\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
+\textbf{Supplementary information:} Supplementary data are available at
 \textit{Bioinformatics}	online.}
@@
 \begin{align*}
-t_i &= \sum_j^n{x_{ij}} \sim  NB(\lambda_i, \theta), \\
-\log(\lambda_i) &\propto \sum_{k}{f_k(z_k)},
+t_i = \sum_j^n{x_{ij}} &\sim  NB(\lambda_i, \theta) \\
+\log(\lambda_i) &\propto \sum_{k}{f_k(z_k)}
 \end{align*}
@@
 \label{eq:xhat}
-\hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}.
+\hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}
 \end{align}
@@
 \begin{equation*}
-\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}.
+\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}
 \end{equation*}
@@
 
-We developed the routines contained in the \texttt{oned} R package to
+We developed the routines contained in the \texttt{dryhic} R package to
 efficiently create sparse representations of contact matrices and further
@@
 Pearson correlation up to a given distance range. Compared to the first,
-this metric gives more weight to changes occuring away from the diagonal.
+this metric gives more weight to changes occurring away from the diagonal.
 The third option is to compute a correlation per distance stratum and then
@@
 classified the comparisons according to the characteristics of each pair
-(cell type, protocol, batch, and treatment).
+(cell type, protocol, batch and treatment).
 
@@
 variable: the total amount of contacts for each bin of the matrix (see
-\ref{sec:model}). The reason for this choice of that the total amount of
+\ref{sec:model}). The reason for this choice is that the total amount of
 contacts is approximately proportional to the local copy number. For
@@
 
-Other biases affect the total amount of contacts in a countinuous but not
+Other biases affect the total amount of contacts in a continuous but not
 necessarily linear way. Figure~\ref{fig:totals}A shows the relationship
-between the amount of contacts and the number of retriction enzyme sites
+between the amount of contacts and the number of restriction enzyme sites
 in T47D, a breast cancer cell line with aberrant and unstable karyotype.
@@
 Figure~\ref{fig:totals}B shows the corrected number of contacts along
-chromosme 17 of a T47D. OneD greatly reduces the wiggling of the total
+chromosome 17 of a T47D. OneD greatly reduces the wiggling of the total
 amount of contacts.
@@
 between the different cell lines (Figures~\ref{fig:aberrant}A and
-\ref{fig:aberrant}C and Supplementary Figure~3). This is an undesirable
+\ref{fig:aberrant}B and Supplementary Figure~3). This is an undesirable
 effect, as it obscures the biological variability. In the same vein, these
 methods decreased the similarity between samples that received the same
-treatment (Figures~\ref{fig:aberrant}A), suggesting that the normalization
+treatment (Figure~\ref{fig:aberrant}A), suggesting that the normalization
 process is detrimental to the biological signal in these two cases. The
@@
 classifier scores and summarized the results with a ROC curve
-(Figures~\ref{fig:aberrant}B and D).
+(Figures~\ref{fig:aberrant}C and D).
 
-All the methods, including the absence of normalizaton, succeeded in
+All the methods, including the absence of normalization, succeeded in
 recognizing the T47D outliers among the K562 samples, but recognizing the
@@
 In these conditions, variations of the protocol had a strong effect on the
-impact of the differen normalization methods (Figure~\ref{fig:normal}A).
+impact of the different normalization methods (Figure~\ref{fig:normal}A).
 For instance, \textit{caib} and \textit{ice} increased the similarity when
@@
 same. The effect was stronger when comparing identical cell types, but the
-same trend appeared when compaing different cell types, indicating that
+same trend appeared when comparing different cell types, indicating that
 these methods may enhance or reduce biological variation, depending on the
@@
 observed that all the methods could identify approximately 50\% of the
-pairs of biologicate pairs, after which their performance diverged
+ biological pairs, after which their performance diverged
 (Figure~\ref{fig:normal}B). OneD achieved the highest area under the curve
@@
 
-Taken togetheer, these results indicate that OneD performs as well as the
+Taken together, these results indicate that OneD performs as well as the
 best normalization methods even with normal karyotypes.
@@
 available implementation \citep{imakaev2012iterative}. This is even more
-significant that current explicit methods \citep{servant2012hitc} are much
+significant as current explicit methods \citep{servant2012hitc} are much
 slower in comparison.
@@
 We measured the running time of the different tools on a 3.3~GHz machine
-with 62 GB~RAM, always using the default paramters. Figure\ref{fig:times}
+with 62 GB~RAM, always using the default parameters. Figure~\ref{fig:times}A
 shows the running times of the different methods on the samples described
@@
 \textit{vanilla} and the slowest is \textit{lgf}, with an over 100-fold
-span between the two.
+span between the two. OneD also scales well in terms of bin size, as can
+be seen in Figure~\ref{fig:times}B, where the different corrections were
+applied to a reduced part of the genome at increasing resolutions.
 
@@
 the regression problem to be solved by the \texttt{mgcv} package is always
-the same at fixed reslution.
+the same at fixed resolution
 
@@
 \begin{figure}
-\centerline{\includegraphics[width=.35\textwidth]{img/times_sina_figure5.pdf}}
+\centerline{\includegraphics[width=.5\textwidth]{img/times_figure5.pdf}}
 \caption{
-Computing time of the bias correction methods. Each dot corresponds to
+Computing time of the bias correction methods. A. Total time for the entire genome at 100 kbp resolution. Each dot corresponds to
 one sample. The only method faster than ours under performs in all sample
-comparisons. Note the logarithmic scale on the y-axis.}
+comparisons. B. Time to correct a reduced genome (chr19-22) of one sample at different resolutions. Note the logarithmic scale on the y-axis on both panels.}
 \label{fig:times}
@@
 data collected on samples with aberrant karyotypes, but it applies
-seemlessly to the case of normal karyotypes. We showed that OneD performs
+seamlessly to the case of normal karyotypes. We showed that OneD performs
 significantly better than other methods when the cells present karyotypic
-aberrations (Figure~\ref{fig:aberrant}), and that it performs equall well
+aberrations (Figure~\ref{fig:aberrant}), and that it performs equally well
 as the best methods otherwise (Figure~\ref{fig:normal}). We also showed
@@
 Depending on the intention of the user, the effect of the copy number
-should either be kept or removed. This is why OneD leaves does not perform
+should either be kept or removed. This is why OneD does not perform
 the correction by default, but allows the user to obtain a
-diploid-equivalent Hi-C map computed from a hidden Marko model. The
+diploid-equivalent Hi-C map computed from a hidden Markov model. The
 resulting matrices have a near constant amount of contacts per bin, but

### 2017-06-05 main.tex
fixing small typos

@@
 \textit{vanilla} and the slowest is \textit{lgf}, with an over 100-fold
-span between the two. OneD also scales well in terms of bin size, as can
+span between the two. OneD also scales well in terms of bin size, as it can
 be seen in Figure~\ref{fig:times}B, where the different corrections were

### 2017-06-06 main.tex
Wrote the section about copy number variation.

@@
 
-\begin{equation*}
+\begin{equation}
+\label{eq:cnvnorm}
 \hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}
-\end{equation*}
+\end{equation}
 
@@
 
+\subsection{Copy number correction}
+
+Removing the explicit experimental biases enhancers the importance of the
+copy number in the signal (Figure~\ref{fig:copy_number}). The copy number
+is not an experimental bias, but it may be considered an additional source
+of bias to be remove. For this reason, OneD also allows the user to
+correct the copy number and output a matrix equivalent to that of a normal
+karyotype.
+
+For this application, OneD segments the linear signal of the total amount
+of contacts into piecewise homoegenous regions. This is carried out by a
+hidden Markov model whereby the averages of the states are constrained to
+be within a discrete factor of each other. This allows the model to detect
+regions with a number of copies equal to 1, 2, 3 \textit{etc}. With the
+inferred number of copies at hand, OneD then normalizes each cell of the
+matrix with equation (\ref{eq:cnvnorm}).
+
+Figure~\ref{fig:cnv_correction} shows the effect of this process on the
+final Hi-C matrix. Panel A indicates that different normalization methods
+will either enhancer or diminish the signal in regions with higher copy
+number. In this example, ICE overcompensates the original bias at the
+center of the picture and fades the signal alost entirely. Concomitantly,
+the signal at the bottom left of the matrix is enhanced to make appear a
+structure not visible in the raw data. On the contrary, \textit{lgf}
+strengthen the central region and the diagonal. OneD reduces the level of
+the central portion by a factor 2 approximately, but does otherwise not
+distort the main features of the region. This example shows that the copy
+number does not have a simple and predictable effect on the final matrix.
+Not taking it into account may open the door to some normalization
+artifacts. The result provided by OneD is not necessarily the right one
+(see Discussion), but at least it does not compensate copy number
+variations as a side effect of some other criteria.
+
+\begin{figure}
+\centerline{\includegraphics[width=.5\textwidth]
+  {img/figure_cnv_correction.pdf}}
+\caption{
+Copy number correction A. Detail of a Hi-C matrix normalized with
+different methods. The central portion has an increased copy number, which
+affects normalization. ICE fades it away, lgf enhances it and OneD reduces
+the signal by about half. B. Profile of the total amount of contacts after
+copy numer correction. The plot shows the same region as
+Figure\ref{fig:totals}B. The brown line represents the raw signal, the
+blue line represents the signal after bias correction, the black line
+represents the signal after bias and copy number correction.}
+\label{fig:cnv_correction}
+\end{figure}
+
+Figure~\ref{fig:cnv_correction}B shows the total sum of contacts after
+correction for the copy number with OneD. The signal that originally
+showed a sharp transition corresponding to the gain or loss of a large
+region, now shows a flat trend throughout the chromosome.
+
+In summary, OneD can be used to obtain a normalized matrix equivalent to
+that of a normal karyotype in cases that the effect of the copy number
+must be removed from the signal.
+
 \subsection{Speed}
@@
 \begin{figure}
-\centerline{\includegraphics[width=.5\textwidth]{img/times_figure5.pdf}}
+\centerline{\includegraphics[width=.5\textwidth]
+  {img/figure_benchmark_time.pdf}}
 \caption{

### 2017-06-07 main.tex

--- 2017-06-05 main.tex
@@
 1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
-Cuartero\,$^{\text{\sfb 1,2}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$ and
-Guillaume Fillion\,$^{\text{\sfb 1,2}}$}
+Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Miguel
+Beato\,$^{\text{\sfb 1,2}}$ and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}
 
@@
 essential component of their biological activity. The advent of the Hi-C
-technology enabled an unprecedented advance in our understanding of genome
-structures. However, Hi-C is subject to systematic biases that can
+technology enabled an unprecedented progress in our understanding of
+genome structures. However, Hi-C is subject to systematic biases that can
 compromise downstream analyses. Several strategies have been proposed to
-remove those biases, but none of them addresses the common issue of
-abnormal karyotypes. Many experiments are performed in cancer cell lines,
-which typically harbor large-scale copy number variations that create
-visible defects on the raw Hi-C maps. The consequences of these widespread
-artifacts on the normalized maps are mostly unexplored.\\
+remove those biases, but the issue of abnormal karyotypes received little
+attention. Many experiments are performed in cancer cell lines, which
+typically harbor large-scale copy number variations that create visible
+defects on the raw Hi-C maps. The consequences of these widespread
+artifacts on the normalized maps are mostly unexplored. \\
 	\textbf{Results:}
-We observed that current normalization methods perform badly in the
-presence of large-scale copy number variations, obscuring biological
-variations and enhancing batch effects. To address this issue, we
-developed an alternative approach designed to take into account chromosomal
-abnormalities. The method, called OneD, increases reproducibility among
-replicates of Hi-C samples with abnormal karyotype, thus significantly
-outperforming previous methods.  In the absence of chromosomal
-aberrations, OneD fared equally well as state-of-the-art methods, making
-it a safe choice for Hi-C normalization. OneD is fast and scales well in
-terms of computing resources for resolutions up to 1 kbp.\\	
-\textbf{Availability:} OneD is implemented as an R package available at
-\href{http://www.github.com/qenvio/dryhic}
-{http://www.github.com/qenvio/dryhic}.\\	
+We observed that current normalization methods are not robust to the
+presence of large-scale copy number variations, potentially obscuring
+biological variations and enhancing batch effects. To address this issue,
+we developed an alternative approach designed to take into account
+chromosomal abnormalities. The method, called \textit{OneD}, increases
+reproducibility among replicates of Hi-C samples with abnormal karyotype,
+thus significantly outperforming previous methods.  In the absence of
+chromosomal aberrations, \textit{OneD} fared equally well as
+state-of-the-art methods, making it a safe choice for Hi-C normalization.
+\textit{OneD} is fast and scales well in terms of computing resources for
+resolutions up to 1 kbp. \\
+\textbf{Availability:} \textit{OneD} is implemented as an R package
+available at \href{http://www.github.com/qenvio/dryhic}
+{http://www.github.com/qenvio/dryhic}. \\
 \textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
@@
 their activity \citep{rowley2016three}. In mammals, this organization
-plays a key role in guiding enhancer-promoter contacts (CITE), in V(D)J
-recombination (CITE) and in X chromosome inactivation (CITE). One of the
-most significant breakthroughs towards this insight was the development of
-the high throughput chromosomal conformation capture technology (Hi-C),
-assaying chromosomal contacts at a genome-wide scale
+plays a key role in guiding enhancer-promoter contacts
+\citep{de2013topology}, in V(D)J recombination \citep{choi2014ctcf} and in
+X chromosome inactivation \citep{galupa2015x}.
+
+One of the most significant breakthroughs towards this insight was the
+development of the high throughput chromosomal conformation capture
+technology (Hi-C), assaying chromosomal contacts at a genome-wide scale
 \citep{lieberman2009comprehensive}. Nowadays, exploring the spatial
-organization of chromatin has become a priority in many fields and
-Hi-C has become part of the standard molecular biology toolbox (CITE).
+organization of chromatin has become a priority in many fields and Hi-C
+has become part of the standard molecular biology toolbox
+\citep{denker2016second}.
 
@@
 sequencing reads have been shown to impact the results
-\citep{yaffe2011probabilistic}. In addition, the general experimental
-biases such as batch effects and protocol variation also apply. It is thus
-important to normalize Hi-C data in order to remove biases and artifacts,
-so that they are not confused with biological signal.
+\citep{yaffe2011probabilistic}. In addition, general experimental biases
+such as batch effects. It is thus important to normalize Hi-C data in
+order to remove biases and artifacts, so that they are not confused with
+biological signal.
 
@@
 explicitly from a defined set of local genomic features, such as the G+C
-content and the availability of restriction sites. This approach is used
-in the method of \cite{yaffe2011probabilistic} and in Hicnorm by
-\cite{hu2012hicnorm}. The second strategy is to implicitly correct
-unknown biases by enforcing some regularity on the data. This approach is
-used in the ICE method of \cite{imakaev2012iterative}, where the total
-amount of contacts of every bin is imposed to be the same. ICE is
-currently the most popular method, due in part to its speed.
+content. This approach is used in the method of
+\cite{yaffe2011probabilistic} and in Hicnorm by \cite{hu2012hicnorm}. The
+second strategy is to implicitly correct unknown biases by enforcing some
+regularity condition on the data. This approach is used in the
+Iterative Correction and Eigenvector decomposition method (\textit{ICE})
+of \cite{imakaev2012iterative}, whereby the total amount of contacts of
+every bin is imposed to be the same. \textit{ICE} is currently the most
+popular method, due in part to its speed.
 
 Neither of these strategies were designed with regard for cell types with
-karyotypic aberrations, most common in cancer cells in culture.  Hi-C is
-very sensitive to aneuploidy, copy number variations and translocations.
+karyotypic aberrations, most common in cancer cells. Hi-C is very
+sensitive to aneuploidy, copy number variations and translocations.
 Actually, these aberrations have so much influence on the outcome that the
@@
 \citep{korbel2013genome}. So far the only attempt to address the issue
-was the correction method caICB by \cite{wu2016computational}. However,
-caICB applies a chromosome-wide correction, effectively excluding the
-numerous cases of partial aneuploidy and regional copy number variations.
+was the chromosome-adjusted Iterative Correction Bias method
+(\textit{caICB}) of \cite{wu2016computational}.  However, caICB applies a
+uniform chromosome-wide correction, effectively excluding the numerous
+cases of partial aneuploidy and regional copy number variations.
 
 Here we propose \textit{OneD}, a method to correct local chromosomal
-abnormalities. OneD explicitly models the contribution of known biases via
-a generalized additive model. The normalized data is more reproducible
-between replicates and across different protocols. Importantly, OneD is
-also applicable when cells have a normal karyotype, where it performs as
-well as the best normalization methods. Finally, the implementation is
-faster than ICE and scales up to 1 kbp resolution with reasonable
+abnormalities in Hi-C experiments. \textit{OneD} explicitly models the
+contribution of known biases via a generalized additive model. The
+normalized data is more reproducible between replicates and across
+different protocols. Importantly, \textit{OneD} is also efficient when
+cells have a normal karyotype, where it performs as well as the best
+normalization methods. Finally, the implementation is as fast as
+\textit{ICE} and it scales up to 1~kbp resolution with reasonable
 computing resources.
@@
 Our approach is to model the tally of contacts for each bin, thus reducing
-the matrix to a one dimension score (hence the name OneD). We assume that
+the matrix to a one dimension score (hence the name \textit{OneD}). We assume that
 the total number of contacts per bin ($t_{i}$) can be approximated by a
@@
 \begin{table*}
-\processtable{Hi-C data sets used in this study.
+\processtable{Hi-C data sets used in this study. Quique you're here.
 \label{tab:samples}}
@@
 
-\subsection{Data processing}
+%\subsection{Data processing}
 
@@
 efficiently create sparse representations of contact matrices and further
-apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections.
+apply \textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
 \texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
-\citep{wu2016computational} were used to carry out the \textit{lgf} and
-\textit{caib} corrections respectively. All the results are based on a
+\citep{wu2016computational} were used to carry out the \textit{LGF} and
+\textit{caICB} corrections respectively. All the results are based on a
 resolution of 100 kbp, but we found no major differences for different
@@
 
-We benchmarked \textbf{\textit{oned}}, against \textbf{\textit{vanilla}},
-\textbf{\textit{ice}} \citep{imakaev2012iterative}, \textbf{\textit{caib}}
-\citep{wu2016computational} and \textbf{\textit{lgf}}
-\citep{hu2012hicnorm, servant2012hitc}. The first three methods correct
-biases implicitly, whereas the fourth method does it explicitly.
-
 To measure the gain or loss of similarity upon normalization, we compared
@@
 fixed effect was the normalization method and the random effect was the
-chromosome. Receiver operating characteristic (ROC) curves were then
+%chromosome. Receiver operating characteristic (ROC) curves were then
 computed using \texttt{ROCR} package \citep{sing2005rocr}.
@@
 
-\subsection{Correction method}
+\subsection{Experimental bias correction}
 
-The principle of OneD is to explicitly model Hi-C biases on a single
-variable: the total amount of contacts for each bin of the matrix (see
-\ref{sec:model}). The reason for this choice is that the total amount of
-contacts is approximately proportional to the local copy number. For
-instance, a duplicated region in a diploid genome will show on average a
-50\% increase in the number of contacts. Discontinuities of the amount of
-contacts thus correspond to changes of the copy number.
+The principle of \textit{OneD} is to explicitly model Hi-C biases on a
+single variable: the total amount of contacts for each bin of the matrix
+(see \ref{sec:model} for detail). The reason for this choice is that the
+total amount of contacts is approximately proportional to the local copy
+number. For instance, a duplicated region in a diploid genome will show on
+average a 50\% increase in the number of contacts. Discontinuities of the
+amount of contacts thus correspond to changes of the copy number.
 
-\begin{figure}[!tp]
+\begin{figure}
 \centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
@@
 
-Other biases affect the total amount of contacts in a continuous but not
-necessarily linear way. Figure~\ref{fig:totals}A shows the relationship
-between the amount of contacts and the number of restriction enzyme sites
-in T47D, a breast cancer cell line with aberrant and unstable karyotype.
-Four clouds are visible. Each corresponds to a copy number between one and
-four. In all of them, the relationship flattens as the number of
-restriction sites increases. For this reason, OneD fits a non-linear
-relationship between the total amount of contacts and the known sources of
-bias (by default the G+C content, the number of restriction sites and the
-mappability of the reads).
+Experimental biases affect the total amount of contacts in a continuous
+but not necessarily linear way. Figure~\ref{fig:totals}A shows the
+relationship between the amount of contacts and the number of restriction
+enzyme sites in T47D, a breast cancer cell line with aberrant and unstable
+karyotype.  Four clouds are visible. Each corresponds to sequences present
+in one to four copies. In all of them, the relationship flattens as the
+number of restriction sites increases. In order to capture this
+relationship, \textit{OneD} fits a non-linear model between the total
+amount of contacts and the known sources of bias (by default the G+C
+content, the number of restriction sites and the mappability of the
+reads).
 
-The biases are estimated genome-wide and each cell of the matrix is then
-corrected using equation (\ref{eq:xhat}). Note that the corrected amount
-of contacts is still proportional to the copy number. The correction thus
-represents a more stable and precise estimation of the copy number.
+The experimental biases are estimated genome-wide and each cell of the
+matrix is then corrected using equation (\ref{eq:xhat}). Note that the
+corrected amount of contacts is still proportional to the copy number.
 Figure~\ref{fig:totals}B shows the corrected number of contacts along
-chromosome 17 of a T47D. OneD greatly reduces the wiggling of the total
-amount of contacts.
+chromosome 17 of T47D. \textit{OneD} greatly reduces the wiggling of the
+total amount of contacts.
 
 We tested the validity of this approach against the Catalogue Of Somatic
-Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure
-\ref{fig:copy_number} shows the Pearson correlation between the corrected
-number of contacts and the copy number estimation for both T47D and K562
-cell lines. Similar results were obtained using Spearman correlation
-(Supplementary Figure 2). All the methods except OneD decrease the
-agreement with the copy number because they partially correct this bias.
-In contrast, OneD enhances the conformity of the signal with the copy
-number. Not correcting for variable copy number at that stage may seem to
-defeat the purpose of a normalization method, but we will see below how
-this can lead to better performance.
+Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. In what follows,
+we benchmarked \textit{OneD}, against \textit{vanilla}, \textit{ICE}
+\citep{imakaev2012iterative}, \textit{caICB} \citep{wu2016computational}
+and the Local Genomic Features method
+\citep[\textit{LGF},][]{hu2012hicnorm, servant2012hitc}. The first three
+methods correct biases implicitly, whereas the fourth method does it
+explicitly. Figure \ref{fig:copy_number} shows the Pearson correlation
+between the corrected number of contacts and the copy number estimation
+for both T47D and K562 cell lines. Similar results were obtained using
+Spearman correlation (Supplementary Figure 2). All the methods except
+\textit{OneD} decrease the agreement between the signal and the copy
+number because they partially correct the discrepancy. In contrast,
+\textit{OneD} enhances the conformity of the signal with the copy number.
+Not correcting for variable copy number at that stage may seem
+counter-intuitive, but the tests below will show this leads to better
+performance.
 
-\begin{figure}
+\begin{figure}[b]
 	\centerline{\includegraphics[width=.45\textwidth]
@@
 leukemia cell line. The new proposal (in blue) outperforms the rest of
-alternatives.  }\label{fig:copy_number}
+alternatives.}
+\label{fig:copy_number}
 \end{figure}
@@
 
-We first benchmarked the performance of OneD on biological samples with an
-aberrant karyotype. A good normalization method should increase the
-similarity between biological replicates by reducing irrelevant
-experimental variation, such as batch effects, laboratory of origin and
-protocol variations. Similarly, a good normalization should decrease the
-similarity between samples that received a different treatment in order to
-enhance the biological differences.
+We first benchmarked the performance of our approach on biological
+samples with an aberrant karyotype. A good normalization method should
+increase the similarity between biological replicates by reducing
+irrelevant experimental variation, such as batch effects, laboratory of
+origin and protocol variations. Similarly, a good normalization should
+decrease the similarity between samples that received a different
+treatment in order to enhance the biological differences.
 
 We assembled two Hi-C data sets obtained from T47D and K562, two cancer
-cell lines with aberrant karyotypes. In each set we spiked two
-samples from the other cell line (see Table~\ref{tab:samples}) in order to
-introduce biological variability. We measured the Pearson correlation of
-observed over expected counts to compare matrices before and after
-normalization by different methods (see \ref{sec:comp}). This gave an
+cell lines with aberrant karyotypes. In each set we spiked two samples
+from the other cell line (see Table~\ref{tab:samples}) in order to
+introduce biological variability. We compared matrices before and after
+normalization by different methods using the Pearson correlation of
+observed over expected counts (see \ref{sec:comp}). This gave an
 indication of the impact of a given normalization method. The results are
@@
 
-The \textit{caib} and \textit{ice} methods increased the similarity
+The \textit{caICB} and \textit{ICE} methods increased the similarity
 between the different cell lines (Figures~\ref{fig:aberrant}A and
@@
 method \textit{vanilla} followed the same trend but to a lesser extent,
-consistent with the fact that it consists of a single \textit{ice}
-iteration.
-
-OneD was the only method to increase the similarity between experiments
-carried out on the same material but with a different protocol
-(Figure~\ref{fig:aberrant}A). In these conditions, \textit{lgf} was
-decreasing the similarity. Taken together, the results show that OneD
-enhanced biological variation while decreasing experimental variation.
+consistent with the fact that it consists of a single \textit{ICE}
+iteration. \textit{OneD} was the only method to increase the similarity
+between experiments carried out on the same material but with a different
+protocol (Figure~\ref{fig:aberrant}A).
 
@@
 All the methods, including the absence of normalization, succeeded in
-recognizing the T47D outliers among the K562 samples, but recognizing the
-K562 outliers among the T47D samples proved more challenging. OneD
-increased the discrimination power compared to the raw matrices, but all
-the other methods decreased it. Actually, they performed little better
-than random on this task.
+identifying the T47D outliers among the K562 samples, but recognizing the
+K562 outliers among the T47D samples proved more challenging.
+\textit{OneD} increased the discrimination power compared to the raw
+matrices, but all the other methods decreased it. Actually, they performed
+little better than random on this task.
 
@@
 
-Taken together, these results show that OneD enhances biological variation
-and reduces experimental noise on samples with an aberrant karyotype.
+Taken together, these results show that correcting experimental biases
+with \textit{OneD} enhances the biological variation and reduces the noise
+on samples with an aberrant karyotype.
 
 \begin{figure}
-\centerline{\includegraphics[width=.50\textwidth]{img/correlation_aberrant_figure3.pdf}}
+\centerline{\includegraphics[width=.50\textwidth]
+  {img/correlation_aberrant_figure3.pdf}}
 \caption{
@@
 
-Does the performance of OneD with aberrant karyotypes comes at the cost of
-decreased performance with normal karyotypes? To address this question, we
-assembled another data set comprised of normal mouse B cells and embryonic
-stem (ES) cells. The cell types were pooled in almost equal proportions
-(see Table~\ref{tab:samples}) and the same tests as above were performed.
+Does the performance of \textit{OneD} for aberrant karyotypes come at the
+cost of decreased performance for normal karyotypes? To address this
+question, we assembled another data set comprised of mouse B cells and
+embryonic stem (ES) cells, both with a normal karyotype The cell types
+were pooled in almost equal proportions (see Table~\ref{tab:samples}) and
+the same tests as above were performed.
 
-In these conditions, variations of the protocol had a strong effect on the
+In these conditions, the experimental protocol had a strong effect on the
 impact of the different normalization methods (Figure~\ref{fig:normal}A).
-For instance, \textit{caib} and \textit{ice} increased the similarity when
-the protocols were different, but decreased it when the protocols were the
-same. The effect was stronger when comparing identical cell types, but the
-same trend appeared when comparing different cell types, indicating that
-these methods may enhance or reduce biological variation, depending on the
-context. Once more, \textit{vanilla} followed the same trend as
-\textit{ice} to a lesser extent. The \textit{lgf} method increased the
-similarity when comparing the same cells with different experimental
-protocols, and had little to no effect in the other cases. This indicates
-that \textit{lgf} is a safe choice in this case.
+For instance, \textit{caICB} and \textit{ICE} increased the similarity
+when the protocols were different, but decreased it when the protocols
+were the same. The effect was stronger when comparing identical cell
+types, but the same trend appeared when comparing different cell types,
+indicating that these methods may enhance or reduce biological variation,
+depending on the context. Once more, \textit{vanilla} followed the same
+trend as \textit{ICE} to a lesser extent. The \textit{LGF} method
+increased the similarity when comparing the same cells with different
+experimental protocols, and had little to no effect in the other cases.
+This indicates that \textit{LGF} is a safe choice in this case.
 
-OneD decreased the similarity between different cell types when using the
-same protocol and increased it between identical cell types when using
-different protocols. In this other two cases, it had little effect. In
-summary, OneD never enhanced the experimental noise and it reduced it in
-one more case than \textit{lgf}.
+\textit{OneD} decreased the similarity between different cell types when
+using the same protocol and increased it between identical cell types when
+using different protocols. In the other two cases, it had little effect.
+In summary, \textit{OneD} never enhanced the experimental noise and it
+reduced it in one more case than \textit{LGF}.
 
@@
 biological pairs, after which their performance diverged
-(Figure~\ref{fig:normal}B). OneD achieved the highest area under the curve
-on this problem, even though it was not always above the others methods at
-every position of the graph.
+(Figure~\ref{fig:normal}B). \textit{OneD} achieved the highest area under
+the curve on this problem, but with a small margin over the other methods
+except \textit{vanilla}.
 
 Using other metrics to compare matrices gave similar results
-(Supplementary Figures 6 and 7). OneD was always among the top scoring
-methods. In these conditions, Spearman correlation of contacts again
-appeared as as the worst comparison metric as it showed a lower
+(Supplementary Figures 6 and 7). \textit{OneD} was always among the top
+scoring methods. In these conditions, Spearman correlation of contacts
+again appeared as the worst comparison metric because it showed a lower
 performance for all the normalization methods.
 
-Taken together, these results indicate that OneD performs as well as the
-best normalization methods even with normal karyotypes.
+Taken together, these results indicate that \textit{OneD} performs as well
+as the best normalization methods, even with normal karyotypes.
 
 \begin{figure}
-\centerline{\includegraphics[width=.50\textwidth]{img/correlation_normal_figure4.pdf}}
+\centerline{\includegraphics[width=.50\textwidth]
+  {img/correlation_normal_figure4.pdf}}
 \caption{
@@
 \end{figure}
+
+\subsection{Copy number correction}
+
+Removing the explicit experimental biases enhances the importance of the
+copy number in the signal (Figure~\ref{fig:copy_number}). The copy number
+is not an experimental bias, but it may be considered an additional source
+of bias to be remove. For this reason, \textit{OneD} also allows the user
+to correct the copy number and output a euploid-equivalent matrix.
+
+Here, \textit{OneD} segments the linear signal of the total amount of
+contacts into piecewise homoegenous regions. This is carried out by a
+hidden Markov model whereby the averages of the states are constrained to
+be within a discrete factor of each other. This allows the model to detect
+regions with a number of copies equal to 1, 2, 3 \textit{etc}. With the
+inferred number of copies at hand, \textit{OneD} then normalizes each cell
+of the matrix with equation (\ref{eq:cnvnorm}).
+
+Figure~\ref{fig:cnv_correction} shows the effect of this process on the
+final Hi-C matrix. Panel \ref{fig:cnv_correction}A indicates that
+different normalization methods can either enhance or diminish the signal
+in regions with higher copy number. In this example, \textit{ICE}
+overcompensates the original bias at the center of the picture and fades
+the signal almost entirely. Concomitantly, the signal at the bottom left
+of the matrix is enhanced and shows a structure that is not visible in the
+raw data. On the contrary, \textit{LGF} strengthen the central region and
+the diagonal. \textit{OneD} reduces the level of the central portion by a
+factor 2 approximately, but does not otherwise distort the main features
+of the region. This example shows that the copy number does not have a
+simple and predictable effect on the final matrix. Not taking it into
+account may open the door to some normalization artifacts. The result
+provided by \textit{OneD} is not necessarily the right one (see
+Discussion), but at least it does not correct copy number variations
+as a side effect of some other criteria.
+
+\begin{figure}
+\centerline{\includegraphics[width=.5\textwidth]
+  {img/figure_cnv_correction.pdf}}
+\caption{
+Copy number correction A. Detail of a Hi-C matrix normalized with
+different methods. The central portion has an increased copy number, which
+affects normalization. \textit{ICE} fades it away, \textit{LGF} enhances
+it and \textit{OneD} reduces the signal by about half. B. Profile of the
+total amount of contacts after copy numer correction. The plot shows the
+same region as Figure\ref{fig:totals}B. The brown line represents the raw
+signal, the blue line represents the signal after bias correction, the
+black line represents the signal after bias and copy number correction.}
+\label{fig:cnv_correction}
+\end{figure}
+
+Figure~\ref{fig:cnv_correction}B shows the total sum of contacts after
+correction for the copy number with \textit{OneD}. The signal that
+originally showed a sharp transition corresponding to the gain or loss of
+a large region, now shows a flat trend throughout the chromosome.
+
+In summary, \textit{OneD} is the only method that can be used to obtain a
+euploid-equivalent normalized matrix in cases that the effect of the copy
+number must be removed from the signal.
 
@@
 
-Unlike the other methods, OneD corrects a single variable instead of the
-whole matrix. It uses only the total number of contacts per bin and thus
-estimates the model parameters much faster than previous explicit methods.
-We measured the running time of the different tools on a 3.3~GHz machine
-with 62 GB~RAM, always using the default parameters. Figure~\ref{fig:times}A
-shows the running times of the different methods on the samples described
-in Table~\ref{tab:samples} at 100 kbp resolution. The fastest method is
-\textit{vanilla} and the slowest is \textit{lgf}, with an over 100-fold
-span between the two. OneD also scales well in terms of bin size, as can
-be seen in Figure~\ref{fig:times}B, where the different corrections were
-applied to a reduced part of the genome at increasing resolutions.
+Unlike the other methods, \textit{OneD} corrects a single variable instead
+of the whole matrix, and thus estimates the model parameters much faster
+than previous explicit methods. We measured the running time of the
+different tools on a 3.3~GHz machine with 62 GB~RAM, always using the
+default parameters. Figure~\ref{fig:times}A shows the running times of the
+different methods on the samples described in Table~\ref{tab:samples} at
+100 kbp resolution. The fastest method is \textit{vanilla} and the slowest
+is \textit{LGF}, with an over 100-fold span between the two.
 
-OneD is the second fastest method and it always ran in less than 1 min in
-the conditions of the benchmark. In all cases, the memory footprint was
-less than 300 MB. Interestingly, the running time of OneD is much more
-homogeneous than that of the other methods. The reason is that the size of
-the regression problem to be solved by the \texttt{mgcv} package is always
-the same at fixed resolution
+\textit{OneD} is the second fastest method and it always ran in less than
+1 min in the conditions of the benchmark. We occasionally encountered
+problems where \textit{ICE} ran faster than \textit{OneD} so it is fair to
+say that none of them has a definite speed advantage. Throughout this
+benchmark, the memory footprint of \textit{OneD} was less than 300 MB.
+Interestingly, the running time of \textit{OneD} is much more homogeneous
+than that of the other methods. The reason is that the size of the
+regression problem to be solved by the \texttt{mgcv} package is always the
+same at fixed resolution.
 
-Taken together, these results show that OneD is very competitive in terms
-of computational speed compared to existing methods.
+At increasing resolution, the relative rank of the methods remained the
+same (Figure~\ref{fig:times}B). The running time of \textit{OneD} remained
+close to that of \textit{ICE}. Taken together, these results show that
+\textit{OneD} is competitive in terms of computational speed compared to
+existing methods.
 
 \begin{figure}
-\centerline{\includegraphics[width=.5\textwidth]{img/times_figure5.pdf}}
+\centerline{\includegraphics[width=.5\textwidth]
+  {img/figure_benchmark_time.pdf}}
 \caption{
-Computing time of the bias correction methods. A. Total time for the entire genome at 100 kbp resolution. Each dot corresponds to
-one sample. The only method faster than ours under performs in all sample
-comparisons. B. Time to correct a reduced genome (chr19-22) of one sample at different resolutions. Note the logarithmic scale on the y-axis on both panels.}
+Computing time of the bias correction methods. A. Total time for the
+entire genome at 100 kbp resolution. Each dot corresponds to one sample.
+The only method faster than ours under performs in all sample comparisons.
+B. Time to correct a reduced genome (chr19-22) of one sample at different
+resolutions. Note the logarithmic scale on the y-axis on both panels.}
 \label{fig:times}
@@
 
-Here we introduced OneD, a fast computational method to normalize Hi-C
-matrices. OneD was developed ground up to address the need to normalize
-data collected on samples with aberrant karyotypes, but it applies
-seamlessly to the case of normal karyotypes. We showed that OneD performs
-significantly better than other methods when the cells present karyotypic
-aberrations (Figure~\ref{fig:aberrant}), and that it performs equally well
-as the best methods otherwise (Figure~\ref{fig:normal}). We also showed
-that OneD is on average faster than \textit{ice}, which makes it very
-competitive from the point of view of computational speed.
+Here we introduced \textit{OneD}, a fast computational method to normalize
+Hi-C matrices. \textit{OneD} was developed ground up to address the need
+to normalize data collected on samples with aberrant karyotypes, but it
+applies seamlessly to the case of normal karyotypes. We showed that
+\textit{OneD} performs significantly better than other methods when the
+cells present karyotypic aberrations (Figure~\ref{fig:aberrant}), and that
+it performs equally well as the best methods otherwise
+(Figure~\ref{fig:normal}). We also showed that \textit{OneD} is
+approximately as fast as \textit{ICE}, which makes it competitive from the
+point of view of computational speed.
 
-The originality of OneD lies in that it projects all the biases onto a
-single variable: the total amount of contact per bin. This allows greater
-running speed, while preserving a good performance on samples with
-karyotypic aberrations. One of the reasons that OneD is better able to
-highlight the biological distinctions between samples is that it does not
-correct the copy number but enhances it (Figures~\ref{fig:totals} and
-\ref{fig:copy_number}).
+The originality of \textit{OneD} lies in that it projects all the biases
+onto a single variable: the total amount of contacts per bin. This allows
+greater running speed, while preserving a good performance on samples with
+karyotypic aberrations. One of the reasons that \textit{OneD} is better
+able to highlight the biological distinctions between samples is that it
+does only corrects the copy number if specifically requested by the user.
+Copy number variations treated as implicit biases have an
+opaque effects on the
 
@@
 Depending on the intention of the user, the effect of the copy number
-should either be kept or removed. This is why OneD does not perform
+should either be kept or removed. This is why \textit{OneD} does not perform
 the correction by default, but allows the user to obtain a
@@
 
-Overall, OneD constitutes a safe choice to normalize Hi-C matrices. If the
+Overall, \textit{OneD} constitutes a safe choice to normalize Hi-C matrices. If the
 karyotype of the sample is aberrant, it enhances the biological variation.

### 2017-06-07 main.tex
Polished and reviewed the text. Two figure legends needs attention.

@@
 1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
-Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Miguel
-Beato\,$^{\text{\sfb 1,2}}$ and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}
+Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Marc A.
+Mart\'i Renom\,$^{\text{\sfb 1,2}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
+and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}
 
@@
 \centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
-\caption{A. Non linear relationship between the number of restriction
-sites per bin and the total number of contacts per bin in T47D. B. Total
-number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
-corrected (blue) signals are shown. The region that corresponds to the
-first 20 Mbp of the chromosome is present in single copy in this clone of
-T47D, explaining that the signal is approximately half compared to the
-rest of the chromosome.  Also notice that the raw signal is noisier than
-the corrected one.}
+\caption{Total amount of contacts per bin. A. Non linear relationship
+between the number of restriction sites and the total number of
+contacts per bin in T47D. B. Total number of contacts per bin on
+chromosome 17 of T47D. Raw (brown) and corrected (blue) signals are shown.
+The long arm of chromosome 17 (the region corresponding to 20-40 Mbp) is
+present in four coipes, explaining that the signal is about twice higher
+than for the short arm.}
 \label{fig:totals}
@@
 {img/copy_number_figure2.pdf}}
-	\caption{
-Pearson correlation between total number of contacts per bin and
-independent copy number estimation (COSMIC) for each of the methods
-compared. Left panel T47D breast cancer cell line, right panel K562
-leukemia cell line. The new proposal (in blue) outperforms the rest of
-alternatives.}
+\caption{Pearson correlation between the total number of contacts per bin
+and an independent estimation of the copy number (COSMIC). Left panel
+T47D breast cancer cell line, right panel K562 leukemia cell line.}
 \label{fig:copy_number}
@@
 \caption{
-Results of the comparison between samples with aberrant karyotype. A and
-B. Average changes compared to raw on the T47D and K562 sets. The bars
-represent 95\% confidence intervals centered on the mean difference of the
-correlation score between a given correction method and the raw data. The
-brown dashed line indicates the value of the average score on raw matrices
-(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
-the curve are indicated in the bottom right corner. The color code is the
-same as panels A and B. The brown lines correspond to raw matrices. All
-results in this figure are based on Pearson correlations between the
-observed over expected counts.}
+Results of the comparison between samples with aberrant karyotypes. A and
+B. Average changes compared to the raw data on the T47D and K562 sets. The
+bars represent 95\% confidence intervals centered on the mean difference
+of the correlation score between a given correction method and the raw
+data. C and D. ROC curves on the T47D and K562 sets.  The areas under the
+curve are indicated in the bottom right corner. The color code is the same
+as panels A and B. The brown lines correspond to raw matrices. All results
+in this figure are based on Pearson correlations between the observed over
+expected counts.}
 \label{fig:aberrant}
@@
 \caption{
-Results of the comparison between samples with normal karyotype.  A.
-Average changes compared to raw on the mouse data set. The bars are as in
-B. ROC curves on the mouse data set. The legend is as in
+Results of the comparison between samples with normal karyotypes. A.
+Average changes compared to the raw data on the mouse data set.  B. ROC
+curves on the mouse data set. The legend is as in
 Figure~\ref{fig:aberrant}.}
@@
 Hi-C matrices. \textit{OneD} was developed ground up to address the need
-to normalize data collected on samples with aberrant karyotypes, but it
+to normalize data from biological samples with aberrant karyotypes, but it
 applies seamlessly to the case of normal karyotypes. We showed that
@@
 able to highlight the biological distinctions between samples is that it
-does only corrects the copy number if specifically requested by the user.
-Copy number variations treated as implicit biases have an
-opaque effects on the
+only corrects the copy number if specifically requested by the user. The
+impact of copy number variations on normalization is rather opaque,
+especially if they are treated as implicit biases
+(Figure~\ref{fig:cnv_correction}). Treating them as explicit biases with
+optional removal seems to be an overall safer strategy.
 
 This raises the question whether variations of the copy number constitute
-a biological signal or an artifact. Note that in the presence of
-translocations, deletions and large scale rearrangements, correcting the
-copy number is equally faulty as not correcting it. Indeed, in this case
-the reference genome on which the matrix is drawn is incorrect and
-changing the values of the pixels cannot account for this information.
+a biological signal or an artifact. If the biological sample contains
+karytoypic aberrations, then its genome is grossly different the reference
+so there can be no correct way to normalize the Hi-C signal. The proper
+approach would be to use the actual genome of the biological sample as a
+reference to construct the contact map. However, this strategy is
+presently unfeasible because assembling mammalian genomes is still a hard
+problem.
 
 Depending on the intention of the user, the effect of the copy number
-should either be kept or removed. This is why \textit{OneD} does not perform
-the correction by default, but allows the user to obtain a
-diploid-equivalent Hi-C map computed from a hidden Markov model. The
+should either be kept or removed. This is why \textit{OneD} does not
+perform the correction by default, but allows the user to obtain a
+euploid-equivalent Hi-C map computed from a hidden Markov model. The
 resulting matrices have a near constant amount of contacts per bin, but
-the artifacts caused by the misfit between the genome of the sample and
+the artifacts caused by the mismatch between the genome of the sample and
 the reference genome are still present (for instance, the artifacts caused
@@
 
-Overall, \textit{OneD} constitutes a safe choice to normalize Hi-C matrices. If the
-karyotype of the sample is aberrant, it enhances the biological variation.
-If not, it performs at least equally well as other methods with an
-advantage in terms of computational speed.
+Overall, \textit{OneD} constitutes a safe choice to normalize Hi-C
+matrices. If the karyotype of the sample is aberrant, it enhances the
+biological variation.  If not, it performs at least equally well as other
+methods in terms of quality and of computational speed.
 

### 2017-06-07 main.tex
Figures fixed, supp. figures added and table fixed

@@
 \begin{table*}
-\processtable{Hi-C data sets used in this study. Quique you're here.
+\processtable{Hi-C data sets used in this study. Sample ID: Unique sample
+identifier. Cell type: Type of cell. Application: Hi-C protocol. RE: Restriction
+enzyme used for digestion. Sequencing core: Laboratory performing the experiment
+(CRG: Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor: Baylor College of Medicine;
+UCSD: University of California San Diego; LICR: Ludwig Institue for Cancer Research). Set(s): Comparison set. Source: SRA, GEO
+or ENCODE raw data identifier.
 \label{tab:samples}}
@@
 GSM987817 & B-cell  & dilution Hi-C &
-  HindIII & UCSC & mm10 & SRR543428-SRR543431 \\
+  HindIII & UCSD & mm10 & SRR543428-SRR543431 \\
 GSM987818 & B-cell  & dilution Hi-C &
-  HindIII & UCSC & mm10 & SRR543432-SRR543442 \\
+  HindIII & UCSD & mm10 & SRR543432-SRR543442 \\
 GSM862720 & ES-cell & dilution Hi-C &
-  HindIII & UCSC & mm10 & SRR443883-SRR443885 \\
+  HindIII & LICR & mm10 & SRR443883-SRR443885 \\
 GSM862721 & ES-cell & dilution Hi-C &
-  HindIII & UCSC & mm10 & SRR400251-SRR400256 \\
+  HindIII & LICR & mm10 & SRR400251-SRR400256 \\
 GSM862722 & ES-cell & dilution Hi-C &
-  NcoI    & UCSC & mm10 & SRR443886-SRR443888 \\
+  NcoI    & LICR & mm10 & SRR443886-SRR443888 \\
   \botrule
@@
 for both T47D and K562 cell lines. Similar results were obtained using
-Spearman correlation (Supplementary Figure 2). All the methods except
+Spearman correlation (Supplementary Figure 1). All the methods except
 \textit{OneD} decrease the agreement between the signal and the copy
@@
 between the different cell lines (Figures~\ref{fig:aberrant}A and
-\ref{fig:aberrant}B and Supplementary Figure~3). This is an undesirable
+\ref{fig:aberrant}B and Supplementary Figures~2 and 3). This is an undesirable
 effect, as it obscures the biological variability. In the same vein, these
@@
 Using the other metrics described in \ref{sec:comp} yielded similar
-results (Supplementary Figures 3, 4 and 5). Note that Spearman correlation
+results (Supplementary Figures 5, 6 and 7). Note that Spearman correlation
 of contacts presented the worst performance for all scenarios, and it thus
@@
 Using other metrics to compare matrices gave similar results
-(Supplementary Figures 6 and 7). \textit{OneD} was always among the top
+(Supplementary Figures 7 and 8). \textit{OneD} was always among the top
 scoring methods. In these conditions, Spearman correlation of contacts

### 2017-06-07 supplementary_figures.tex
Figures fixed, supp. figures added and table fixed

\PassOptionsToPackage{utf8}{inputenc}
\documentclass[12pt]{report}

\usepackage[margin=2cm]{geometry}

\usepackage{graphicx}

\renewcommand{\thefigure}{S\arabic{figure}}

\begin{document}

\begin{figure}[b]
	\centerline{\includegraphics[width=\textwidth]{img/copy_number_suppfigure2.pdf}}
	\caption{
Spearman correlation between total number of contacts per bin and
independent copy number estimation (COSMIC) for each of the methods
compared. Left panel T47D breast cancer cell line, right panel K562
leukemia cell line. The new proposal (in blue) outperforms the rest of
alternatives.}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{img/correlation_aberrant_boxplots1.pdf}}
\caption{Boxplots representing the results of the pair-wise comparisons of the
	samples included in the T47D set. X axis: different correction methods. Y
    axis: correlation. Each panel groups paris of samples with the correspondind
    characteristics (in terms of cell type, protocol, batch and
    treatment). A. Pearson correlation between observed over expected
    counts. B. Spearman correlation between observed counts. C. Stratum-adjusted correlation
    coefficient (SCC) between observed counts. D. Reproducibility score of
    observed counts.}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{img/correlation_aberrant_boxplots2.pdf}}
\caption{Boxplots representing the results of the pair-wise comparisons of the
	samples included in the K562 set. X axis: different correction methods. Y
    axis: correlation. Each panel groups paris of samples with the correspondind
    characteristics (in terms of cell type, protocol, batch and
    treatment). A. Pearson correlation between observed over expected
    counts. B. Spearman correlation between observed counts. C. Stratum-adjusted correlation
    coefficient (SCC) between observed counts. D. Reproducibility score of
    observed counts.}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{img/correlation_aberrant_spearman.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotype. A and
B. Average changes compared to raw on the T47D and K562 sets. The bars
represent 95\% confidence intervals centered on the mean difference of the
correlation score between a given correction method and the raw data. The
brown dashed line indicates the value of the average score on raw matrices
(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
the curve are indicated in the bottom right corner. The color code is the
same as panels A and B. The brown lines correspond to raw matrices. All
results in this figure are based on Spearman correlations between the
observed counts.}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{img/correlation_aberrant_scc.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotype. A and
B. Average changes compared to raw on the T47D and K562 sets. The bars
represent 95\% confidence intervals centered on the mean difference of the
correlation score between a given correction method and the raw data. The
brown dashed line indicates the value of the average score on raw matrices
(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
the curve are indicated in the bottom right corner. The color code is the
same as panels A and B. The brown lines correspond to raw matrices. All
results in this figure are based on stratum-adjusted correlations between the
observed counts.}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{img/correlation_aberrant_repro.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotype. A and
B. Average changes compared to raw on the T47D and K562 sets. The bars
represent 95\% confidence intervals centered on the mean difference of the
correlation score between a given correction method and the raw data. The
brown dashed line indicates the value of the average score on raw matrices
(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
the curve are indicated in the bottom right corner. The color code is the
same as panels A and B. The brown lines correspond to raw matrices. All
results in this figure are based on the reproducibiliry score between the
observed counts.}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{img/correlation_normal_boxplots.pdf}}
\caption{Boxplots representing the results of the pair-wise comparisons of the
	samples included in the mm10 set. X axis: different correction methods. Y
    axis: correlation. Each panel groups paris of samples with the correspondind
    characteristics (in terms of cell type, protocol, batch and
    treatment). A. Pearson correlation between observed over expected
    counts. B. Spearman correlation between observed counts. C. Stratum-adjusted correlation
    coefficient (SCC) between observed counts. D. Reproducibility score of
    observed counts.}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=.75\textwidth]{img/correlation_normal_other_metrics.pdf}}
\caption{Results of the comparison between samples with normal karyotype. A., C.
    and E.  Average changes compared to raw. The bars
represent 95\% confidence intervals centered on the mean difference of the
correlation score between a given correction method and the raw data. The
brown dashed line indicates the value of the average score on raw matrices
(set to 0). B., D. and F. ROC curves. The areas under
the curve are indicated in the bottom right corner. The color code is the
same as panels A, C. B. The brown lines correspond to raw matrices. Results in
panels A. and B. on Spearman correlations between the observed counts. Results in
panels C. and D. on Stratum-adjusted correlation coefficient (SCC) between observed counts. Results in
panels E. and F. on reproducibility score of observed counts.}
\end{figure}

\end{document}

### 2017-06-07 main.tex
Fixing Figures 1 and 6 legends

@@
 contacts per bin in T47D. B. Total number of contacts per bin on
-chromosome 17 of T47D. Raw (brown) and corrected (blue) signals are shown.
-The long arm of chromosome 17 (the region corresponding to 20-40 Mbp) is
-present in four coipes, explaining that the signal is about twice higher
+chromosome 17 of T47D. The brown line represents the raw
+signal, the blue line represents the signal after bias correction, the
+black line represents the signal after bias and copy number correction.
+The long arm of chromosome 17 (the region corresponding to 20-80 Mbp) is
+present in four copies, explaining that the signal is about twice higher
 than for the short arm.}
@@
 total amount of contacts after copy numer correction. The plot shows the
-same region as Figure\ref{fig:totals}B. The brown line represents the raw
+same region as panel A. The brown line represents the raw
 signal, the blue line represents the signal after bias correction, the
@@
 Competitiveness, ‘Centro de Excelencia Severo Ochoa 2013-2017’,
-SEV-2012-0208, ERC Synergy Grant 609989 and...
+SEV-2012-0208 and ERC Synergy Grant 609989. R.S. was supported by an EMBO Long-term
+Fellowship (ALTF 1201-2014) and a Marie Curie Individual Fellowship (H2020-MSCA-IF-2014).
 

### 2017-06-07 main.tex
Verb missing in line 98
Most changes
Comments

- lines 104 and 107: what do you mean by explicitily and implicitily, respectively?
- line 192: what do you mean by "the content"?
- lines 358-360: there seems to be mixed text pieces, and it converts wrong to the PDF
- F. Le Dily told me that that the ERC/Synergy has to be acknowledged as follows: "We received funding from the European Research Council under the European Union's Seventh Framework Programme (FP7/2007-2013)/ERC Synergy grant agreement 609989 (4DGenome). The content of this manuscript reflects only the author’s views and the Union i

@@
 \citep{yaffe2011probabilistic}. In addition, general experimental biases
-such as batch effects. It is thus important to normalize Hi-C data in
+such as batch effects exist. It is thus important to normalize Hi-C data in
 order to remove biases and artifacts, so that they are not confused with
@@
 
-Neither of these strategies were designed with regard for cell types with
+Neither of these strategies were designed for cell types with
 karyotypic aberrations, most common in cancer cells. Hi-C is very
@@
 bins. The values are stored in the cells of the contact matrix ($x_{ij}$),
-quantifying of the interaction between the tow loci at positions $i$ and
+quantifying of the interaction between the two loci at positions $i$ and
 $j$.
@@
 
-All data were processed though a pipeline based on TADBit
+All data were processed through a pipeline based on TADbit
 \citep{serra2016structural}. Briefly, after controlling the quality of
-fastq files, paired-end reads were mapped in the corresponding reference
+FASTQ files, paired-end reads were mapped to the corresponding reference
 genome (hg38 and mm10) taking into account the restriction enzyme site.
-Non-informative contacts were removed applying the following TADBit
+Non-informative contacts were removed applying the following TADbit
 filters: self-circle, dangling-ends, error, extra dangling-end, duplicated
@@
 {img/copy_number_figure2.pdf}}
-\caption{Pearson correlation between the total number of contacts per bin
+\caption{Pearson correlation between the corrected number of contacts per bin
 and an independent estimation of the copy number (COSMIC). Left panel
@@
 of contacts presented the worst performance for all scenarios, and it thus
-seems to be a poor choice as a metric to compare HiC matrices.
+seems to be a poor choice as a metric to compare Hi-C matrices.
 
@@
 1 min in the conditions of the benchmark. We occasionally encountered
-problems where \textit{ICE} ran faster than \textit{OneD} so it is fair to
+samples where \textit{ICE} ran faster than \textit{OneD} so it is fair to
 say that none of them has a definite speed advantage. Throughout this
@@
 a biological signal or an artifact. If the biological sample contains
-karytoypic aberrations, then its genome is grossly different the reference
+karytoypic aberrations, then its genome is grossly different from the reference
 so there can be no correct way to normalize the Hi-C signal. The proper

### 2017-06-08 main.tex
Figures improved

--- 2017-06-07 main.tex
@@
 signal, the blue line represents the signal after bias correction, the
-black line represents the signal after bias and copy number correction.
+black line represents the signal after bias and copy number (CN) correction.
 The long arm of chromosome 17 (the region corresponding to 20-80 Mbp) is
@@
 signal, the blue line represents the signal after bias correction, the
-black line represents the signal after bias and copy number correction.}
+black line represents the signal after bias and copy number (CN) correction.}
 \label{fig:cnv_correction}

### 2017-06-08 main.tex
Included Marc's comments

@@
 Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Marc A.
-Mart\'i Renom\,$^{\text{\sfb 1,2}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
+Mart\'i Renom\,$^{\text{\sfb 1,2,3,4}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
 and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}
@@
 Centre for Genomic Regulation (CRG), The Barcelona Institute of Science
-and Technology (BIST), Barcelona, Spain \\
-$^{\text{\sf 2}}$Universitat Pompeu Fabra (UPF), Barcelona, Spain}
+and Technology (BIST), Dr. Aiguader 88, 08003, Barcelona, Spain \\
+$^{\text{\sf 2}}$Universitat Pompeu Fabra (UPF), Barcelona, Spain \\
+$^{\text{\sf 3}}$CNAG-CRG, Centre for Genomic Regulation (CRG), Barcelona
+Institute of Science and Technology (BIST), Baldiri i Reixac 4, 08028 Barcelona,
+Spain \\
+$^{\text{\sf 4}}$ICREA, Pg. Lluís Companys 23, 08010 Barcelona, Spain}
 
@@
 has become part of the standard molecular biology toolbox
-\citep{denker2016second}.
+\citep{dekker2013exploring}.
 
-Contrary to the precursor technologies 3C, 4C and 5C \citep{de2012decade},
+Contrary to the precursor technologies 3C, 4C and 5C \citep{dekker2002capturing,simonis2006nuclear,dostie2006chromosome,de2012decade},
 Hi-C interrogates all possible pairwise interactions between restriction
@@
 
-Once the parameters of the model are estimated, we rescale the estimated
-means $\{\lambda_i\}$ to obtain a correction vector $\{\lambda_i'\}$ that
+Once the parameters of the model are estimated, estimated
+means $\{\lambda_i\}$ are rescaled to obtain a correction vector $\{\lambda_i'\}$ that
 can be used to compute the corrected counts ($\hat{x}_{ij}$).
@@
 The model is fitted with the Baum-Welch algorithm \citep{baum1966} until
-convergence, following the detail given by \cite{filion2010systematic}.
+convergence, following a previously described protocol \citep{filion2010systematic}.
 The Viterbi path is then computed and corresponds to the inferred copy
@@
 stadhouders2017transcription, lin2012global, dixon2012topological} and
-unpublished Hi-C data of different cell types and organisms. The details
-can be found in Table \ref{tab:samples}.
+unpublished Hi-C data of different cell types and organisms
+(Table \ref{tab:samples}).
 
@@
 We also used array-based copy-number segmentation of the two cell
-lines obtained from COSMIC database \citep{forbes2010cosmic}.
+lines obtained from the COSMIC database \citep{forbes2010cosmic}.
 
@@
 \textit{caICB} corrections respectively. All the results are based on a
-resolution of 100 kbp, but we found no major differences for different
-values (not shown).
+resolution of 100 kbp, but we found no major differences at different
+resolutions (not shown).
 
@@
 in one to four copies. In all of them, the relationship flattens as the
-number of restriction sites increases. In order to capture this
+number of restriction sites increases. To capture this
 relationship, \textit{OneD} fits a non-linear model between the total
@@
 
-We tested the validity of this approach against the Catalogue Of Somatic
-Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. In what follows,
-we benchmarked \textit{OneD}, against \textit{vanilla}, \textit{ICE}
-\citep{imakaev2012iterative}, \textit{caICB} \citep{wu2016computational}
-and the Local Genomic Features method
-\citep[\textit{LGF},][]{hu2012hicnorm, servant2012hitc}. The first three
-methods correct biases implicitly, whereas the fourth method does it
-explicitly. Figure \ref{fig:copy_number} shows the Pearson correlation
+% Marc's comment here was regarding lack of motivation and explanation for the
+% copy number comparison. I've rescued some text from an older version. We
+% should also comment here about CN being a bias or providing biological
+% information, so to correct or not to correct for CN
+
+In what follows, we benchmarked \textit{OneD},
+against \textit{vanilla}, \textit{ICE} \citep{imakaev2012iterative}, \textit{caICB} \citep{wu2016computational}
+and the Local Genomic Features method \citep[\textit{LGF},][]{hu2012hicnorm,
+servant2012hitc}. The first three methods correct biases implicitly, whereas the
+fourth method does it explicitly.
+
+Given that our model is based on total number of contacts, we reasoned that a preliminary test
+would be to check if the corrected number of contacts per bin reflects the copy
+number (as measured by an independent technique such as array-based copy-number
+segmentation). We tested the validity of this approach against the Catalogue Of Somatic
+Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure \ref{fig:copy_number} shows the Pearson correlation
 between the corrected number of contacts and the copy number estimation
@@
 {img/copy_number_figure2.pdf}}
-\caption{Pearson correlation between the corrected number of contacts per bin
+\caption{Pearson correlation between the total number of contacts per bin
 and an independent estimation of the copy number (COSMIC). Left panel
@@
 decrease the similarity between samples that received a different
-treatment in order to enhance the biological differences.
+treatment to enhance the biological differences.
 
@@
 cell lines with aberrant karyotypes. In each set we spiked two samples
-from the other cell line (see Table~\ref{tab:samples}) in order to
+from the other cell line (Table~\ref{tab:samples}) to
 introduce biological variability. We compared matrices before and after
@@
 classifier scores and summarized the results with a ROC curve
-(Figures~\ref{fig:aberrant}C and D).
-
-All the methods, including the absence of normalization, succeeded in
-identifying the T47D outliers among the K562 samples, but recognizing the
-K562 outliers among the T47D samples proved more challenging.
-\textit{OneD} increased the discrimination power compared to the raw
-matrices, but all the other methods decreased it. Actually, they performed
-little better than random on this task.
-
-Using the other metrics described in \ref{sec:comp} yielded similar
-results (Supplementary Figures 5, 6 and 7). Note that Spearman correlation
-of contacts presented the worst performance for all scenarios, and it thus
-seems to be a poor choice as a metric to compare Hi-C matrices.
+(Figures~\ref{fig:aberrant}C and D). All the methods, including the absence of
+normalization, succeeded in identifying the T47D outliers among the K562
+samples, but recognizing the K562 outliers among the T47D samples proved more
+challenging. \textit{OneD} increased the discrimination power compared to the
+raw matrices, but all the other methods decreased it. Actually, they performed
+little better than random on this task. Using the other metrics described
+in \ref{sec:comp} yielded similar results (Supplementary Figures 5, 6 and
+7). Note that Spearman correlation of contacts presented the worst performance
+for all scenarios, and it thus seems to be a poor choice as a metric to compare
+Hi-C matrices.
 
@@
 the curve on this problem, but with a small margin over the other methods
-except \textit{vanilla}.
-
-Using other metrics to compare matrices gave similar results
-(Supplementary Figures 7 and 8). \textit{OneD} was always among the top
-scoring methods. In these conditions, Spearman correlation of contacts
-again appeared as the worst comparison metric because it showed a lower
-performance for all the normalization methods.
+except \textit{vanilla}. Using other metrics to compare matrices gave similar
+results (Supplementary Figures 7 and 8). \textit{OneD} was always among the top
+scoring methods. In these conditions, Spearman correlation of contacts again
+appeared as the worst comparison metric because it showed a lower performance
+for all the normalization methods.
 
@@
 of bias to be remove. For this reason, \textit{OneD} also allows the user
-to correct the copy number and output a euploid-equivalent matrix.
+to correct the copy number and output a euploid-equivalent matrix. To do
+so, \textit{OneD} segments the linear signal of the total amount of contacts
+into piecewise homoegenous regions. This is carried out by a hidden Markov model
+whereby the averages of the states are constrained to be within a discrete
+factor of each other. This allows the model to detect regions with a number of
+copies equal to 1, 2, 3 \textit{etc}. With the inferred number of copies at
+hand, \textit{OneD} then normalizes each cell of the matrix with equation
+(\ref{eq:cnvnorm}).
 
-Here, \textit{OneD} segments the linear signal of the total amount of
-contacts into piecewise homoegenous regions. This is carried out by a
-hidden Markov model whereby the averages of the states are constrained to
-be within a discrete factor of each other. This allows the model to detect
-regions with a number of copies equal to 1, 2, 3 \textit{etc}. With the
-inferred number of copies at hand, \textit{OneD} then normalizes each cell
-of the matrix with equation (\ref{eq:cnvnorm}).
-
-Figure~\ref{fig:cnv_correction} shows the effect of this process on the
-final Hi-C matrix. Panel \ref{fig:cnv_correction}A indicates that
-different normalization methods can either enhance or diminish the signal
-in regions with higher copy number. In this example, \textit{ICE}
-overcompensates the original bias at the center of the picture and fades
-the signal almost entirely. Concomitantly, the signal at the bottom left
-of the matrix is enhanced and shows a structure that is not visible in the
-raw data. On the contrary, \textit{LGF} strengthen the central region and
-the diagonal. \textit{OneD} reduces the level of the central portion by a
-factor 2 approximately, but does not otherwise distort the main features
-of the region. This example shows that the copy number does not have a
-simple and predictable effect on the final matrix. Not taking it into
-account may open the door to some normalization artifacts. The result
-provided by \textit{OneD} is not necessarily the right one (see
-Discussion), but at least it does not correct copy number variations
-as a side effect of some other criteria.
+Different normalization methods can either enhance or diminish the signal
+in regions with higher copy number (Figure~\ref{fig:cnv_correction}). In this
+example, \textit{ICE} overcompensates the original bias at the center of the
+picture and fades the signal almost entirely. Concomitantly, the signal at the
+bottom left of the matrix is enhanced and shows a structure that is not visible
+in the raw data. On the contrary, \textit{LGF} strengthen the central region and
+the diagonal. \textit{OneD} reduces the level of the central portion by a factor
+2 approximately, but does not otherwise distort the main features of the
+region. This example shows that the copy number does not have a simple and
+predictable effect on the final matrix. Not taking it into account may open the
+door to some normalization artifacts. The result provided by \textit{OneD} is
+not necessarily the right one (see Discussion), but at least it does not correct
+copy number variations as a side effect of some other criteria.
 
@@
 
-In summary, \textit{OneD} is the only method that can be used to obtain a
-euploid-equivalent normalized matrix in cases that the effect of the copy
-number must be removed from the signal.
+In summary, \textit{OneD} can be used to obtain a euploid-equivalent normalized
+matrix in cases that the effect of the copy number must be removed from the signal.
 
@@
 Finally, we compared the computational speed of the different
-normalization methods. One of the main reasons for the broad acceptance of
-\textit{vanilla} and \textit{ICE} as standards is the speed of the
-available implementation \citep{imakaev2012iterative}. This is even more
-significant as current explicit methods \citep{servant2012hitc} are much
-slower in comparison.
+normalization methods. \textit{vanilla} and \textit{ICE} have broad acceptance
+for their conceptual simplicity, easy use and
+speed \citep{imakaev2012iterative}. This is even more significant as current
+explicit methods \citep{servant2012hitc} are much slower in comparison.
 
@@
 greater running speed, while preserving a good performance on samples with
-karyotypic aberrations. One of the reasons that \textit{OneD} is better
-able to highlight the biological distinctions between samples is that it
-only corrects the copy number if specifically requested by the user. The
-impact of copy number variations on normalization is rather opaque,
-especially if they are treated as implicit biases
-(Figure~\ref{fig:cnv_correction}). Treating them as explicit biases with
-optional removal seems to be an overall safer strategy.
+karyotypic aberrations. One of the reasons why \textit{OneD} is able to better
+highlight the biological distinctions between samples is that it only corrects
+the copy number if specifically requested by the user. The impact of copy number
+variations on normalization is rather opaque, especially if they are treated as
+implicit biases (Figure~\ref{fig:cnv_correction}). Treating them as explicit
+biases with optional removal seems to be an overall safer strategy.
 
@@
 a biological signal or an artifact. If the biological sample contains
-karytoypic aberrations, then its genome is grossly different from the reference
-so there can be no correct way to normalize the Hi-C signal. The proper
-approach would be to use the actual genome of the biological sample as a
-reference to construct the contact map. However, this strategy is
-presently unfeasible because assembling mammalian genomes is still a hard
-problem.
+karytoypic aberrations, then its genome is grossly different then the reference
+genome, which makes signal correction very challenging. The proper approach
+would be to use the actual genome of the biological sample as a reference to
+construct the contact map. However, this strategy is presently unfeasible
+because assembling mammalian genomes is still a hard problem.
 

### 2017-06-08 main.tex
Added funding

@@
 
-This work was supported by the Spanish Ministry of Economy and
-Competitiveness, ‘Centro de Excelencia Severo Ochoa 2013-2017’,
-SEV-2012-0208 and ERC Synergy Grant 609989. R.S. was supported by an EMBO Long-term
-Fellowship (ALTF 1201-2014) and a Marie Curie Individual Fellowship (H2020-MSCA-IF-2014).
+This work was partially supported by the Spanish Ministry of Economy and
+Competitiveness 'Centro de Excelencia Severo Ochoa 2013-2017' (SEV-2012-0208)
+and ACER to CRG. R.S. was supported by an EMBO Long-term
+Fellowship (ALTF 1201-2014) and a Marie Curie Individual Fellowship
+(H2020-MSCA-IF-2014). We received funding from the European Research Council under
+the European Union's Seventh Framework Programme (FP7/2007-2013)/ERC Synergy
+grant agreement 609989 (4DGenome). The content of this manuscript reflects only
+the author’s views and the Union is not liable for any use that may be made of
+the information contained therein.
 

### 2017-06-08 main.tex
Minor changes

@@
 sequencing reads have been shown to impact the results
-\citep{yaffe2011probabilistic}. In addition, general experimental biases
-such as batch effects exist. It is thus important to normalize Hi-C data in
+\citep{yaffe2011probabilistic}, in addition to general experimental biases
+such as batch effects. It is thus important to normalize Hi-C data in
 order to remove biases and artifacts, so that they are not confused with
@@
 fixed effect was the normalization method and the random effect was the
-%chromosome. Receiver operating characteristic (ROC) curves were then
-computed using \texttt{ROCR} package \citep{sing2005rocr}.
+chromosome. Receiver operating characteristic (ROC) curves were computed
+using \texttt{ROCR} package \citep{sing2005rocr}.
 
@@
 We would like to thank the members of the 4DGenome Synergy project for the
-fruitful discussions during project meetings. EV wants to acknowledge
+fruitful discussions during project meetings. E.V. wants to acknowledge
 the members of Miguel Beato's laboratory for their insights during lab

### 2017-06-09 main.tex
Reviewed and addressed Quique's three points. Added contributions.

@@
 genomes have an underlying three-dimensional structure contributing to
-their activity \citep{rowley2016three}. In mammals, this organization
-plays a key role in guiding enhancer-promoter contacts
-\citep{de2013topology}, in V(D)J recombination \citep{choi2014ctcf} and in
-X chromosome inactivation \citep{galupa2015x}.
-
+their activity \citep{rowley2016three, dekker20163d, pezic2017more}. In
+mammals, this organization plays a key role in guiding enhancer-promoter
+contacts \citep{de2013topology}, in V(D)J recombination
+\citep{choi2014ctcf} and in X chromosome inactivation \citep{galupa2015x}.
 One of the most significant breakthroughs towards this insight was the
@@
 
-Contrary to the precursor technologies 3C, 4C and 5C \citep{dekker2002capturing,simonis2006nuclear,dostie2006chromosome,de2012decade},
-Hi-C interrogates all possible pairwise interactions between restriction
-fragments. However, this does not guarantee that the method has no bias.
-On the contrary, local genome features such as the G+C content, the
-availability of restriction enzyme sites and the mappability of the
-sequencing reads have been shown to impact the results
-\citep{yaffe2011probabilistic}, in addition to general experimental biases
-such as batch effects. It is thus important to normalize Hi-C data in
-order to remove biases and artifacts, so that they are not confused with
-biological signal.
+Contrary to the precursor technologies 3C, 4C and 5C
+\citep{dekker2002capturing, simonis2006nuclear,
+dostie2006chromosome,de2012decade}, Hi-C interrogates all possible
+pairwise interactions between restriction fragments. However, this does
+not guarantee that the method has no bias.  On the contrary, local genome
+features such as the G+C content, the availability of restriction enzyme
+sites and the mappability of the sequencing reads have been shown to
+impact the results \citep{yaffe2011probabilistic}, in addition to general
+experimental biases such as batch effects. It is thus important to
+normalize Hi-C data in order to remove biases and artifacts, so that they
+are not confused with biological signal.
 
@@
 
-Neither of these strategies were designed for cell types with
-karyotypic aberrations, most common in cancer cells. Hi-C is very
-sensitive to aneuploidy, copy number variations and translocations.
-Actually, these aberrations have so much influence on the outcome that the
-artifacts can be used to re-assemble the target genome
-\citep{korbel2013genome}. So far the only attempt to address the issue
-was the chromosome-adjusted Iterative Correction Bias method
-(\textit{caICB}) of \cite{wu2016computational}.  However, caICB applies a
-uniform chromosome-wide correction, effectively excluding the numerous
+Neither of these strategies were designed for cell types with karyotypic
+aberrations, most common in cancer. Yet, Hi-C is very sensitive to
+aneuploidy, copy number variations and translocations.  Actually, these
+aberrations have so much influence on the outcome that the artifacts can
+be used to re-assemble the target genome \citep{korbel2013genome}. An
+additional complication is that karyotypic aberrations are not
+experimental biases, so it is unclear whether they should be corrected at
+all or be considered part of the biological signal.
+
+So far, the only attempt to address the issue was the chromosome-adjusted
+Iterative Correction Bias method (\textit{caICB}) of
+\cite{wu2016computational}. However, \textit{caICB} applies a uniform
+chromosome-wide copy number correction, effectively excluding the numerous
 cases of partial aneuploidy and regional copy number variations.
@@
 Our approach is to model the tally of contacts for each bin, thus reducing
-the matrix to a one dimension score (hence the name \textit{OneD}). We assume that
-the total number of contacts per bin ($t_{i}$) can be approximated by a
-negative binomial distribution. This choice is sensible because the amount
-of contacts is a discrete variable and because the negative binomial
-distribution allows for overdispersion. We further assume that the
-explicit sources of bias have independent contributions to the mean of the
-distribution for a given bin ($\lambda_i$).
+the matrix to a one dimension score (hence the name \textit{OneD}). We
+assume that the total number of contacts per bin ($t_{i}$) can be
+approximated by a negative binomial distribution. This choice is sensible
+because the amount of contacts is a discrete variable and because the
+negative binomial distribution allows for overdispersion. We further
+assume that the explicit sources of bias have independent contributions to
+the mean of the distribution for a given bin ($\lambda_i$).
 
@@
 
-Once the parameters of the model are estimated, estimated
-means $\{\lambda_i\}$ are rescaled to obtain a correction vector $\{\lambda_i'\}$ that
-can be used to compute the corrected counts ($\hat{x}_{ij}$).
+Once the parameters of the model are estimated, estimated means
+$\{\lambda_i\}$ are rescaled to obtain a correction vector
+$\{\lambda_i'\}$ that can be used to compute the corrected counts
+($\hat{x}_{ij}$).
 
@@
 The model is fitted with the Baum-Welch algorithm \citep{baum1966} until
-convergence, following a previously described protocol \citep{filion2010systematic}.
-The Viterbi path is then computed and corresponds to the inferred copy
-number of each bin ($c_i$).
+convergence, following a previously described protocol
+\citep{filion2010systematic}.  The Viterbi path is then computed and
+corresponds to the inferred copy number of each bin ($c_i$).
 
@@
 \processtable{Hi-C data sets used in this study. Sample ID: Unique sample
-identifier. Cell type: Type of cell. Application: Hi-C protocol. RE: Restriction
-enzyme used for digestion. Sequencing core: Laboratory performing the experiment
-(CRG: Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor: Baylor College of Medicine;
-UCSD: University of California San Diego; LICR: Ludwig Institue for Cancer Research). Set(s): Comparison set. Source: SRA, GEO
-or ENCODE raw data identifier.
+identifier in the respective projects. Cell type: source of the biological
+sample. Application: Hi-C protocol. RE: Restriction enzyme used for the
+digestion. Sequencing core: Laboratory performing the experiment (CRG:
+Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor:
+Baylor College of Medicine; UCSD: University of California San Diego;
+LICR: Ludwig Institue for Cancer Research). Set(s): Sets in which the
+sample was included (see text for detail).  Source: SRA, GEO or ENCODE raw
+data identifier.
 \label{tab:samples}}
@@
   MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
-b7fa2d8db\_bfac48760 & B-cell  & \textit{in situ} Hi-C &
+b7fa2d8db\_bfac48760 & B cell  & \textit{in situ} Hi-C &
   DpnII & CRG & mm10 & GSE96611 \\
-fc3e8b36a\_7bf1bf374 & ES-cell & \textit{in situ} Hi-C &
+fc3e8b36a\_7bf1bf374 & ES cell & \textit{in situ} Hi-C &
   DpnII & CRG & mm10 & GSE96611 \\
-b7fa2d8db\_7284b867a & B-cell  & \textit{in situ} Hi-C &
+b7fa2d8db\_7284b867a & B cell  & \textit{in situ} Hi-C &
   DpnII & CRG & mm10 & GSE96611 \\
-fc3e8b36a\_38bfd1b33 & ES-cell & \textit{in situ} Hi-C &
+fc3e8b36a\_38bfd1b33 & ES cell & \textit{in situ} Hi-C &
   DpnII & CRG & mm10 & GSE96611 \\
-b7fa2d8db\_58e812fc2 & B-cell  & \textit{in situ} Hi-C &
+b7fa2d8db\_58e812fc2 & B cell  & \textit{in situ} Hi-C &
   DpnII & CRG & mm10 & GSE96611 \\
-fc3e8b36a\_c990a254e & ES-cell & \textit{in situ} Hi-C &
+fc3e8b36a\_c990a254e & ES cell & \textit{in situ} Hi-C &
   DpnII & CRG & mm10 & GSE96611 \\
-b7fa2d8db\_73f11d923 & B-cell  & \textit{in situ} Hi-C &
+b7fa2d8db\_73f11d923 & B cell  & \textit{in situ} Hi-C &
   DpnII & CRG & mm10 & GSE96611 \\
-fc3e8b36a\_4bf044f18 & ES-cell & \textit{in situ} Hi-C &
+fc3e8b36a\_4bf044f18 & ES cell & \textit{in situ} Hi-C &
   DpnII & CRG & mm10 & GSE96611 \\
-GSM987817 & B-cell  & dilution Hi-C &
+GSM987817 & B cell  & dilution Hi-C &
   HindIII & UCSD & mm10 & SRR543428-SRR543431 \\
-GSM987818 & B-cell  & dilution Hi-C &
+GSM987818 & B cell  & dilution Hi-C &
   HindIII & UCSD & mm10 & SRR543432-SRR543442 \\
-GSM862720 & ES-cell & dilution Hi-C &
+GSM862720 & ES cell & dilution Hi-C &
   HindIII & LICR & mm10 & SRR443883-SRR443885 \\
-GSM862721 & ES-cell & dilution Hi-C &
+GSM862721 & ES cell & dilution Hi-C &
   HindIII & LICR & mm10 & SRR400251-SRR400256 \\
-GSM862722 & ES-cell & dilution Hi-C &
+GSM862722 & ES cell & dilution Hi-C &
   NcoI    & LICR & mm10 & SRR443886-SRR443888 \\
@@
 
-% Marc's comment here was regarding lack of motivation and explanation for the
-% copy number comparison. I've rescued some text from an older version. We
-% should also comment here about CN being a bias or providing biological
-% information, so to correct or not to correct for CN
+% Marc's comment here was regarding lack of motivation and explanation for
+% the copy number comparison. I've rescued some text from an older
+% version. We should also comment here about CN being a bias or providing
+% biological information, so to correct or not to correct for CN
 
-In what follows, we benchmarked \textit{OneD},
-against \textit{vanilla}, \textit{ICE} \citep{imakaev2012iterative}, \textit{caICB} \citep{wu2016computational}
-and the Local Genomic Features method \citep[\textit{LGF},][]{hu2012hicnorm,
-servant2012hitc}. The first three methods correct biases implicitly, whereas the
-fourth method does it explicitly.
+In what follows, we benchmarked \textit{OneD}, against \textit{vanilla},
+\textit{ICE} \citep{imakaev2012iterative}, \textit{caICB}
+\citep{wu2016computational} and the Local Genomic Features method
+\citep[\textit{LGF},][]{hu2012hicnorm, servant2012hitc}. The first three
+methods correct biases implicitly, whereas the fourth method does it
+explicitly.
 
-Given that our model is based on total number of contacts, we reasoned that a preliminary test
-would be to check if the corrected number of contacts per bin reflects the copy
-number (as measured by an independent technique such as array-based copy-number
-segmentation). We tested the validity of this approach against the Catalogue Of Somatic
-Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure \ref{fig:copy_number} shows the Pearson correlation
-between the corrected number of contacts and the copy number estimation
-for both T47D and K562 cell lines. Similar results were obtained using
-Spearman correlation (Supplementary Figure 1). All the methods except
-\textit{OneD} decrease the agreement between the signal and the copy
-number because they partially correct the discrepancy. In contrast,
-\textit{OneD} enhances the conformity of the signal with the copy number.
-Not correcting for variable copy number at that stage may seem
-counter-intuitive, but the tests below will show this leads to better
+Given that our model is based on total number of contacts, we reasoned
+that a preliminary test would be to check if the corrected number of
+contacts per bin reflects the copy number (as measured by an independent
+technique such as array-based copy-number segmentation). We tested the
+validity of this approach against the Catalogue Of Somatic Mutations In
+Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure \ref{fig:copy_number}
+shows the Pearson correlation between the corrected number of contacts and
+the copy number estimation for both T47D and K562 cell lines. Similar
+results were obtained using Spearman correlation (Supplementary Figure 1).
+All the methods except \textit{OneD} decrease the agreement between the
+signal and the copy number because they partially correct the discrepancy.
+In contrast, \textit{OneD} enhances the conformity of the signal with the
+copy number.  Not correcting for variable copy number at that stage may
+seem counter-intuitive, but the tests below will show this leads to better
 performance.
@@
 between the different cell lines (Figures~\ref{fig:aberrant}A and
-\ref{fig:aberrant}B and Supplementary Figures~2 and 3). This is an undesirable
-effect, as it obscures the biological variability. In the same vein, these
-methods decreased the similarity between samples that received the same
-treatment (Figure~\ref{fig:aberrant}A), suggesting that the normalization
-process is detrimental to the biological signal in these two cases. The
-method \textit{vanilla} followed the same trend but to a lesser extent,
-consistent with the fact that it consists of a single \textit{ICE}
+\ref{fig:aberrant}B and Supplementary Figures~2 and 3). This is an
+undesirable effect, as it obscures the biological variability. In the same
+vein, these methods decreased the similarity between samples that received
+the same treatment (Figure~\ref{fig:aberrant}A), suggesting that the
+normalization process is detrimental to the biological signal in these two
+cases. The method \textit{vanilla} followed the same trend but to a lesser
+extent, consistent with the fact that it consists of a single \textit{ICE}
 iteration. \textit{OneD} was the only method to increase the similarity
@@
 classifier scores and summarized the results with a ROC curve
-(Figures~\ref{fig:aberrant}C and D). All the methods, including the absence of
-normalization, succeeded in identifying the T47D outliers among the K562
-samples, but recognizing the K562 outliers among the T47D samples proved more
-challenging. \textit{OneD} increased the discrimination power compared to the
-raw matrices, but all the other methods decreased it. Actually, they performed
-little better than random on this task. Using the other metrics described
-in \ref{sec:comp} yielded similar results (Supplementary Figures 5, 6 and
-7). Note that Spearman correlation of contacts presented the worst performance
-for all scenarios, and it thus seems to be a poor choice as a metric to compare
-Hi-C matrices.
+(Figures~\ref{fig:aberrant}C and D). All the methods, including the
+absence of normalization, succeeded in identifying the T47D outliers among
+the K562 samples, but recognizing the K562 outliers among the T47D samples
+proved more challenging. \textit{OneD} increased the discrimination power
+compared to the raw matrices, but all the other methods decreased it.
+Actually, they performed little better than random on this task. Using the
+other metrics described in \ref{sec:comp} yielded similar results
+(Supplementary Figures 5, 6 and 7). Note that Spearman correlation of
+contacts presented the worst performance for all scenarios, and it thus
+seems to be a poor choice as a metric to compare Hi-C matrices.
 
@@
 the curve on this problem, but with a small margin over the other methods
-except \textit{vanilla}. Using other metrics to compare matrices gave similar
-results (Supplementary Figures 7 and 8). \textit{OneD} was always among the top
-scoring methods. In these conditions, Spearman correlation of contacts again
-appeared as the worst comparison metric because it showed a lower performance
-for all the normalization methods.
+except \textit{vanilla}. Using other metrics to compare matrices gave
+similar results (Supplementary Figures 7 and 8). \textit{OneD} was always
+among the top scoring methods. In these conditions, Spearman correlation
+of contacts again appeared as the worst comparison metric because it
+showed a lower performance for all the normalization methods.
 
@@
 to correct the copy number and output a euploid-equivalent matrix. To do
-so, \textit{OneD} segments the linear signal of the total amount of contacts
-into piecewise homoegenous regions. This is carried out by a hidden Markov model
-whereby the averages of the states are constrained to be within a discrete
-factor of each other. This allows the model to detect regions with a number of
-copies equal to 1, 2, 3 \textit{etc}. With the inferred number of copies at
-hand, \textit{OneD} then normalizes each cell of the matrix with equation
-(\ref{eq:cnvnorm}).
+so, \textit{OneD} segments the linear signal of the total amount of
+contacts into piecewise homoegenous regions. This is carried out by a
+hidden Markov model whereby the averages of the states are constrained to
+be within a discrete factor of each other. This allows the model to detect
+regions with a number of copies equal to 1, 2, 3 \textit{etc}. With the
+inferred number of copies at hand, \textit{OneD} then normalizes each cell
+of the matrix with equation (\ref{eq:cnvnorm}).
 
 Different normalization methods can either enhance or diminish the signal
-in regions with higher copy number (Figure~\ref{fig:cnv_correction}). In this
-example, \textit{ICE} overcompensates the original bias at the center of the
-picture and fades the signal almost entirely. Concomitantly, the signal at the
-bottom left of the matrix is enhanced and shows a structure that is not visible
-in the raw data. On the contrary, \textit{LGF} strengthen the central region and
-the diagonal. \textit{OneD} reduces the level of the central portion by a factor
-2 approximately, but does not otherwise distort the main features of the
-region. This example shows that the copy number does not have a simple and
-predictable effect on the final matrix. Not taking it into account may open the
-door to some normalization artifacts. The result provided by \textit{OneD} is
-not necessarily the right one (see Discussion), but at least it does not correct
-copy number variations as a side effect of some other criteria.
+in regions with higher copy number (Figure~\ref{fig:cnv_correction}). In
+this example, \textit{ICE} overcompensates the original bias at the center
+of the picture and fades the signal almost entirely. Concomitantly, the
+signal at the bottom left of the matrix is enhanced and shows a structure
+that is not visible in the raw data. On the contrary, \textit{LGF}
+strengthen the central region and the diagonal. \textit{OneD} reduces the
+level of the central portion by a factor 2 approximately, but does not
+otherwise distort the main features of the region. This example shows that
+the copy number does not have a simple and predictable effect on the final
+matrix. Not taking it into account may open the door to some normalization
+artifacts. The result provided by \textit{OneD} is not necessarily the
+right one (see Discussion), but at least it does not correct copy number
+variations as a side effect of some other criteria.
 
@@
 total amount of contacts after copy numer correction. The plot shows the
-same region as panel A. The brown line represents the raw
-signal, the blue line represents the signal after bias correction, the
-black line represents the signal after bias and copy number (CN) correction.}
+same region as panel A. The brown line represents the raw signal, the blue
+line represents the signal after bias correction, the black line
+represents the signal after bias and copy number (CN) correction.}
 \label{fig:cnv_correction}
@@
 Figure~\ref{fig:cnv_correction}B shows the total sum of contacts after
-correction for the copy number with \textit{OneD}. The signal that
-originally showed a sharp transition corresponding to the gain or loss of
-a large region, now shows a flat trend throughout the chromosome.
+correction for the copy number with \textit{OneD}. The region around
+72.5-75.0 Mbp showed an elevated amount of contacts in the raw data. After
+copy number correction, the signal is brought to the same level as the
+flanking regions.
 
-In summary, \textit{OneD} can be used to obtain a euploid-equivalent normalized
-matrix in cases that the effect of the copy number must be removed from the signal.
+In summary, \textit{OneD} can be used to obtain a euploid-equivalent
+normalized matrix in cases that the effect of the copy number must be
+removed from the signal.
 
@@
 Finally, we compared the computational speed of the different
-normalization methods. \textit{vanilla} and \textit{ICE} have broad acceptance
-for their conceptual simplicity, easy use and
-speed \citep{imakaev2012iterative}. This is even more significant as current
+normalization methods. \textit{vanilla} and \textit{ICE} have broad
+acceptance for their conceptual simplicity, easy use and speed
+\citep{imakaev2012iterative}. This is even more significant as current
 explicit methods \citep{servant2012hitc} are much slower in comparison.
@@
 
-At increasing resolution, the relative rank of the methods remained the
-same (Figure~\ref{fig:times}B). The running time of \textit{OneD} remained
-close to that of \textit{ICE}. Taken together, these results show that
-\textit{OneD} is competitive in terms of computational speed compared to
-existing methods.
+We also performed a benchmark on a smaller data set at increasing
+resolution (Figure~\ref{fig:times}B). On this benchmark, \textit{ICE} ran
+slightly faster than \textit{OneD}, and the the rank of the methods
+remained the same at all resolutions. The running time of \textit{OneD}
+remained close to that of \textit{ICE}. Taken together, these results show
+that \textit{OneD} is competitive in terms of computational speed compared
+to existing methods.
 
@@
 greater running speed, while preserving a good performance on samples with
-karyotypic aberrations. One of the reasons why \textit{OneD} is able to better
-highlight the biological distinctions between samples is that it only corrects
-the copy number if specifically requested by the user. The impact of copy number
-variations on normalization is rather opaque, especially if they are treated as
-implicit biases (Figure~\ref{fig:cnv_correction}). Treating them as explicit
-biases with optional removal seems to be an overall safer strategy.
+karyotypic aberrations. One of the reasons why \textit{OneD} is able to
+better highlight the biological distinctions between samples is that it
+only corrects the copy number if specifically requested by the user. The
+impact of copy number variations on normalization is rather opaque,
+especially if they are treated as implicit biases
+(Figure~\ref{fig:cnv_correction}). Treating them as explicit biases with
+optional removal seems to be an overall safer strategy.
 
@@
 a biological signal or an artifact. If the biological sample contains
-karytoypic aberrations, then its genome is grossly different then the reference
-genome, which makes signal correction very challenging. The proper approach
-would be to use the actual genome of the biological sample as a reference to
-construct the contact map. However, this strategy is presently unfeasible
-because assembling mammalian genomes is still a hard problem.
+karytoypic aberrations, then its genome is grossly different then the
+reference genome, which makes signal correction very challenging. The
+proper approach would be to use the actual genome of the biological sample
+as a reference to construct the contact map. However, this strategy is
+presently unfeasible because assembling mammalian genomes is still a hard
+problem.
 
@@
 
+\section*{Author contributions}
+
+Conceptualization, E.V.; Methodology, E.V. and G.F.; Software, E.V. and
+G.F.; Formal Analysis, E.V.; Investigation, E.V., F.D., Y.C. and R.S.;
+Resources, M.B.; Data curation, J.Q., Writing – Original Draft, E.V. and
+G.F., Writing – Review \& Editing, F.D. J.Q., R.S., Y.C, T.G., M.M. and
+M.B.; Visualization, E.V.; Supervision, G.F. and M.B., Project
+Administration, M.B.; Funding Acquisition, M.B., T.G., M.M. and G.F.
+
 \section*{Funding}
@@
 This work was partially supported by the Spanish Ministry of Economy and
-Competitiveness 'Centro de Excelencia Severo Ochoa 2013-2017' (SEV-2012-0208)
-and ACER to CRG. R.S. was supported by an EMBO Long-term
+Competitiveness `Centro de Excelencia Severo Ochoa 2013-2017'
+(SEV-2012-0208) and ACER to CRG. R.S. was supported by an EMBO Long-term
 Fellowship (ALTF 1201-2014) and a Marie Curie Individual Fellowship
-(H2020-MSCA-IF-2014). We received funding from the European Research Council under
-the European Union's Seventh Framework Programme (FP7/2007-2013)/ERC Synergy
-grant agreement 609989 (4DGenome). The content of this manuscript reflects only
-the author’s views and the Union is not liable for any use that may be made of
-the information contained therein.
+(H2020-MSCA-IF-2014). We received funding from the European Research
+Council under the European Union's Seventh Framework Programme
+(FP7/2007-2013)/ERC Synergy grant agreement 609989 (4DGenome). The content
+of this manuscript reflects only the author's views and the Union is not
+liable for any use that may be made of the information contained therein.
 

### 2017-06-09 main.tex
Text version 1.0. Ready for first submission.

@@
 Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Marc A.
-Mart\'i Renom\,$^{\text{\sfb 1,2,3,4}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
+Mart\'i-Renom\,$^{\text{\sfb 1,2,3,4}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
 and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}
@@
 presence of large-scale copy number variations, potentially obscuring
-biological variations and enhancing batch effects. To address this issue,
+biological differences and enhancing batch effects. To address this issue,
 we developed an alternative approach designed to take into account
@@
 reproducibility among replicates of Hi-C samples with abnormal karyotype,
-thus significantly outperforming previous methods.  In the absence of
-chromosomal aberrations, \textit{OneD} fared equally well as
-state-of-the-art methods, making it a safe choice for Hi-C normalization.
-\textit{OneD} is fast and scales well in terms of computing resources for
-resolutions up to 1 kbp. \\
+outperforming previous methods significantly. On normal karyotypes,
+\textit{OneD} fared equally well as state-of-the-art methods, making it a
+safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
+in terms of computing resources for resolutions up to 1 kbp.  \\
 \textbf{Availability:} \textit{OneD} is implemented as an R package
@@
 \citep{choi2014ctcf} and in X chromosome inactivation \citep{galupa2015x}.
-One of the most significant breakthroughs towards this insight was the
-development of the high throughput chromosomal conformation capture
-technology (Hi-C), assaying chromosomal contacts at a genome-wide scale
+A significant breakthrough towards this insight was the development of
+the high throughput chromosomal conformation capture technology (Hi-C),
+assaying chromosomal contacts at a genome-wide scale
 \citep{lieberman2009comprehensive}. Nowadays, exploring the spatial
@@
 aneuploidy, copy number variations and translocations.  Actually, these
-aberrations have so much influence on the outcome that the artifacts can
-be used to re-assemble the target genome \citep{korbel2013genome}. An
+aberrations have so much influence on the outcome that they can be used as
+signatureto re-assemble the target genome \citep{korbel2013genome}. An
 additional complication is that karyotypic aberrations are not
@@
 bins. The values are stored in the cells of the contact matrix ($x_{ij}$),
-quantifying of the interaction between the two loci at positions $i$ and
+quantifying the interaction between the two loci at positions $i$ and
 $j$.
@@
 
-Once the parameters of the model are estimated, estimated means
+Once the parameters of the model are determined, the estimated means
 $\{\lambda_i\}$ are rescaled to obtain a correction vector
@@
 In line with previous methods
-\citep{yaffe2011probabilistic,hu2012hicnorm}, the features
-used to fit the model are the local G+C content, read mappability and the 
-content and number of restriction sites. The model and the implementation
-can be easily extended with any user-provided genomic feature.
+\citep{yaffe2011probabilistic,hu2012hicnorm}, the default features
+used to fit the model are the local G+C content, the read mappability and
+the content and number of restriction sites. The model and the
+implementation can be modifed or extended with any user-provided genomic
+features.
 
 \subsection{Copy number correction}
+\label{sec:hmm}
 
 Briefly, a hidden Markov model with emissions distributed as a Student's
-$t$ variable is fitted on the corrected total amount of contacts per bin.
-The model consists of 7 states that correspond to 1, 2, 3, 4, 5, 6 and 8
-copies of the target, for a total of 3 emission parameters (a single
-scaling parameter, a single standard deviation and a single degree of
-freedom for all the states) and 21 transition parameters.
+$t$ variable is fitted on the corrected total amount of contacts per bin
+($\lambda_i'$). The model consists of 7 states that correspond to 1, 2,
+3, 4, 5, 6 and 8 copies of the target, for a total of 3 emission
+parameters (a single scaling parameter, a single standard deviation and a
+single degree of freedom for all the states) and 21 transition parameters.
 
 The model is fitted with the Baum-Welch algorithm \citep{baum1966} until
-convergence, following a previously described protocol
+convergence, following a previously described implementation
 \citep{filion2010systematic}.  The Viterbi path is then computed and
@@
 \label{eq:cnvnorm}
-\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}
+\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}.
 \end{equation}
@@
 
-We used several experiments comprising T47D breast cancer cell
-lines (7 samples), K562 leukemia cell lines (4 samples), both with
-aberrant karyotypes, and mouse primary B (6 samples ) and ES (7 samples)
-cells, both with normal diploid karyotypes. The experiments were carried
-out in different laboratories, following either the original Hi-C protocol
-\citep{lieberman2009comprehensive} or the newer \textit{in situ} version
-\citep{rao20143d}, and using different restriction enzymes (DpnII,
-HindIII, MboI and NcoI).
-
-We also used array-based copy-number segmentation of the two cell
-lines obtained from the COSMIC database \citep{forbes2010cosmic}.
+We used several experiments comprising T47D breast cancer cell lines (7
+samples), K562 leukemia cell lines (4 samples), both with aberrant
+karyotypes; and mouse primary B cells (6 samples ) and ES cells (7
+samples), both with normal diploid karyotypes. The experiments were
+carried out in different laboratories, following either the original Hi-C
+protocol \citep{lieberman2009comprehensive} or the newer \textit{in situ}
+version \citep{rao20143d}, and using different restriction enzymes (DpnII,
+HindIII, MboI and NcoI). We also used array-based copy-number
+segmentation of the two cell lines obtained from the COSMIC database
+\citep{forbes2010cosmic}.
 
@@
 Non-informative contacts were removed applying the following TADbit
-filters: self-circle, dangling-ends, error, extra dangling-end, duplicated
-and random-breaks. For more details, see the methods section of
+filters: \texttt{self-circle}, \texttt{dangling-ends}, \texttt{error},
+\texttt{extra-dangling-ends}, \texttt{duplicated} and
+\texttt{random-breaks}. For more details, see the methods section of
 \cite{stadhouders2017transcription}. In addition, the pipeline is
@@
 The third option is to compute a correlation per distance stratum and then
-obtain a stratum-adjusted correlation coefficient (SCC) as defined in
+obtain a stratum-adjusted correlation coefficient (SCC) as defined by
 \cite{yang2017hicrep}. Finally, the last option, proposed by
@@
 chromosome. Receiver operating characteristic (ROC) curves were computed
-using \texttt{ROCR} package \citep{sing2005rocr}.
+using the \texttt{ROCR} package \citep{sing2005rocr}.
 
@@
 relationship between the amount of contacts and the number of restriction
-enzyme sites in T47D, a breast cancer cell line with aberrant and unstable
+enzyme sites in T47D, a breast cancer cell line with an aberrant
 karyotype.  Four clouds are visible. Each corresponds to sequences present
@@
 chromosome 17 of T47D. \textit{OneD} greatly reduces the wiggling of the
-total amount of contacts.
+total amount of contacts (blue line).
 
@@
 shows the Pearson correlation between the corrected number of contacts and
-the copy number estimation for both T47D and K562 cell lines. Similar
-results were obtained using Spearman correlation (Supplementary Figure 1).
-All the methods except \textit{OneD} decrease the agreement between the
-signal and the copy number because they partially correct the discrepancy.
-In contrast, \textit{OneD} enhances the conformity of the signal with the
-copy number.  Not correcting for variable copy number at that stage may
-seem counter-intuitive, but the tests below will show this leads to better
+the copy number estimation for T47D and K562 (a leukemia cell line with an
+aberrant karyotype). Similar results were obtained using Spearman
+correlation (Supplementary Figure 1).  All the methods except
+\textit{OneD} decreased the agreement between the signal and the copy
+number because they partially corrected the discrepancy. In contrast,
+\textit{OneD} enhanced the conformity of the signal with the copy number.
+Not correcting for variable copy number at that stage may seem
+counter-intuitive, but the tests below will show this leads to better
 performance.
@@
 increase the similarity between biological replicates by reducing
-irrelevant experimental variation, such as batch effects, laboratory of
+irrelevant experimental variance, such as batch effects, laboratory of
 origin and protocol variations. Similarly, a good normalization should
-decrease the similarity between samples that received a different
-treatment to enhance the biological differences.
+decrease the similarity between different samples to enhance the
+biological differences.
 
-We assembled two Hi-C data sets obtained from T47D and K562, two cancer
-cell lines with aberrant karyotypes. In each set we spiked two samples
-from the other cell line (Table~\ref{tab:samples}) to
-introduce biological variability. We compared matrices before and after
-normalization by different methods using the Pearson correlation of
-observed over expected counts (see \ref{sec:comp}). This gave an
-indication of the impact of a given normalization method. The results are
-summarized in Figure~\ref{fig:aberrant}.
+We assembled two Hi-C data sets obtained from T47D and K562 cells.  In
+each set we spiked two samples from the other cell line
+(Table~\ref{tab:samples}) to introduce biological variability. We compared
+matrices before and after normalization by different methods using the
+Pearson correlation of observed over expected counts (see \ref{sec:comp}).
+This gave an indication of the impact of a given normalization method. The
+results are summarized in Figure~\ref{fig:aberrant}.
 
@@
 \ref{fig:aberrant}B and Supplementary Figures~2 and 3). This is an
-undesirable effect, as it obscures the biological variability. In the same
-vein, these methods decreased the similarity between samples that received
-the same treatment (Figure~\ref{fig:aberrant}A), suggesting that the
+undesirable effect, as it obscures the biological variability. Likewise,
+these methods decreased the similarity between samples that received the
+same treatment (Figure~\ref{fig:aberrant}A), suggesting that the
 normalization process is detrimental to the biological signal in these two
@@
 contacts presented the worst performance for all scenarios, and it thus
-seems to be a poor choice as a metric to compare Hi-C matrices.
+seems to be a poor choice for a metric to compare Hi-C matrices.
 
@@
 
-Does the performance of \textit{OneD} for aberrant karyotypes come at the
-cost of decreased performance for normal karyotypes? To address this
+Does the performance of \textit{OneD} on aberrant karyotypes come at the
+cost of decreased performance on normal karyotypes? To address this
 question, we assembled another data set comprised of mouse B cells and
-embryonic stem (ES) cells, both with a normal karyotype The cell types
+embryonic stem (ES) cells, both with a normal karyotype. The cell types
 were pooled in almost equal proportions (see Table~\ref{tab:samples}) and
@@
 depending on the context. Once more, \textit{vanilla} followed the same
-trend as \textit{ICE} to a lesser extent. The \textit{LGF} method
+trend as \textit{ICE} but to a lesser extent. The \textit{LGF} method
 increased the similarity when comparing the same cells with different
@@
 using different protocols. In the other two cases, it had little effect.
-In summary, \textit{OneD} never enhanced the experimental noise and it
+In summary, \textit{OneD} never enhanced the experimental noise and even
 reduced it in one more case than \textit{LGF}.
@@
 except \textit{vanilla}. Using other metrics to compare matrices gave
-similar results (Supplementary Figures 7 and 8). \textit{OneD} was always
-among the top scoring methods. In these conditions, Spearman correlation
+similar results: \textit{OneD} was always among the top scoring methods
+(Supplementary Figures 7 and 8). In these conditions, Spearman correlation
 of contacts again appeared as the worst comparison metric because it
@@
 is not an experimental bias, but it may be considered an additional source
-of bias to be remove. For this reason, \textit{OneD} also allows the user
-to correct the copy number and output a euploid-equivalent matrix. To do
+of bias to be removed. For this reason, \textit{OneD} also allows the user
+to correct the copy number and output an euploid-equivalent matrix. To do
 so, \textit{OneD} segments the linear signal of the total amount of
-contacts into piecewise homoegenous regions. This is carried out by a
+contacts into piecewise homogeneous regions. This is carried out by a
 hidden Markov model whereby the averages of the states are constrained to
-be within a discrete factor of each other. This allows the model to detect
-regions with a number of copies equal to 1, 2, 3 \textit{etc}. With the
-inferred number of copies at hand, \textit{OneD} then normalizes each cell
-of the matrix with equation (\ref{eq:cnvnorm}).
+be an integer number, up to a scaling factor (see \ref{sec:hmm}). This
+allows the model to detect regions with a number of copies equal to 1, 2,
+3 \textit{etc}. With the inferred number of copies at hand, \textit{OneD}
+then normalizes each cell of the matrix with equation (\ref{eq:cnvnorm}).
 
@@
 in regions with higher copy number (Figure~\ref{fig:cnv_correction}). In
-this example, \textit{ICE} overcompensates the original bias at the center
-of the picture and fades the signal almost entirely. Concomitantly, the
-signal at the bottom left of the matrix is enhanced and shows a structure
-that is not visible in the raw data. On the contrary, \textit{LGF}
-strengthen the central region and the diagonal. \textit{OneD} reduces the
-level of the central portion by a factor 2 approximately, but does not
-otherwise distort the main features of the region. This example shows that
-the copy number does not have a simple and predictable effect on the final
-matrix. Not taking it into account may open the door to some normalization
-artifacts. The result provided by \textit{OneD} is not necessarily the
-right one (see Discussion), but at least it does not correct copy number
-variations as a side effect of some other criteria.
+this example, \textit{ICE} overcompensated the original bias at the center
+of the picture and faded the signal almost entirely. Concomitantly, the
+signal at the bottom left of the matrix was enhanced and showed a
+structure that was not visible in the raw data. On the contrary,
+\textit{LGF} strengthened the central region and the diagonal.
+\textit{OneD} reduced the level of the central portion by a factor 2
+approximately, but did not otherwise distort the main features of the
+region. This example shows that the copy number does not have a simple and
+predictable effect on the final matrix. Not taking it into account may
+open the door to some normalization artifacts.
 
@@
 copy number correction, the signal is brought to the same level as the
-flanking regions.
+flanking regions. The result provided by \textit{OneD} is not necessarily
+the right one (see Discussion), but at least it does not correct copy
+number variations as a side effect of some other criteria.
 
-In summary, \textit{OneD} can be used to obtain a euploid-equivalent
-normalized matrix in cases that the effect of the copy number must be
+In summary, \textit{OneD} can be used to obtain an euploid-equivalent
+normalized matrix in cases where the effect of the copy number must be
 removed from the signal.
@@
 normalization methods. \textit{vanilla} and \textit{ICE} have broad
-acceptance for their conceptual simplicity, easy use and speed
+acceptance for their conceptual simplicity, ease of use and speed
 \citep{imakaev2012iterative}. This is even more significant as current
@@
 different methods on the samples described in Table~\ref{tab:samples} at
-100 kbp resolution. The fastest method is \textit{vanilla} and the slowest
-is \textit{LGF}, with an over 100-fold span between the two.
+100 kbp resolution. The fastest method was \textit{vanilla} and the
+slowest is \textit{LGF}, with an over 100-fold span between the two.
 
-\textit{OneD} is the second fastest method and it always ran in less than
-1 min in the conditions of the benchmark. We occasionally encountered
-samples where \textit{ICE} ran faster than \textit{OneD} so it is fair to
-say that none of them has a definite speed advantage. Throughout this
-benchmark, the memory footprint of \textit{OneD} was less than 300 MB.
-Interestingly, the running time of \textit{OneD} is much more homogeneous
-than that of the other methods. The reason is that the size of the
-regression problem to be solved by the \texttt{mgcv} package is always the
-same at fixed resolution.
+\textit{OneD} was the second fastest method and it always ran in less than
+1~min in the conditions of the benchmark. Throughout this benchmark, the
+memory footprint of \textit{OneD} was less than 300 MB.  Interestingly,
+the running time of \textit{OneD} was much more homogeneous than that of
+the other methods. The reason is that the size of the regression problem
+to be solved by the \texttt{mgcv} package is always the same at fixed
+resolution.
 
@@
 slightly faster than \textit{OneD}, and the the rank of the methods
-remained the same at all resolutions. The running time of \textit{OneD}
-remained close to that of \textit{ICE}. Taken together, these results show
+remained the same at all resolutions. Taken together, these results show
 that \textit{OneD} is competitive in terms of computational speed compared
@@
 cells present karyotypic aberrations (Figure~\ref{fig:aberrant}), and that
-it performs equally well as the best methods otherwise
+it performs equally well as the best methods on euploid genomes
 (Figure~\ref{fig:normal}). We also showed that \textit{OneD} is
@@
 a biological signal or an artifact. If the biological sample contains
-karytoypic aberrations, then its genome is grossly different then the
+karytoypic aberrations, then its genome is grossly different from the
 reference genome, which makes signal correction very challenging. The
@@
 
-Overall, \textit{OneD} constitutes a safe choice to normalize Hi-C
-matrices. If the karyotype of the sample is aberrant, it enhances the
-biological variation.  If not, it performs at least equally well as other
-methods in terms of quality and of computational speed.
+Overall, \textit{OneD} constitutes a novel computational approach to
+normalize Hi-C matrices. If the karyotype of the sample is aberrant, it
+enhances the biological variation.  If not, it performs at least equally
+well as other methods in terms of quality and of computational speed.
 
@@
 Resources, M.B.; Data curation, J.Q., Writing – Original Draft, E.V. and
-G.F., Writing – Review \& Editing, F.D. J.Q., R.S., Y.C, T.G., M.M. and
-M.B.; Visualization, E.V.; Supervision, G.F. and M.B., Project
-Administration, M.B.; Funding Acquisition, M.B., T.G., M.M. and G.F.
+G.F., Writing – Review \& Editing, F.D. J.Q., R.S., Y.C, T.G., M.A.M-R.
+and M.B.; Visualization, E.V.; Supervision, G.F. and M.B., Project
+Administration, M.B.; Funding Acquisition, M.B., T.G., M.A.M-R. and G.F.
 

### 2017-06-09 bland.tex
Added bland pdf.

--- 2017-06-09 main.tex
@@
-\PassOptionsToPackage{utf8}{inputenc}
-\documentclass{bioinfo}
-\copyrightyear{2017} \pubyear{2017}
+\documentclass{article}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{amsthm}
+\usepackage[english]{babel}
+\usepackage{graphicx}
+\usepackage{natbib}
+\usepackage{authblk}
+\usepackage{pdflscape}
 
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Original Article}
+\title{OneD: increasing reproducibility of Hi-C Samples with
+abnormal karyotypes}
+
+\author[1,2,*]{Enrique Vidal}
+\author[1,2]{François le Dily}
+\author[1,2]{Javier Quilez}
+\author[1,2]{Ralph Stadhouders}
+\author[1,2]{Yasmina Cuartero}
+\author[1,2]{Thomas Graf}
+\author[1,2,3,4]{Marc A.  Mart\'i-Renom}
+\author[1,2]{Miguel Beato}
+\author[1,2]{Guillaume J. Filion}
+\affil[1]{Gene Regulation, Stem Cells and Cancer Program, Centre for
+Genomic Regulation (CRG), The Barcelona Institute of Science and
+Technology (BIST), Dr. Aiguader 88, 08003, Barcelona, Spain}
+\affil[2]{Universitat Pompeu Fabra (UPF), Barcelona, Spain}
+\affil[3]{CNAG-CRG, Centre for Genomic Regulation (CRG), Barcelona
+Institute of Science and Technology (BIST), Baldiri i Reixac 4, 08028
+Barcelona, Spain}
+\affil[4]{ICREA, Pg. Lluís Companys 23, 08010 Barcelona, Spain}
+\affil[*]{\textbf{Contact:} enrique.vidal@crg.eu}
 
 \begin{document}
-\firstpage{1}
 
-\subtitle{Genome Analysis}
+\maketitle
 
-\title[OneD]{OneD: increasing reproducibility of Hi-C Samples with
-abnormal karyotypes}
-
-\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
-François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
-1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
-Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Marc A.
-Mart\'i-Renom\,$^{\text{\sfb 1,2,3,4}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
-and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}
-
-\address{$^{\text{\sf 1}}$Gene Regulation, Stem Cells and Cancer Program,
-Centre for Genomic Regulation (CRG), The Barcelona Institute of Science
-and Technology (BIST), Dr. Aiguader 88, 08003, Barcelona, Spain \\
-$^{\text{\sf 2}}$Universitat Pompeu Fabra (UPF), Barcelona, Spain \\
-$^{\text{\sf 3}}$CNAG-CRG, Centre for Genomic Regulation (CRG), Barcelona
-Institute of Science and Technology (BIST), Baldiri i Reixac 4, 08028 Barcelona,
-Spain \\
-$^{\text{\sf 4}}$ICREA, Pg. Lluís Companys 23, 08010 Barcelona, Spain}
-
-\corresp{$^\ast$To whom correspondence should be addressed.}
-
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
-
-\editor{Associate Editor: XXXXXXX}
-
-\abstract{
-  \textbf{Motivation:} The three-dimensional conformation of genomes is an
+\begin{abstract}
+The three-dimensional conformation of genomes is an
 essential component of their biological activity. The advent of the Hi-C
@@
 defects on the raw Hi-C maps. The consequences of these widespread
-artifacts on the normalized maps are mostly unexplored. \\
-	\textbf{Results:}
+artifacts on the normalized maps are mostly unexplored.
 We observed that current normalization methods are not robust to the
@@
 safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
-in terms of computing resources for resolutions up to 1 kbp.  \\
-\textbf{Availability:} \textit{OneD} is implemented as an R package
-available at \href{http://www.github.com/qenvio/dryhic}
-{http://www.github.com/qenvio/dryhic}. \\
-\textbf{Contact:} \href{enrique.vidal@crg.eu}{enrique.vidal@crg.eu}\\	
-\textbf{Supplementary information:} Supplementary data are available at
-\textit{Bioinformatics}	online.}
-
-\maketitle
+in terms of computing resources for resolutions up to 1 kbp.
+\textit{OneD} is implemented as an R package
+available at http://www.github.com/qenvio/dryhic.
+\end{abstract}
 
@@
 %\enlargethispage{12pt}
-
-\begin{methods}
 
@@
 
+\begin{landscape}
+
 \begin{table*}
-\processtable{Hi-C data sets used in this study. Sample ID: Unique sample
-identifier in the respective projects. Cell type: source of the biological
-sample. Application: Hi-C protocol. RE: Restriction enzyme used for the
-digestion. Sequencing core: Laboratory performing the experiment (CRG:
-Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor:
-Baylor College of Medicine; UCSD: University of California San Diego;
-LICR: Ludwig Institue for Cancer Research). Set(s): Sets in which the
-sample was included (see text for detail).  Source: SRA, GEO or ENCODE raw
-data identifier.
-\label{tab:samples}}
+\label{tab:samples}
 {\begin{tabular}{ccccccc}
-  \toprule
+  \hline
   \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
@@
   \textbf{Source} \\
-  \midrule
+  \hline
 dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
@@
   NcoI    & LICR & mm10 & SRR443886-SRR443888 \\
-  \botrule
+  \hline
 \end{tabular}}{}
+\caption{Hi-C data sets used in this study. Sample ID: Unique sample
+identifier in the respective projects. Cell type: source of the biological
+sample. Application: Hi-C protocol. RE: Restriction enzyme used for the
+digestion. Sequencing core: Laboratory performing the experiment (CRG:
+Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor:
+Baylor College of Medicine; UCSD: University of California San Diego;
+LICR: Ludwig Institue for Cancer Research). Set(s): Sets in which the
+sample was included (see text for detail).  Source: SRA, GEO or ENCODE raw
+data identifier.}
 \end{table*}
+
+\end{landscape}
 
@@
 
-\end{methods}
-
 %% --------------- Results --------------- %%
@@
 \begin{figure}
-\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
+\centerline{\includegraphics[width=.99\textwidth]{img/figure1.pdf}}
 \caption{Total amount of contacts per bin. A. Non linear relationship
@@
 
-\begin{figure}[b]
-	\centerline{\includegraphics[width=.45\textwidth]
+\begin{figure}
+	\centerline{\includegraphics[width=.99\textwidth]
 {img/copy_number_figure2.pdf}}
@@
 \begin{figure}
-\centerline{\includegraphics[width=.50\textwidth]
+\centerline{\includegraphics[width=.99\textwidth]
   {img/correlation_aberrant_figure3.pdf}}
@@
 \begin{figure}
-\centerline{\includegraphics[width=.50\textwidth]
+\centerline{\includegraphics[width=.99\textwidth]
   {img/correlation_normal_figure4.pdf}}
@@
 \begin{figure}
-\centerline{\includegraphics[width=.5\textwidth]
+\centerline{\includegraphics[width=.99\textwidth]
   {img/figure_cnv_correction.pdf}}
@@
 \begin{figure}
-\centerline{\includegraphics[width=.5\textwidth]
+\centerline{\includegraphics[width=.99\textwidth]
   {img/figure_benchmark_time.pdf}}

### 2017-06-12 manuscript_nar.tex
Added manuscript in NAR template

--- 2017-06-09 bland.tex
@@
-\documentclass{article}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage[english]{babel}
-\usepackage{graphicx}
-\usepackage{natbib}
-\usepackage{authblk}
-\usepackage{pdflscape}
+\documentclass[a4,center,fleqn]{NAR}
 
-\title{OneD: increasing reproducibility of Hi-C Samples with
-abnormal karyotypes}
+% Enter dates of publication
+\copyrightyear{2017}
+\pubdate{** ***** ****}
+\pubyear{2017}
+\jvolume{**}
+\jissue{**}
 
-\author[1,2,*]{Enrique Vidal}
-\author[1,2]{François le Dily}
-\author[1,2]{Javier Quilez}
-\author[1,2]{Ralph Stadhouders}
-\author[1,2]{Yasmina Cuartero}
-\author[1,2]{Thomas Graf}
-\author[1,2,3,4]{Marc A.  Mart\'i-Renom}
-\author[1,2]{Miguel Beato}
-\author[1,2]{Guillaume J. Filion}
-\affil[1]{Gene Regulation, Stem Cells and Cancer Program, Centre for
-Genomic Regulation (CRG), The Barcelona Institute of Science and
-Technology (BIST), Dr. Aiguader 88, 08003, Barcelona, Spain}
-\affil[2]{Universitat Pompeu Fabra (UPF), Barcelona, Spain}
-\affil[3]{CNAG-CRG, Centre for Genomic Regulation (CRG), Barcelona
-Institute of Science and Technology (BIST), Baldiri i Reixac 4, 08028
-Barcelona, Spain}
-\affil[4]{ICREA, Pg. Lluís Companys 23, 08010 Barcelona, Spain}
-\affil[*]{\textbf{Contact:} enrique.vidal@crg.eu}
+%\articlesubtype{This is the article type (optional)}
 
 \begin{document}
+
+\title{OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes}
+
+\author{%
+Enrique Vidal\,$^{1,2,*}$,
+Fran\c{c}ois le Dily\,$^{1,2}$,
+Javier Quilez\,$^{1,2}$,
+Ralph Stadhouders\,$^{1,2}$,
+Yasmina Cuartero\,$^{1,2,3}$,
+Thomas Graf\,$^{1,2}$,
+Marc A.  Mart\'i-Renom\,$^{1,2,3,4}$,
+Miguel Beato\,$^{1,2}$,
+and Guillaume J. Filion\,$^{1,2}$%
+\footnote{To whom correspondence should be addressed.
+Tel: +34 93 316 01 15; Fax: +34 93 316 00 99; Email: enrique.vidal@crg.eu}}
+
+\address{%
+$^{1}$Gene Regulation, Stem Cells and Cancer Program, Centre for
+Genomic Regulation (CRG), The Barcelona Institute of Science and
+Technology (BIST), Dr. Aiguader 88, 08003, Barcelona, Spain
+and
+$^{2}$Universitat Pompeu Fabra (UPF), Barcelona, Spain
+and
+$^{3}$CNAG-CRG, Centre for Genomic Regulation (CRG), Barcelona
+Institute of Science and Technology (BIST), Baldiri i Reixac 4, 08028
+Barcelona, Spain
+and
+$^{4}$ICREA, Pg. Lluís Companys 23, 08010 Barcelona, Spain}
+
+% Affiliation must include:
+% Department name, institution name, full road and district address,
+% state, Zip or postal code, country
+
+\history{%
+Received January 1, ****;
+Revised February 1, ****;
+Accepted March 1, ****}
 
@@
 in terms of computing resources for resolutions up to 1 kbp.
-\textit{OneD} is implemented as an R package
-available at http://www.github.com/qenvio/dryhic.
+%\textit{OneD} is implemented as an R package
+%available at http://www.github.com/qenvio/dryhic.
 \end{abstract}
-
-%% --------------- Introduction --------------- %%
 
@@
 
+% **************************************************************
+% Keep this command to avoid text of first page running into the
+% first page footnotes
+\enlargethispage{-65.1pt}
+% **************************************************************
+
 So far, the only attempt to address the issue was the chromosome-adjusted
@@
 
-%\enlargethispage{12pt}
-
-\section{Methods}
+\section{MATERIALS AND METHODS}
 
@@
 
-\begin{landscape}
+%\begin{landscape}
 
@@
 
-\end{landscape}
+%\end{landscape}
 
@@
 
-We developed the routines contained in the \texttt{dryhic} R package to
+We developed the routines contained in the \texttt{dryhic} R package (available
+at http://www.github.com/qenvio/dryhic) to
 efficiently create sparse representations of contact matrices and further
@@
 
-%% --------------- Results --------------- %%
-
-\section{Results}
+\section{RESULTS}
 
@@
 \begin{figure}
-\centerline{\includegraphics[width=.99\textwidth]{img/figure1.pdf}}
+\centerline{\includegraphics[width=.49\textwidth]{img/figure1.pdf}}
 \caption{Total amount of contacts per bin. A. Non linear relationship
@@
 \begin{figure}
-	\centerline{\includegraphics[width=.99\textwidth]
+	\centerline{\includegraphics[width=.39\textwidth]
 {img/copy_number_figure2.pdf}}
@@
 \begin{figure}
-\centerline{\includegraphics[width=.99\textwidth]
+\centerline{\includegraphics[width=.49\textwidth]
   {img/correlation_aberrant_figure3.pdf}}
@@
 \begin{figure}
-\centerline{\includegraphics[width=.99\textwidth]
+\centerline{\includegraphics[width=.49\textwidth]
   {img/correlation_normal_figure4.pdf}}
@@
 \begin{figure}
-\centerline{\includegraphics[width=.99\textwidth]
+\centerline{\includegraphics[width=.49\textwidth]
   {img/figure_cnv_correction.pdf}}
@@
 \begin{figure}
-\centerline{\includegraphics[width=.99\textwidth]
+\centerline{\includegraphics[width=.49\textwidth]
   {img/figure_benchmark_time.pdf}}
@@
 
-%% --------------- Discussion --------------- %%
-
-\section{Discussion and Conclusions}
+\section{DISCUSSION}
 
@@
 
+\section{CONCLUSION}
+
 Overall, \textit{OneD} constitutes a novel computational approach to
@@
 
-\section*{Acknowledgements}
+\section{ACKNOWLEDGEMENTS}
 
@@
 
-\section*{Author contributions}
-
-Conceptualization, E.V.; Methodology, E.V. and G.F.; Software, E.V. and
-G.F.; Formal Analysis, E.V.; Investigation, E.V., F.D., Y.C. and R.S.;
-Resources, M.B.; Data curation, J.Q., Writing – Original Draft, E.V. and
-G.F., Writing – Review \& Editing, F.D. J.Q., R.S., Y.C, T.G., M.A.M-R.
-and M.B.; Visualization, E.V.; Supervision, G.F. and M.B., Project
-Administration, M.B.; Funding Acquisition, M.B., T.G., M.A.M-R. and G.F.
-
-\section*{Funding}
+\section{FUNDING}
 
@@
 
-\vspace*{-12pt}
+\subsubsection{Conflict of interest statement.} None declared.
+\newpage
 

### 2017-11-18 manuscript_nar.tex
First draft after first round of revisions.

@@
 
-\title{OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes}
+\title{OneD: increasing reproducibility of Hi-C Samples with abnormal
+karyotypes}
 
@@
 and
-$^{4}$ICREA, Pg. Lluís Companys 23, 08010 Barcelona, Spain}
+$^{4}$ICREA, Pg. Llu\'is Companys 23, 08010 Barcelona, Spain}
 
@@
 safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
-in terms of computing resources for resolutions up to 1 kbp.
+in terms of computing resources for resolutions up to 1~kbp.
 %\textit{OneD} is implemented as an R package
@@
 aberrations have so much influence on the outcome that they can be used as
-signatureto re-assemble the target genome \citep{korbel2013genome}. An
+signature to re-assemble the target genome \citep{korbel2013genome}. An
 additional complication is that karyotypic aberrations are not
@@
 \subsection{Model}
-\label{sec:model}
 
@@
 \subsection{Copy number correction}
-\label{sec:hmm}
 
@@
 $t$ variable is fitted on the corrected total amount of contacts per bin
-($\lambda_i'$). The model consists of 7 states that correspond to 1, 2,
-3, 4, 5, 6 and 8 copies of the target, for a total of 3 emission
-parameters (a single scaling parameter, a single standard deviation and a
-single degree of freedom for all the states) and 21 transition parameters.
+($\lambda_i'$). The model consists of 8 states that correspond to 1, 2, 3,
+4, 5, 6, 7, and `8 or more' copies of the target, for a total of 4
+emission parameters (a single position parameter for states 1-7, a
+position parameter for state `8 or more', a single standard deviation for
+all the states and a single degree of freedom for all the states) and 56
+transition parameters.
 
@@
 unpublished Hi-C data of different cell types and organisms
-(Table \ref{tab:samples}).
+(Supplementary Table~1).
 
@@
 %\begin{landscape}
-
-\begin{table*}
-\label{tab:samples}
-{\begin{tabular}{ccccccc}
-  \hline
-  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
-  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set(s)} &
-  \textbf{Source} \\
-  \hline
-dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
-  DpnII & CRG & T47D, K562 & NA \\
-b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
-  DpnII & CRG & T47D & NA \\
-dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
-  DpnII & CRG & T47D, K562 & NA \\
-HindIII\_T0 & T47D & dilution Hi-C &
-  HindIII & CRG & T47D & SRR1054341 \\
-NcoI\_T0    & T47D & dilution Hi-C &
-  NcoI      & CRG & T47D & SRR1054343 \\
-ENCLB758KFU & T47D & dilution Hi-C &
-  HindIII   & UMass & T47D & ENCLB758KFU \\
-ENCLB183QHG & T47D & dilution Hi-C &
-  HindIII   & UMass & T47D & ENCLB183QHG \\
-HIC069  & K562 & \textit{in situ} Hi-C &
-  MboI & Baylor & T47D, K562 &    SRR1658693 \\
-HIC070  & K562 & \textit{in situ} Hi-C &
-  MboI & Baylor & T47D, K562 &    SRR1658694 \\
-HIC071  & K562 & \textit{in situ} Hi-C &
-  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
-HIC074  & K562 & \textit{in situ} Hi-C &
-  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
-b7fa2d8db\_bfac48760 & B cell  & \textit{in situ} Hi-C &
-  DpnII & CRG & mm10 & GSE96611 \\
-fc3e8b36a\_7bf1bf374 & ES cell & \textit{in situ} Hi-C &
-  DpnII & CRG & mm10 & GSE96611 \\
-b7fa2d8db\_7284b867a & B cell  & \textit{in situ} Hi-C &
-  DpnII & CRG & mm10 & GSE96611 \\
-fc3e8b36a\_38bfd1b33 & ES cell & \textit{in situ} Hi-C &
-  DpnII & CRG & mm10 & GSE96611 \\
-b7fa2d8db\_58e812fc2 & B cell  & \textit{in situ} Hi-C &
-  DpnII & CRG & mm10 & GSE96611 \\
-fc3e8b36a\_c990a254e & ES cell & \textit{in situ} Hi-C &
-  DpnII & CRG & mm10 & GSE96611 \\
-b7fa2d8db\_73f11d923 & B cell  & \textit{in situ} Hi-C &
-  DpnII & CRG & mm10 & GSE96611 \\
-fc3e8b36a\_4bf044f18 & ES cell & \textit{in situ} Hi-C &
-  DpnII & CRG & mm10 & GSE96611 \\
-GSM987817 & B cell  & dilution Hi-C &
-  HindIII & UCSD & mm10 & SRR543428-SRR543431 \\
-GSM987818 & B cell  & dilution Hi-C &
-  HindIII & UCSD & mm10 & SRR543432-SRR543442 \\
-GSM862720 & ES cell & dilution Hi-C &
-  HindIII & LICR & mm10 & SRR443883-SRR443885 \\
-GSM862721 & ES cell & dilution Hi-C &
-  HindIII & LICR & mm10 & SRR400251-SRR400256 \\
-GSM862722 & ES cell & dilution Hi-C &
-  NcoI    & LICR & mm10 & SRR443886-SRR443888 \\
-  \hline
-\end{tabular}}{}
-\caption{Hi-C data sets used in this study. Sample ID: Unique sample
-identifier in the respective projects. Cell type: source of the biological
-sample. Application: Hi-C protocol. RE: Restriction enzyme used for the
-digestion. Sequencing core: Laboratory performing the experiment (CRG:
-Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor:
-Baylor College of Medicine; UCSD: University of California San Diego;
-LICR: Ludwig Institue for Cancer Research). Set(s): Sets in which the
-sample was included (see text for detail).  Source: SRA, GEO or ENCODE raw
-data identifier.}
-\end{table*}
-
+%
+%\begin{table*}
+%\label{tab:samples}
+%{\begin{tabular}{ccccccc}
+%  \hline
+%  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
+%  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set(s)} &
+%  \textbf{Source} \\
+%  \hline
+%dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
+%  DpnII & CRG & T47D, K562 & NA \\
+%b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
+%  DpnII & CRG & T47D & NA \\
+%dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
+%  DpnII & CRG & T47D, K562 & NA \\
+%HindIII\_T0 & T47D & dilution Hi-C &
+%  HindIII & CRG & T47D & SRR1054341 \\
+%NcoI\_T0    & T47D & dilution Hi-C &
+%  NcoI      & CRG & T47D & SRR1054343 \\
+%ENCLB758KFU & T47D & dilution Hi-C &
+%  HindIII   & UMass & T47D & ENCLB758KFU \\
+%ENCLB183QHG & T47D & dilution Hi-C &
+%  HindIII   & UMass & T47D & ENCLB183QHG \\
+%HIC069  & K562 & \textit{in situ} Hi-C &
+%  MboI & Baylor & T47D, K562 &    SRR1658693 \\
+%HIC070  & K562 & \textit{in situ} Hi-C &
+%  MboI & Baylor & T47D, K562 &    SRR1658694 \\
+%HIC071  & K562 & \textit{in situ} Hi-C &
+%  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
+%HIC074  & K562 & \textit{in situ} Hi-C &
+%  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
+%b7fa2d8db\_bfac48760 & B cell  & \textit{in situ} Hi-C &
+%  DpnII & CRG & mm10 & GSE96611 \\
+%fc3e8b36a\_7bf1bf374 & ES cell & \textit{in situ} Hi-C &
+%  DpnII & CRG & mm10 & GSE96611 \\
+%b7fa2d8db\_7284b867a & B cell  & \textit{in situ} Hi-C &
+%  DpnII & CRG & mm10 & GSE96611 \\
+%fc3e8b36a\_38bfd1b33 & ES cell & \textit{in situ} Hi-C &
+%  DpnII & CRG & mm10 & GSE96611 \\
+%b7fa2d8db\_58e812fc2 & B cell  & \textit{in situ} Hi-C &
+%  DpnII & CRG & mm10 & GSE96611 \\
+%fc3e8b36a\_c990a254e & ES cell & \textit{in situ} Hi-C &
+%  DpnII & CRG & mm10 & GSE96611 \\
+%b7fa2d8db\_73f11d923 & B cell  & \textit{in situ} Hi-C &
+%  DpnII & CRG & mm10 & GSE96611 \\
+%fc3e8b36a\_4bf044f18 & ES cell & \textit{in situ} Hi-C &
+%  DpnII & CRG & mm10 & GSE96611 \\
+%GSM987817 & B cell  & dilution Hi-C &
+%  HindIII & UCSD & mm10 & SRR543428-SRR543431 \\
+%GSM987818 & B cell  & dilution Hi-C &
+%  HindIII & UCSD & mm10 & SRR543432-SRR543442 \\
+%GSM862720 & ES cell & dilution Hi-C &
+%  HindIII & LICR & mm10 & SRR443883-SRR443885 \\
+%GSM862721 & ES cell & dilution Hi-C &
+%  HindIII & LICR & mm10 & SRR400251-SRR400256 \\
+%GSM862722 & ES cell & dilution Hi-C &
+%  NcoI    & LICR & mm10 & SRR443886-SRR443888 \\
+%  \hline
+%\end{tabular}}{}
+%\caption{Hi-C data sets used in this study. Sample ID: Unique sample
+%identifier in the respective projects. Cell type: source of the biological
+%sample. Application: Hi-C protocol. RE: Restriction enzyme used for the
+%digestion. Sequencing core: Laboratory performing the experiment (CRG:
+%Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor:
+%Baylor College of Medicine; UCSD: University of California San Diego;
+%LICR: Ludwig Institue for Cancer Research). Set(s): Sets in which the
+%sample was included (see text for detail).  Source: SRA, GEO or ENCODE raw
+%data identifier.}
+%\end{table*}
+%
 %\end{landscape}
@@
 \cite{stadhouders2017transcription}. In addition, the pipeline is
-available from the supplementary material published by
+available from the Supplementary Material published by
 \cite{quilez2017managing}.
 
-We developed the routines contained in the \texttt{dryhic} R package (available
-at http://www.github.com/qenvio/dryhic) to
-efficiently create sparse representations of contact matrices and further
-apply \textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
+We developed the routines contained in the \texttt{dryhic} R package
+(available at http://www.github.com/qenvio/dryhic) to efficiently create
+sparse representations of contact matrices and further apply
+\textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
 \texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
@@
 \textit{caICB} corrections respectively. All the results are based on a
-resolution of 100 kbp, but we found no major differences at different
+resolution of 100~kbp, but we found no major differences at different
 resolutions (not shown).
 
+\subsection{Simulations}
+
+Simulations were based on altering four experimental Hi-C samples obtained
+from diploid mouse cell lines (either B cell or ES cell, and from either
+CRG or UCSC). The simulation strategy was the following: First we selected
+uniformly at random the amount of copy number break points (from 3 to 10).
+We then placed the break points along chromosomes 18 and 19 uniformly at
+random. Next, we assigned a copy number (2, 3, 4 or 10 with equal
+probability) to each segment delimited by the break points. We computed
+the outer product of this simulated copy number profile and multiplied it
+element-wise with the original contact matrices of chromosomes 18 and 19.
+The resulting matrices were used as input for correction methods. For each
+simulation (100 in total), we measured the pairwise reproducibility score
+defined by \citet{yan2017hicspector} and computed the average for pairs
+from the same cell type minus the average for pairs from different cell
+types.
+
 \subsection{Comparison of Hi-C matrices}
-\label{sec:comp}
 
@@
 obtain a stratum-adjusted correlation coefficient (SCC) as defined by
-\cite{yang2017hicrep}. Finally, the last option, proposed by
-\cite{yan2017hicspector} is to measure the Pearson correlation between the
-last eigenvectors of the Laplacian of the Hi-C matrix. This approach
-borrows the concepts of spectral clustering \citep{von2007tutorial} and
-amounts to comparing high level features of the matrix.
+\cite{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
+\citet{yan2017hicspector} sums the distances between the leading twenty
+eigenvectors of the Laplacian of the Hi-C matrix. This approach borrows
+the concepts of spectral clustering \citep{von2007tutorial} and amounts to
+comparing high level features of the matrix.
 
@@
 from K562, the second contained the samples from K562 plus two samples
-from T47D, the third contained all the mouse samples (see Table
-\ref{tab:samples} for details). Given a set of experiments and a metric,
-we first computed all pairwise combinations between experiments and then
-classified the comparisons according to the characteristics of each pair
-(cell type, protocol, batch and treatment).
+from T47D, the third contained all the mouse samples (see Supplementary
+Table~1 for details). Given a set of experiments and a metric, we first
+computed all pairwise combinations between experiments and then classified
+the comparisons according to the characteristics of each pair (cell type,
+protocol, batch and treatment).
 
@@
 
-\subsection{Experimental bias correction}
+\subsection{Bias correction in Hi-C experiments}
 
 The principle of \textit{OneD} is to explicitly model Hi-C biases on a
-single variable: the total amount of contacts for each bin of the matrix
-(see \ref{sec:model} for detail). The reason for this choice is that the
-total amount of contacts is approximately proportional to the local copy
-number. For instance, a duplicated region in a diploid genome will show on
-average a 50\% increase in the number of contacts. Discontinuities of the
-amount of contacts thus correspond to changes of the copy number.
+single variable: the total amount of contacts for each bin of the matrix,
+also referred to as the \textit{contact profile}. The reason for this
+choice is that the total amount of contacts is approximately proportional
+to the local copy number. For instance, a duplicated region in a diploid
+genome will show on average a 50\% increase in the number of contacts.
+Discontinuities of the amount of contacts thus correspond to changes of
+the copy number.
 
 \begin{figure}
-\centerline{\includegraphics[width=.49\textwidth]{img/figure1.pdf}}
-\caption{Total amount of contacts per bin. A. Non linear relationship
-between the number of restriction sites and the total number of
-contacts per bin in T47D. B. Total number of contacts per bin on
-chromosome 17 of T47D. The brown line represents the raw
-signal, the blue line represents the signal after bias correction, the
-black line represents the signal after bias and copy number (CN) correction.
-The long arm of chromosome 17 (the region corresponding to 20-80 Mbp) is
-present in four copies, explaining that the signal is about twice higher
-than for the short arm.}
+\centerline{\includegraphics[width=.49\textwidth]
+  {nar_figures/figure_1.eps}}
+\caption{Modeling biases from the contact profile. A. Non linear
+relationship between the number of restriction sites and the total number
+of contacts per bin in T47D. B. Total number of contacts per bin on
+chromosome 17 of T47D. The brown line represents the raw signal, the blue
+line represents the signal after bias correction, the black line
+represents the signal after bias and copy number (CN) correction. The long
+arm of chromosome 17 (the region corresponding to 20-80 Mbp) is present in
+four copies, explaining that the signal is about twice higher than for the
+short arm.}
 \label{fig:totals}
@@
 
-Experimental biases affect the total amount of contacts in a continuous
-but not necessarily linear way. Figure~\ref{fig:totals}A shows the
-relationship between the amount of contacts and the number of restriction
-enzyme sites in T47D, a breast cancer cell line with an aberrant
-karyotype.  Four clouds are visible. Each corresponds to sequences present
-in one to four copies. In all of them, the relationship flattens as the
-number of restriction sites increases. To capture this
-relationship, \textit{OneD} fits a non-linear model between the total
-amount of contacts and the known sources of bias (by default the G+C
-content, the number of restriction sites and the mappability of the
-reads).
+Experimental biases affect the contact profile in a continuous but not
+necessarily linear way. Figure~\ref{fig:totals}A shows the relationship
+between the amount of contacts and the number of restriction enzyme sites
+in T47D, a breast cancer cell line with an aberrant karyotype.  Four
+clouds are visible. Each corresponds to sequences present in one to four
+copies. In all of them, the relationship flattens as the number of
+restriction sites increases. To capture this relationship, \textit{OneD}
+fits a non-linear model between the total amount of contacts and the known
+sources of bias (by default the G+C content, the number of restriction
+sites and the mappability of the reads).
 
 The experimental biases are estimated genome-wide and each cell of the
-matrix is then corrected using equation (\ref{eq:xhat}). Note that the
-corrected amount of contacts is still proportional to the copy number.
-Figure~\ref{fig:totals}B shows the corrected number of contacts along
-chromosome 17 of T47D. \textit{OneD} greatly reduces the wiggling of the
-total amount of contacts (blue line).
+Hi-C matrix is then corrected using equation (\ref{eq:xhat}). Note that
+the corrected contact profile is still proportional to the copy number.
+In cancer cell lines, the contact profile corrected by \textit{OneD} is
+highly correlated to the copy number estimated from an external source
+(Supplementary Figure~1). Figure~\ref{fig:totals}B shows the corrected
+contact profile along chromosome 17 of T47D. \textit{OneD} greatly reduces
+the wiggling of the contact profile (dark blue line).
 
-% Marc's comment here was regarding lack of motivation and explanation for
-% the copy number comparison. I've rescued some text from an older
-% version. We should also comment here about CN being a bias or providing
-% biological information, so to correct or not to correct for CN
+Optionally, \textit{OneD} allows the user to also correct the Hi-C signal
+for the copy number. In this case, a hidden Markov model is fitted on the
+corrected contact profile in order to infer the local copy number. Each
+cell of the Hi-C matrix is then further corrected using equation
+(\ref{eq:cnvnorm}), \textit{i.e.} it is divided by the square root of the
+inferred copy number. In essence, this approach flatens the contact
+profile so that it is no longer proportional to the local copy number. The
+light blue line in figure~\ref{fig:totals}B shows the result of this
+process. After this correction, the contact profile fluctuates around a
+genome-wide constant value.
 
-In what follows, we benchmarked \textit{OneD}, against \textit{vanilla},
-\textit{ICE} \citep{imakaev2012iterative}, \textit{caICB}
-\citep{wu2016computational} and the Local Genomic Features method
-\citep[\textit{LGF},][]{hu2012hicnorm, servant2012hitc}. The first three
-methods correct biases implicitly, whereas the fourth method does it
-explicitly.
+In what follows, we benchmarked \textit{OneD} and \textit{OneD} with copy
+number correction (\textit{OneD+CN}) against \textit{vanilla},
+\textit{ICE} \citep{imakaev2012iterative}, the \textit{KR} algorithm
+\citep{knight2013fast}, \textit{caICB} \citep{wu2016computational} and the
+Local Genomic Features method \citep[\textit{LGF},][]{hu2012hicnorm,
+servant2012hitc}. The first four methods correct biases implicitly,
+whereas the fourth method does it explicitly.
 
-Given that our model is based on total number of contacts, we reasoned
-that a preliminary test would be to check if the corrected number of
-contacts per bin reflects the copy number (as measured by an independent
-technique such as array-based copy-number segmentation). We tested the
-validity of this approach against the Catalogue Of Somatic Mutations In
-Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure \ref{fig:copy_number}
-shows the Pearson correlation between the corrected number of contacts and
-the copy number estimation for T47D and K562 (a leukemia cell line with an
-aberrant karyotype). Similar results were obtained using Spearman
-correlation (Supplementary Figure 1).  All the methods except
-\textit{OneD} decreased the agreement between the signal and the copy
-number because they partially corrected the discrepancy. In contrast,
-\textit{OneD} enhanced the conformity of the signal with the copy number.
-Not correcting for variable copy number at that stage may seem
-counter-intuitive, but the tests below will show this leads to better
-performance.
+We used four metrics to compare Hi-C matrices (see Materials and Methods).
+For consistency and concision, the results based on the reproducibility
+score of \citet{yan2017hicspector} are shown in the main figures, and the
+results based on the other metrics are shown in the Supplementary
+Material.
 
-\begin{figure}
-	\centerline{\includegraphics[width=.39\textwidth]
-{img/copy_number_figure2.pdf}}
-\caption{Pearson correlation between the total number of contacts per bin
-and an independent estimation of the copy number (COSMIC). Left panel
-T47D breast cancer cell line, right panel K562 leukemia cell line.}
-\label{fig:copy_number}
-\end{figure}
+%Given that our model is based on the total number of contacts, we reasoned
+%that a preliminary test would be to check if the corrected number of
+%contacts per bin reflects the copy number (as measured by an independent
+%technique such as array-based copy-number segmentation). We tested the
+%validity of this approach against the Catalogue Of Somatic Mutations In
+%Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure \ref{fig:copy_number}
+%shows the Pearson correlation between the corrected number of contacts and
+%the copy number estimation for T47D and K562 (a leukemia cell line with an
+%aberrant karyotype). Similar results were obtained using Spearman
+%correlation (Supplementary Figure 1).  All the methods except
+%\textit{OneD} decreased the agreement between the signal and the copy
+%number because they partially corrected the discrepancy. In contrast,
+%\textit{OneD} enhanced the conformity of the signal with the copy number.
+%Not correcting for variable copy number at that stage may seem
+%counter-intuitive, but the tests below will show this leads to better
+%performance.
+%
+%
+%\begin{figure}
+%\centerline{\includegraphics[width=.39\textwidth]
+%  {nar_figures/figure_2.eps}}
+%\caption{Pearson correlation between the total number of contacts per bin
+%and an independent estimation of the copy number (COSMIC). Left panel
+%T47D breast cancer cell line, right panel K562 leukemia cell line.}
+%\label{fig:copy_number}
+%\end{figure}
+%
 
@@
 irrelevant experimental variance, such as batch effects, laboratory of
-origin and protocol variations. Similarly, a good normalization should
-decrease the similarity between different samples to enhance the
+origin and protocol variations. Likewise, a good normalization should
+increase the contrast between different samples to enhance the
 biological differences.
 
-We assembled two Hi-C data sets obtained from T47D and K562 cells.  In
-each set we spiked two samples from the other cell line
-(Table~\ref{tab:samples}) to introduce biological variability. We compared
-matrices before and after normalization by different methods using the
-Pearson correlation of observed over expected counts (see \ref{sec:comp}).
-This gave an indication of the impact of a given normalization method. The
-results are summarized in Figure~\ref{fig:aberrant}.
+We collected multiple Hi-C data sets obtained from T47D and K562 cells,
+both with aberrant karyotypes. The first pool, referred to as the T47D data
+set contained six T47D samples and two K562 samples, the second, refered
+to as the K562 data set contained eight T47D samples and two T47D samples.
+We compared matrices before and after normalization by different methods
+(see Materials and Methods, distributions shown in Supplementary Figures~2
+and 3). This gave an indication of the impact of a given normalization
+method on the biological variation (differences between samples from T47D
+and K562) and on the technical variation (differences among samples from
+the same cell type). The results are summarized in
+Figure~\ref{fig:aberrant}.
 
-The \textit{caICB} and \textit{ICE} methods increased the similarity
-between the different cell lines (Figures~\ref{fig:aberrant}A and
-\ref{fig:aberrant}B and Supplementary Figures~2 and 3). This is an
-undesirable effect, as it obscures the biological variability. Likewise,
-these methods decreased the similarity between samples that received the
-same treatment (Figure~\ref{fig:aberrant}A), suggesting that the
-normalization process is detrimental to the biological signal in these two
-cases. The method \textit{vanilla} followed the same trend but to a lesser
-extent, consistent with the fact that it consists of a single \textit{ICE}
-iteration. \textit{OneD} was the only method to increase the similarity
-between experiments carried out on the same material but with a different
-protocol (Figure~\ref{fig:aberrant}A).
+\begin{figure}
+\centerline{\includegraphics[width=.49\textwidth]
+  {nar_figures/figure_2.eps}}
+\caption{Removing biases from Hi-C on aberrant karyotypes. A and B.
+Average changes compared to the raw data on the T47D and K562 data sets.
+The bars represent 95\% confidence intervals centered on the mean
+difference of the reproducibility score between a given correction method
+and the raw data. C and D. ROC curves on the T47D and K562 sets. The areas
+under the curve are indicated in the bottom right corner. The color code
+is the same as in panels A and B.}
+\label{fig:aberrant}
+\end{figure}
+
+On the T47D data set, the methods that increased the reproducibility
+between identical cell types also increased it between different cell
+types (Figures~\ref{fig:aberrant}A), meaning that none of them enhanced
+exclusively the biological signal. However, the difference was most marked
+for \textit{OneD} and \textit{caICB}. On the K562 data set
+(figure~\ref{fig:aberrant}B), both methods increased the reproducibility
+between identical cell types but not between different cell types
+Figures~\ref{fig:aberrant}B), which is the desired behavior of a
+correction method. On both data sets, \textit{OneD+CN}, \textit{ICE} and
+\textit{KR} gave inferior results, while \textit{raw+CN} (copy number
+correction alone), \textit{LGF} and \textit{vanilla} were not competitive.
+The conclusions were unchanged when using other metrics to compare Hi-C
+matrices (Supplementary Figures~4, 5 and 6).
 
@@
 classifier scores and summarized the results with a ROC curve
-(Figures~\ref{fig:aberrant}C and D). All the methods, including the
-absence of normalization, succeeded in identifying the T47D outliers among
-the K562 samples, but recognizing the K562 outliers among the T47D samples
-proved more challenging. \textit{OneD} increased the discrimination power
-compared to the raw matrices, but all the other methods decreased it.
-Actually, they performed little better than random on this task. Using the
-other metrics described in \ref{sec:comp} yielded similar results
-(Supplementary Figures 5, 6 and 7). Note that Spearman correlation of
-contacts presented the worst performance for all scenarios, and it thus
-seems to be a poor choice for a metric to compare Hi-C matrices.
+(Figures~\ref{fig:aberrant}C and D). \textit{OneD} had the larest area
+under the curve in both tests, and the corresponding ROC curve was above
+the others throughout. Using other metrics to compare Hi-C matrices
+yielded similar results (Supplementary Figures~4, 5 and 6).
 
-Taken together, these results show that correcting experimental biases
-with \textit{OneD} enhances the biological variation and reduces the noise
-on samples with an aberrant karyotype.
-
-\begin{figure}
-\centerline{\includegraphics[width=.49\textwidth]
-  {img/correlation_aberrant_figure3.pdf}}
-\caption{
-Results of the comparison between samples with aberrant karyotypes. A and
-B. Average changes compared to the raw data on the T47D and K562 sets. The
-bars represent 95\% confidence intervals centered on the mean difference
-of the correlation score between a given correction method and the raw
-data. C and D. ROC curves on the T47D and K562 sets.  The areas under the
-curve are indicated in the bottom right corner. The color code is the same
-as panels A and B. The brown lines correspond to raw matrices. All results
-in this figure are based on Pearson correlations between the observed over
-expected counts.}
-\label{fig:aberrant}
-\end{figure}
+Taken together, these results show that \textit{OneD} enhances the
+biological variation and reduces the noise on samples with an aberrant
+karyotype.
 
@@
 cost of decreased performance on normal karyotypes? To address this
-question, we assembled another data set comprised of mouse B cells and
-embryonic stem (ES) cells, both with a normal karyotype. The cell types
-were pooled in almost equal proportions (see Table~\ref{tab:samples}) and
-the same tests as above were performed.
+question, we collected data from mouse B cells and embryonic stem (ES)
+cells, both with a normal karyotype. The cell types were pooled in almost
+equal proportions (see Supplementary Table~1) and we performe the same
+tests as above using the same metrics (distributions shown in
+Supplementary Figure~7)
 
-In these conditions, the experimental protocol had a strong effect on the
-impact of the different normalization methods (Figure~\ref{fig:normal}A).
-For instance, \textit{caICB} and \textit{ICE} increased the similarity
-when the protocols were different, but decreased it when the protocols
-were the same. The effect was stronger when comparing identical cell
-types, but the same trend appeared when comparing different cell types,
-indicating that these methods may enhance or reduce biological variation,
-depending on the context. Once more, \textit{vanilla} followed the same
-trend as \textit{ICE} but to a lesser extent. The \textit{LGF} method
-increased the similarity when comparing the same cells with different
-experimental protocols, and had little to no effect in the other cases.
-This indicates that \textit{LGF} is a safe choice in this case.
+In this test, the reproducibility between different cell types was barely
+affected (Figure~\ref{fig:normal}A). All the methods increased the
+reproducibility between identical cell types (except one, see below), but
+only when the experimental conditions were different. This means that all
+the methods were able to remove some technical biases. \textit{ICE} and
+\textit{KR} showed the stongest increase in reproducibility, but
+\textit{OneD}, \textit{OneD+CN}, and \textit{caICB} had competitive
+performance. \textit{vanilla} and \textit{LGF} were less competitive, and
+\textit{raw+CN} (copy number correction alone) performed poorly, because
+it is equivalent to no normalization on diploid cells. Other metrics gave
+similar results (Supplementary Figures~8, 9 and 10).
 
-\textit{OneD} decreased the similarity between different cell types when
-using the same protocol and increased it between identical cell types when
-using different protocols. In the other two cases, it had little effect.
-In summary, \textit{OneD} never enhanced the experimental noise and even
-reduced it in one more case than \textit{LGF}.
-
-When interpreting the similarity scores as classification scores, we
-observed that all the methods could identify approximately 50\% of the
-biological pairs, after which their performance diverged
-(Figure~\ref{fig:normal}B). \textit{OneD} achieved the highest area under
-the curve on this problem, but with a small margin over the other methods
-except \textit{vanilla}. Using other metrics to compare matrices gave
-similar results: \textit{OneD} was always among the top scoring methods
-(Supplementary Figures 7 and 8). In these conditions, Spearman correlation
-of contacts again appeared as the worst comparison metric because it
-showed a lower performance for all the normalization methods.
+As above, we used the reproducibility score for classification and
+compared the tools with a ROC curve (Figure~\ref{fig:normal}B). The
+performance of the all tools were high and less variable than for
+aneuploid cell lines. \textit{OneD} achieved the highest area under the
+curve on this problem, but with a small margin over \textit{KR} and
+\textit{ICE}. The performance of \textit{OneD+CN} was the same as that of
+\textit{OneD} because copy numer correction has no effect on diploid cell
+lines. Using other metrics to compare matrices gave similar results
+(Supplementary Figures~8, 9 and 10).
 
@@
 \centerline{\includegraphics[width=.49\textwidth]
-  {img/correlation_normal_figure4.pdf}}
-\caption{
-Results of the comparison between samples with normal karyotypes. A.
-Average changes compared to the raw data on the mouse data set.  B. ROC
-curves on the mouse data set. The legend is as in
+  {nar_figures/figure_3.eps}}
+\caption{Removing biases from Hi-C on normal karyotypes. A. Average
+changes compared to the raw data on the mouse data set.  B. ROC curves on
+the mouse data set. The legend and color code are as in
 Figure~\ref{fig:aberrant}.}
@@
 
-Removing the explicit experimental biases enhances the importance of the
-copy number in the signal (Figure~\ref{fig:copy_number}). The copy number
-is not an experimental bias, but it may be considered an additional source
-of bias to be removed. For this reason, \textit{OneD} also allows the user
-to correct the copy number and output an euploid-equivalent matrix. To do
-so, \textit{OneD} segments the linear signal of the total amount of
-contacts into piecewise homogeneous regions. This is carried out by a
-hidden Markov model whereby the averages of the states are constrained to
-be an integer number, up to a scaling factor (see \ref{sec:hmm}). This
-allows the model to detect regions with a number of copies equal to 1, 2,
-3 \textit{etc}. With the inferred number of copies at hand, \textit{OneD}
-then normalizes each cell of the matrix with equation (\ref{eq:cnvnorm}).
+The results so far suggest that the copy number correction lowers the
+performance of \textit{OneD}. A possible limitation in Hi-C is the absence
+of ground truth. We thus used simulations to generate matrices with
+defined karyotypic aberrations (see Materials and Methods) and tested the
+capacity of the correction methods to distinguish different cell types
+(Supplementary Figure~11). On this test, \textit{LGF} and \textit{OneD}
+performed best. In comparison, \textit{OneD+CN} performed less well,
+suggesting that correcting the copy number attenuates the biological
+signal. This is corroborated by the fact that the copy number correction
+alone has a lower performance than no correction at all. Thus, copy number
+correction can indeed have side effects that lower the biological signal.
 
-Different normalization methods can either enhance or diminish the signal
-in regions with higher copy number (Figure~\ref{fig:cnv_correction}). In
-this example, \textit{ICE} overcompensated the original bias at the center
-of the picture and faded the signal almost entirely. Concomitantly, the
+However, those conclusions are based on data sets where the karyotype is
+uniform. One sometimes needs to compare samples from cells with different
+karyotype, for instance in cancer samples where the karyotypes may change
+through time. In such conditions, copy number correction can remove a
+feature that is considered irrelevant in order to highlight other
+biological differences.
+
+To test this idea, we collected a Hi-C data set from breast cancer cell
+lines and measured the reproducibility score bewteen identical cell types
+after normalization with the different tools. In this case, we found that
+\textit{OneD+CN} outperformed \textit{OneD} and gave the highest overall
+reproducibility (Figure~\ref{fig:cnv_correction}A). Even though the
+variability is high in this case, this suggests that the karyotypes of
+these cells have diverged, so that the copy number correction performed by
+\textit{OneD} increased the reproducibility between samples from the same
+cell type.
+
+\begin{figure}
+\centerline{\includegraphics[width=.49\textwidth]
+  {nar_figures/figure_4.eps}}
+\caption{Copy number correction. A. Removing biases from Hi-C on breast
+cancer cells. The reproducibility score between samples from the same
+breast cancer cell type was computed after normalization. The vertical bar
+shows the reproducibility score of the raw data. Shown are comparisons for
+different (top) or identical (bottom) experimental conditions. B. Detail
+of a Hi-C matrix normalized with different methods. The central portion
+has an increased copy number, which affects normalization. \textit{ICE}
+fades it away, \textit{LGF} enhances it and \textit{OneD} reduces the
+signal by about half.}
+\label{fig:cnv_correction}
+\end{figure}
+
+Copy number correction is also useful to give euploid-equivalent
+representations of samples from aberrant karyotypes.
+Figure~\ref{fig:cnv_correction}B shows an example where a region at the
+center of the matrix is duplicated. \textit{ICE} overcompensated the
+original bias and faded the signal almost entirely. Concomitantly, the
 signal at the bottom left of the matrix was enhanced and showed a
@@
 approximately, but did not otherwise distort the main features of the
-region. This example shows that the copy number does not have a simple and
-predictable effect on the final matrix. Not taking it into account may
-open the door to some normalization artifacts.
+region.
+
+In summary, \textit{OneD+CN} can improve the reproducibility of samples
+with unstable karyotypes, and it can be used to obtain an
+euploid-equivalent normalized matrix.
+
+\subsection{Capture Hi-C}
+
+Matrix-balancing methods (here \textit{ICE}, \textit{vanilla} and the
+\textit{KR} algorithm) are specifically tailored for Hi-C because it is
+the dominant 3C-derived technology. As a consequence, their performance is
+typically lower when they are used for variants such as and Capture Hi-C
+\citep{jager2015capture}. In this case, the contacts with one or more
+regions of interest are enriched through direct purification, essentially
+erasing all the other contacts from the Hi-C matrix. The lack of
+homogeneity of the signal in such technologies makes normalization
+particularly challenging.
+
+We reasoned that \textit{OneD} should handle such cases gracefully,
+because the signal is formally equivalent to an aberrant karyotype where
+most of the genome is present in zero copy. We thus evaluated the
+performance of \textit{OneD} on a Capture Hi-C data set at 5~kbp
+resolution centered on a 6~Mbp domain of chromosome 6 in T47D cells. We
+found that the reproducibility score between replicates was higher after
+normalization with \textit{OneD} than with matrix-balancing methods
+(figure~\ref{fig:chic}A). The performance of \textit{OneD} was only
+slightly lower than the raw data, suggesting that it barely disturbs the
+biological signal. Other tools were not included beause they either
+crashed (\textit{caICB}) or had prohibitive run time (\textit{LGF}).
 
@@
 \centerline{\includegraphics[width=.49\textwidth]
-  {img/figure_cnv_correction.pdf}}
-\caption{
-Copy number correction A. Detail of a Hi-C matrix normalized with
-different methods. The central portion has an increased copy number, which
-affects normalization. \textit{ICE} fades it away, \textit{LGF} enhances
-it and \textit{OneD} reduces the signal by about half. B. Profile of the
-total amount of contacts after copy numer correction. The plot shows the
-same region as panel A. The brown line represents the raw signal, the blue
-line represents the signal after bias correction, the black line
-represents the signal after bias and copy number (CN) correction.}
-\label{fig:cnv_correction}
+  {nar_figures/figure_5.eps}}
+\caption{Bias removal on Capture Hi-C data A. Performance on the Capture
+Hi-C data. Two replicates centered on a 6~Mbp domain on chromosome 6 in
+T47D cells were normalized by the indicated tools and their
+reproducibility score was computed. B. Performance on virtual Capture Hi-C
+data. Two Hi-C experiments were post-processed to produce an output
+similar to the Capture Hi-C experiment depicted in panel A. The signal
+outside the domain on chromosome 6 was removed and the data were
+normalized (capt), or the data were normalized before removing the signal
+oustside the domain on chromosome 6 (full). Four matrices per method were
+generated and the reproducibility score between the pairs from different
+replicates were computed. Note that the scores are identical for the raw
+data because in this case the matrices of the same replicate are
+identical.}
+\label{fig:chic}
 \end{figure}
 
-Figure~\ref{fig:cnv_correction}B shows the total sum of contacts after
-correction for the copy number with \textit{OneD}. The region around
-72.5-75.0 Mbp showed an elevated amount of contacts in the raw data. After
-copy number correction, the signal is brought to the same level as the
-flanking regions. The result provided by \textit{OneD} is not necessarily
-the right one (see Discussion), but at least it does not correct copy
-number variations as a side effect of some other criteria.
+To gain additional insight, we also set up a ``virtual Capture Hi-C'' data
+set by removing the signal outside the same region of chromosome 6 from
+Hi-C experiments performed in the same cell type. In this setup, we could
+restrict the data and then normalize, or normalize and then restrict the
+data. This allowed us to probe how sensitive the methods are to the shape
+of the input data. Here we found that \textit{OneD} significantly
+outperformed the other methods: it featured the highest reproducibility
+score between replicates with practically no influence from the shape of
+the data (figure~\ref{fig:chic}B). Also, in this case, \textit{OneD}
+enhanced the reproducibility bewteen replicates compared to the raw data.
+This confirms the view that capture methods are a natural framework for
+the statistical model underlying \textit{OneD}. In comparison,
+matrix-balancing methods showed lower reproducibility between replicates,
+together with more variability due to the shape of the data.
 
-In summary, \textit{OneD} can be used to obtain an euploid-equivalent
-normalized matrix in cases where the effect of the copy number must be
-removed from the signal.
+In conclusion, \textit{OneD} is suitable for removing experimental biases
+from data acquired with Capture Hi-C.
 
@@
 Finally, we compared the computational speed of the different
-normalization methods. \textit{vanilla} and \textit{ICE} have broad
-acceptance for their conceptual simplicity, ease of use and speed
+normalization methods. \textit{vanilla} and \textit{ICE} are broadly used
+because of their conceptual simplicity, ease of use and speed
 \citep{imakaev2012iterative}. This is even more significant as current
@@
 
-Unlike the other methods, \textit{OneD} corrects a single variable instead
-of the whole matrix, and thus estimates the model parameters much faster
-than previous explicit methods. We measured the running time of the
-different tools on a 3.3~GHz machine with 62 GB~RAM, always using the
-default parameters. Figure~\ref{fig:times}A shows the running times of the
-different methods on the samples described in Table~\ref{tab:samples} at
-100 kbp resolution. The fastest method was \textit{vanilla} and the
-slowest is \textit{LGF}, with an over 100-fold span between the two.
+\textit{OneD} corrects a single variable instead of the whole matrix, and
+thus estimates the model parameters much faster than previous explicit
+methods. We measured the running time of the different tools on a 3.3~GHz
+machine with 62 GB~RAM, always using the default parameters.
+Figure~\ref{fig:times}A shows the running times of the different methods
+on all the samples shown in Supplementary Figure~1 at 100~kbp resolution.
+The fastest method was \textit{vanilla} and the slowest \textit{LGF}, with
+an over 100-fold span between the two.
 
-\textit{OneD} was the second fastest method and it always ran in less than
-1~min in the conditions of the benchmark. Throughout this benchmark, the
-memory footprint of \textit{OneD} was less than 300 MB.  Interestingly,
-the running time of \textit{OneD} was much more homogeneous than that of
-the other methods. The reason is that the size of the regression problem
-to be solved by the \texttt{mgcv} package is always the same at fixed
-resolution.
-
-We also performed a benchmark on a smaller data set at increasing
-resolution (Figure~\ref{fig:times}B). On this benchmark, \textit{ICE} ran
-slightly faster than \textit{OneD}, and the the rank of the methods
-remained the same at all resolutions. Taken together, these results show
-that \textit{OneD} is competitive in terms of computational speed compared
-to existing methods.
+\textit{OneD} was the third fastest method and it always ran in less than
+1~min in the conditions of the benchmark. The copy number correction
+increased the running time, but it remained unert 1~min in general.
+Throughout this benchmark, the memory footprint of \textit{OneD} was less
+than 300 MB.  Interestingly, the running time of \textit{OneD} was much
+more homogeneous than that of the other methods. The reason is that the
+size of the regression problem to be solved by the \texttt{mgcv} package
+is always the same at fixed resolution.
 
@@
 \centerline{\includegraphics[width=.49\textwidth]
-  {img/figure_benchmark_time.pdf}}
-\caption{
-Computing time of the bias correction methods. A. Total time for the
-entire genome at 100 kbp resolution. Each dot corresponds to one sample.
-The only method faster than ours under performs in all sample comparisons.
-B. Time to correct a reduced genome (chr19-22) of one sample at different
-resolutions. Note the logarithmic scale on the y-axis on both panels.}
+  {nar_figures/figure_6.eps}}
+\caption{Computing time of the bias correction methods. A. Total time to
+process samples listed in Supplementary Table~1 at 100~kbp resolution.
+Each dot corresponds to a sample. B. Total time to process sample HIC070
+at the indicated bin size and sequencing depths. Note the logarithmic
+scale on the y-axis on both panels.}
 \label{fig:times}
 \end{figure}
+
+To better understand these results, we also investigated the influence of
+bin size and sequencing depth on the running time. We performed a
+benchmark on the sample with highest sequencing depth in the data set
+(HIC070, see Supplementary Table~1) at different bin sizes while
+downsampling the contact matrices to mimic lower resolutions. On this
+test, we included only the methods competing with \textit{OneD} in terms
+of speed, \textit{i.e.} the matrix-balacing methods.
+Figure~\ref{fig:times}B shows that the running time of \textit{OneD} does
+not depend on the sequencing depth. Strikingly, \textit{OneD} also gives
+nearly identical results at very different sequencing depths
+(Supplementary Figure~12). \textit{vanilla} was the overall fastest method
+and \textit{OneD} was usually the faster than \textit{ICE} and \textit{KR}
+at high but not at low sequencing depth. At low resolution (100~kbp) the
+speed advantage of \textit{OneD} appeared from low sequencing depth, but
+at high resolution reslution (5~kbp) it appeared only at high sequening
+depth.
+
+Taken together, these results show that \textit{OneD} is competitive in
+terms of speed. It is particularly adapted to projects where the
+sequencing depth is high, as the running time is essentially not affected.
 
@@
 (Figure~\ref{fig:normal}). We also showed that \textit{OneD} is
-approximately as fast as \textit{ICE}, which makes it competitive from the
-point of view of computational speed.
+approximately as fast as \textit{ICE} (Figure~\ref{fig:times}), which
+makes it competitive from the point of view of computational speed.
 
-The originality of \textit{OneD} lies in that it projects all the biases
-onto a single variable: the total amount of contacts per bin. This allows
-greater running speed, while preserving a good performance on samples with
-karyotypic aberrations. One of the reasons why \textit{OneD} is able to
-better highlight the biological distinctions between samples is that it
-only corrects the copy number if specifically requested by the user. The
-impact of copy number variations on normalization is rather opaque,
-especially if they are treated as implicit biases
-(Figure~\ref{fig:cnv_correction}). Treating them as explicit biases with
-optional removal seems to be an overall safer strategy.
+The originality of \textit{OneD} lies in the estimation of the explicit
+biases and the copy number from a single variable: the contact profile.
+This allows greater running speed, while preserving a good performance on
+samples with karyotypic aberrations. One of the reasons why \textit{OneD}
+is able to better highlight the biological distinctions between samples is
+that it only corrects the copy number if specifically requested by the
+user. The impact of copy number variations on normalization is rather
+opaque, especially if they are treated as implicit biases
+(Figure~\ref{fig:cnv_correction}B). Treating them as explicit biases with
+optional removal seems to be an overall safer strategy. In homogeneous
+data sets, correcting for the copy number can blur the biological signal,
+but when karyotypes are variable or unstable, it may increase the
+reproducibility (Figure~\ref{fig:cnv_correction}A).
 

### 2017-11-19 manuscript_nar.tex
adding GM results and minor typos

@@
 Thomas Graf\,$^{1,2}$,
-Marc A.  Mart\'i-Renom\,$^{1,2,3,4}$,
+Marc A.  Marti-Renom\,$^{1,2,3,4}$,
 Miguel Beato\,$^{1,2}$,
@@
 safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
-in terms of computing resources for resolutions up to 1~kbp.
+in terms of computing resources for resolutions up to 2~kbp.
 %\textit{OneD} is implemented as an R package
@@
 normalization methods. Finally, the implementation is as fast as
-\textit{ICE} and it scales up to 1~kbp resolution with reasonable
+\textit{ICE} and it scales up to 5~kbp resolution with reasonable
 computing resources.
@@
 $t$ variable is fitted on the corrected total amount of contacts per bin
-($\lambda_i'$). The model consists of 8 states that correspond to 1, 2, 3,
+($\hat{t}_i = \sum_j^n{\hat{x}_{ij}}$). The model consists of 8 states that correspond to 1, 2, 3,
 4, 5, 6, 7, and `8 or more' copies of the target, for a total of 4
@@
 
-We used several experiments comprising T47D breast cancer cell lines (7
-samples), K562 leukemia cell lines (4 samples), both with aberrant
-karyotypes; and mouse primary B cells (6 samples ) and ES cells (7
-samples), both with normal diploid karyotypes. The experiments were
+We used several experiments comprising T47D breast cancer cell lines (6
+samples), K562 leukemia cell lines (8 samples), both with aberrant
+karyotypes; mouse primary B cells (6 samples) and ES cells (7
+samples), both with normal diploid karyotypes; GM12878 B-lymphoblastoid cells
+(58 samples), with diploid karyotype; BT474, MCF10, MCF-7 and SKBR3 breast cell
+lines (1 sample each). The experiments were
 carried out in different laboratories, following either the original Hi-C
@@
 version \citep{rao20143d}, and using different restriction enzymes (DpnII,
-HindIII, MboI and NcoI). We also used array-based copy-number
+HindIII, MspI, MboI and NcoI). We also used array-based copy-number
 segmentation of the two cell lines obtained from the COSMIC database
@@
 sparse representations of contact matrices and further apply
-\textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
+\textit{vanilla}, \textit{ICE}, \textit{KR} and \textit{oneD} corrections.
 \texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
@@
 
-We defined three data sets to measure experimental reproducibility after
+We defined five data sets to measure experimental reproducibility after
 normalization: The first contained the samples from T47D plus two samples
 from K562, the second contained the samples from K562 plus two samples
-from T47D, the third contained all the mouse samples (see Supplementary
+from T47D, the third contained all the mouse samples, the fourth contained all
+GM12878 samples, the fifth contained one sample of each breast cells (see Supplementary
 Table~1 for details). Given a set of experiments and a metric, we first
@@
 chromosome. Receiver operating characteristic (ROC) curves were computed
-using the \texttt{ROCR} package \citep{sing2005rocr}.
+using the \texttt{ROCR} package \citep{sing2005rocr} for the three first sets
+(those including samples of different cell types).
 
@@
 servant2012hitc}. The first four methods correct biases implicitly,
-whereas the fourth method does it explicitly.
+whereas the fifth method does it explicitly.
 
@@
 between identical cell types but not between different cell types
-Figures~\ref{fig:aberrant}B), which is the desired behavior of a
+%Figures~\ref{fig:aberrant}B)
+, which is the desired behavior of a
 correction method. On both data sets, \textit{OneD+CN}, \textit{ICE} and
@@
 classifier scores and summarized the results with a ROC curve
-(Figures~\ref{fig:aberrant}C and D). \textit{OneD} had the larest area
+(Figures~\ref{fig:aberrant}C and D). \textit{OneD} had the largest area
 under the curve in both tests, and the corresponding ROC curve was above
@@
 
+We took advantage of the wide range of protocols used in the GM12878 set to
+further investigate the performance of each method. The differences between
+methods were minor and \textit{OneD} showed the greatest improvement in
+reporducibility (Supplementary Figure~11).
+
 Taken together, these results indicate that \textit{OneD} performs as well
@@
 capacity of the correction methods to distinguish different cell types
-(Supplementary Figure~11). On this test, \textit{LGF} and \textit{OneD}
+(Supplementary Figure~12). On this test, \textit{LGF} and \textit{OneD}
 performed best. In comparison, \textit{OneD+CN} performed less well,
@@
 
-\textit{OneD} was the third fastest method and it always ran in less than
+On average, \textit{OneD} was the second fastest method and it always ran in less than
 1~min in the conditions of the benchmark. The copy number correction
-increased the running time, but it remained unert 1~min in general.
+increased the running time, but it remained under 1~min in general.
 Throughout this benchmark, the memory footprint of \textit{OneD} was less
@@
 nearly identical results at very different sequencing depths
-(Supplementary Figure~12). \textit{vanilla} was the overall fastest method
-and \textit{OneD} was usually the faster than \textit{ICE} and \textit{KR}
+(Supplementary Figure~13). \textit{vanilla} was the overall fastest method
+and \textit{OneD} was usually faster than \textit{ICE} and \textit{KR}
 at high but not at low sequencing depth. At low resolution (100~kbp) the

### 2017-11-20 manuscript_nar.tex
Minor edits.

@@
 
-Matrix-balancing methods (here \textit{ICE}, \textit{vanilla} and the
-\textit{KR} algorithm) are specifically tailored for Hi-C because it is
-the dominant 3C-derived technology. As a consequence, their performance is
-typically lower when they are used for variants such as and Capture Hi-C
+Matrix-balancing methods (here \textit{ICE}, \textit{vanilla} and
+\textit{KR}) are specifically tailored for Hi-C because it is the dominant
+3C-derived technology. As a consequence, their performance is typically
+lower when they are used for variants such as Capture Hi-C
 \citep{jager2015capture}. In this case, the contacts with one or more
@@
 slightly lower than the raw data, suggesting that it barely disturbs the
-biological signal. Other tools were not included beause they either
-crashed (\textit{caICB}) or had prohibitive run time (\textit{LGF}).
+biological signal. Other tools were not included beause they had
+prohibitive run time.
 

### 2017-11-22 manuscript_nar.tex
adding supp figures and fixing abstract
Adding @jaquol comments

--- 2017-11-19 manuscript_nar.tex
@@
 safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
-in terms of computing resources for resolutions up to 2~kbp.
+in terms of computing resources for resolutions up to 5~kbp.
 %\textit{OneD} is implemented as an R package
@@
 simplest metric is the Spearman correlation applied to intra-chromosomal
-contacts up to a given distance (5 Mb in what follows). The second option
+contacts up to a given distance (5 Mbp in what follows). The second option
 is to measure the similarity of observed over expected contacts via the
@@
 of contacts per bin in T47D. B. Total number of contacts per bin on
-chromosome 17 of T47D. The brown line represents the raw signal, the blue
-line represents the signal after bias correction, the black line
-represents the signal after bias and copy number (CN) correction. The long
-arm of chromosome 17 (the region corresponding to 20-80 Mbp) is present in
-four copies, explaining that the signal is about twice higher than for the
-short arm.}
+chromosome 17 of T47D. The purple line represents the raw signal, the dark blue
+line represents the signal after OneD bias correction, the light blue line
+represents the signal after OneD bias and copy number (CN) correction and
+the orange line represents the signal after ICE bias correction. The long arm
+of chromosome 17 (the region corresponding to 20-80 Mbp) is present in
+four copies, explaining that the signal is about twice higher than for
+the short arm.}
 \label{fig:totals}
@@
 set contained six T47D samples and two K562 samples, the second, refered
-to as the K562 data set contained eight T47D samples and two T47D samples.
+to as the K562 data set contained eight K562 samples and two T47D samples.
 We compared matrices before and after normalization by different methods
@@
 difference of the reproducibility score between a given correction method
-and the raw data. C and D. ROC curves on the T47D and K562 sets. The areas
-under the curve are indicated in the bottom right corner. The color code
-is the same as in panels A and B.}
+and the raw data. Upper left panels: Different cell type samples
+processed using different protocols. Upper right panels: Same cell
+type samples processed using different protocols. Lower left panels:
+Same cell type samples processed using the same protocol. C and D. ROC
+curves on the T47D and K562 sets. The areas under the curve are
+indicated in the bottom right corner. The color code is the same as in
+panels A and B.}
 \label{fig:aberrant}
@@
 compared the tools with a ROC curve (Figure~\ref{fig:normal}B). The
-performance of the all tools were high and less variable than for
+performance of all the tools were high and less variable than for
 aneuploid cell lines. \textit{OneD} achieved the highest area under the
@@
 \caption{Removing biases from Hi-C on normal karyotypes. A. Average
-changes compared to the raw data on the mouse data set.  B. ROC curves on
-the mouse data set. The legend and color code are as in
-Figure~\ref{fig:aberrant}.}
+  changes compared to the raw data on the mouse data set.
+  Upper left panel: Different cell type samples
+processed using different protocols. Upper right panel: Same cell
+type samples processed using different protocols. Lower left panel:
+Same cell type samples processed using the same protocol. Lower right panel:
+Different cell type samples processed using the same protocol. B. ROC curves on
+the mouse data set. The legend and color code are as in Figure~\ref{fig:aberrant}.}
 \label{fig:normal}

### 2017-11-23 manuscript_nar.tex
Minor changes (need the commit to merge).

--- 2017-11-20 manuscript_nar.tex
@@
 
-The originality of \textit{OneD} lies in the estimation of the explicit
-biases and the copy number from a single variable: the contact profile.
-This allows greater running speed, while preserving a good performance on
-samples with karyotypic aberrations. One of the reasons why \textit{OneD}
-is able to better highlight the biological distinctions between samples is
-that it only corrects the copy number if specifically requested by the
-user. The impact of copy number variations on normalization is rather
-opaque, especially if they are treated as implicit biases
-(Figure~\ref{fig:cnv_correction}B). Treating them as explicit biases with
-optional removal seems to be an overall safer strategy. In homogeneous
-data sets, correcting for the copy number can blur the biological signal,
-but when karyotypes are variable or unstable, it may increase the
-reproducibility (Figure~\ref{fig:cnv_correction}A).
+The originality of \textit{OneD} lies in the estimation from a single
+variable of the explicit biases with a Negative Binomial model and of the
+copy number. This allows greater running speed, while preserving a good
+performance on samples with karyotypic aberrations. One of the reasons why
+\textit{OneD} is able to better highlight the biological distinctions
+between samples is that it only corrects the copy number if specifically
+requested by the user. The impact of copy number variations on
+normalization is rather opaque, especially if they are treated as implicit
+biases (Figure~\ref{fig:cnv_correction}B). Treating them as explicit
+biases with optional removal seems to be an overall safer strategy. In
+homogeneous data sets, correcting for the copy number can blur the
+biological signal, but when karyotypes are variable or unstable, it may
+increase the reproducibility (Figure~\ref{fig:cnv_correction}A).
 

### 2017-11-23 supplementary_figures_nar.tex

\PassOptionsToPackage{utf8}{inputenc}
\documentclass[12pt]{report}

\usepackage[margin=2cm]{geometry}

\usepackage{graphicx}

\renewcommand{\thefigure}{S\arabic{figure}}

\begin{document}

\begin{figure}[b]
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_1.eps}}
	\caption{
    Pearson correlation between contact profiles and independent copy number
	estimation (COSMIC). A. T47D breast cancer cell line (6 samples). B. K562
	leukemia cell line (8 samples). Each dot represents a sample. The new
	proposal (in dark blue) outperforms the rest of alternatives.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_2.eps}}
    \caption{
    Distribution of the pair-wise comparisons between the samples included in
    the T47D set. X axis: similarity metric. Y axis: correction
    method. Each panel groups pairs of samples with the corresponding
    characteristics (in terms of cell type and protocol). A. Pearson correlation
    between observed over expected counts. B. Spearman correlation between
    observed counts. C. Stratum-adjusted correlation coefficient (SCC) between
    observed counts. D. Reproducibility score of observed counts.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_3.eps}}
    \caption{
    Distribution of the pair-wise comparisons between the samples included in
    the K562 set. X axis: similarity metric. Y axis: correction
    method. Each panel groups pairs of samples with the corresponding
    characteristics (in terms of cell type and protocol). A. Pearson correlation
    between observed over expected counts. B. Spearman correlation between
    observed counts. C. Stratum-adjusted correlation coefficient (SCC) between
    observed counts. D. Reproducibility score of observed counts.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_4.eps}}
    \caption{
    Removing biases from Hi-C on aberrant karyotypes (Spearman correlation). A and B. Average changes
    compared to the raw data on the T47D and K562 data sets. The bars represent
    95\% confidence intervals centered on the mean difference of the
    spearman correlation between a given correction method and the raw data. C
    and D. ROC curves on the T47D and K562 sets. The areas under the curve are
    indicated in the bottom right corner. The color code is the same as in panels A and B.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_5.eps}}
    \caption{
    Removing biases from Hi-C on aberrant karyotypes (Pearson correlation). A and B. Average changes
    compared to the raw data on the T47D and K562 data sets. The bars represent
    95\% confidence intervals centered on the mean difference of the
    Pearson correlation of the observed over expected between a given correction method and the raw data. C
    and D. ROC curves on the T47D and K562 sets. The areas under the curve are
    indicated in the bottom right corner. The color code is the same as in panels A and B.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_6.eps}}
    \caption{
    Removing biases from Hi-C on aberrant karyotypes (SCC). A and B. Average changes
    compared to the raw data on the T47D and K562 data sets. The bars represent
    95\% confidence intervals centered on the mean difference of the
    stratum-adjusted correlation coefficient (SCC) between a given correction method and the raw data. C
    and D. ROC curves on the T47D and K562 sets. The areas under the curve are
    indicated in the bottom right corner. The color code is the same as in panels A and B.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_7.eps}}
    \caption{
    Distribution of the pair-wise comparisons between the samples included in
    the mm10 set. X axis: similarity metric. Y axis: correction
    method. Each panel groups pairs of samples with the corresponding
    characteristics (in terms of cell type and protocol). A. Pearson correlation
    between observed over expected counts. B. Spearman correlation between
    observed counts. C. Stratum-adjusted correlation coefficient (SCC) between
    observed counts. D. Reproducibility score of observed counts.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_8.eps}}
    \caption{
    Removing biases from Hi-C on normal karyotypes (Spearman correlation). A. Average changes
    compared to the raw data on the mm10 data set. The bars represent
    95\% confidence intervals centered on the mean difference of the
    Spearman correlation between a given correction method and the raw
    data. B. ROC curves on the mm10 set. The areas under the curve are indicated
    in the bottom right corner. The color code is the same as in panel A.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_9.eps}}
    \caption{
    Removing biases from Hi-C on normal karyotypes (Pearson correlation). A. Average changes
    compared to the raw data on the mm10 data set. The bars represent
    95\% confidence intervals centered on the mean difference of the
    Pearson correlation of the observed over expected correlation between a given correction method and the raw
    data. B. ROC curves on the mm10 set. The areas under the curve are indicated
    in the bottom right corner. The color code is the same as in panel A.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_10.eps}}
    \caption{
    Removing biases from Hi-C on normal karyotypes (SCC). A. Average changes
    compared to the raw data on the mm10 data set. The bars represent
    95\% confidence intervals centered on the mean difference of the
    stratum-adjusted correlation coefficient (SCC) between a given correction method and the raw
    data. B. ROC curves on the mm10 set. The areas under the curve are indicated
    in the bottom right corner. The color code is the same as in panel A.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_11.eps}}
    \caption{
    Removing biases from Hi-C on GM12878 samples. X axis: reproducibility
    score. Y axis: correction method. The bars represent 95\% confidence intervals centered on the mean of the
    reproducibility score of each correction method. The vertical dark purple
    line marks the reproducibility score for the uncorrected data (raw). Upper panel: pairs of
    samples processed using different protocols. Lower panel: pairs of samples
    processed using the same protocol.}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_12.eps}}
    \caption{
    Distribution of the simulation results. X axis: Difference in
    reproducibility score between pairs of samples of the same cell type and
    pairs of samples of different cell type. Y axis: correction method}
\end{figure}

\begin{figure}
	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_13.eps}}
    \caption{Correlation between subsamples at different resolutions and
    depths. X axis: sequencing depth. Y axis: Pearson correlation between the
    correction vectors of two different subsamples. Each color represents a
    correction method and each panel a bining resolution. Note the Y axis limits
    are specific to each panel.}
\end{figure}

\end{document}

### 2017-11-23 supplementary_figures_nar.tex
Edited legends of supplementary figures.

@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_1.eps}}
-	\caption{
-    Pearson correlation between contact profiles and independent copy number
-	estimation (COSMIC). A. T47D breast cancer cell line (6 samples). B. K562
-	leukemia cell line (8 samples). Each dot represents a sample. The new
-	proposal (in dark blue) outperforms the rest of alternatives.}
+  \caption{ Pearson correlation between contact profiles and independent
+copy number estimation from the COSMIC data base (Forbes \textit{et al}.,
+2010). A. T47D breast cancer cell line (6 samples). B. K562 leukemia cell
+line (8 samples). Each dot represents a sample normalized by a different
+tool. The data normalized by \textit{OneD} correlates better with the copy
+number than the unnormalized data.}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_2.eps}}
-    \caption{
-    Distribution of the pair-wise comparisons between the samples included in
-    the T47D set. X axis: similarity metric. Y axis: correction
+  \caption{ Distribution of the pair-wise comparisons between the samples
+included in the T47D set. X axis: similarity metric. Y axis: correction
 method. Each panel groups pairs of samples with the corresponding
-    characteristics (in terms of cell type and protocol). A. Pearson correlation
-    between observed over expected counts. B. Spearman correlation between
-    observed counts. C. Stratum-adjusted correlation coefficient (SCC) between
-    observed counts. D. Reproducibility score of observed counts.}
+characteristics (in terms of cell type and protocol). A. Pearson
+correlation between observed over expected counts. B. Spearman correlation
+between observed counts. C. Stratum-adjusted correlation coefficient (SCC)
+between observed counts. D. Reproducibility score of observed counts.}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_3.eps}}
-    \caption{
-    Distribution of the pair-wise comparisons between the samples included in
-    the K562 set. X axis: similarity metric. Y axis: correction
-    method. Each panel groups pairs of samples with the corresponding
-    characteristics (in terms of cell type and protocol). A. Pearson correlation
-    between observed over expected counts. B. Spearman correlation between
-    observed counts. C. Stratum-adjusted correlation coefficient (SCC) between
-    observed counts. D. Reproducibility score of observed counts.}
-\end{figure}
+  \caption{Distribution of the pair-wise comparisons between the
+samples included in the K562 set. X axis: similarity metric. Y axis:
+correction method. Each panel groups pairs of samples with the
+corresponding characteristics (in terms of cell type and protocol). A.
+Pearson correlation between observed over expected counts. B. Spearman
+correlation between observed counts. C. Stratum-adjusted correlation
+coefficient (SCC) between observed counts. D. Reproducibility score of
+observed counts.} \end{figure}
 
@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_4.eps}}
-    \caption{
-    Removing biases from Hi-C on aberrant karyotypes (Spearman correlation). A and B. Average changes
-    compared to the raw data on the T47D and K562 data sets. The bars represent
-    95\% confidence intervals centered on the mean difference of the
-    spearman correlation between a given correction method and the raw data. C
-    and D. ROC curves on the T47D and K562 sets. The areas under the curve are
-    indicated in the bottom right corner. The color code is the same as in panels A and B.}
+    \caption{Removing biases from Hi-C on aberrant karyotypes (Spearman
+correlation). A and B. Average changes compared to the raw data on the
+T47D and K562 data sets. The bars represent 95\% confidence intervals
+centered on the mean difference of the Spearman correlation between a
+given correction method and the raw data. C and D. ROC curves on the T47D
+and K562 sets. The areas under the curve are indicated in the bottom right
+corner. The color code is the same as in panels A and B.}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_5.eps}}
-    \caption{
-    Removing biases from Hi-C on aberrant karyotypes (Pearson correlation). A and B. Average changes
-    compared to the raw data on the T47D and K562 data sets. The bars represent
-    95\% confidence intervals centered on the mean difference of the
-    Pearson correlation of the observed over expected between a given correction method and the raw data. C
+  \caption{Removing biases from Hi-C on aberrant karyotypes (Pearson
+correlation). A and B. Average changes compared to the raw data on the
+T47D and K562 data sets. The bars represent 95\% confidence intervals
+centered on the mean difference of the Pearson correlation of the observed
+over expected counts between a given correction method and the raw data. C
 and D. ROC curves on the T47D and K562 sets. The areas under the curve are
-    indicated in the bottom right corner. The color code is the same as in panels A and B.}
+indicated in the bottom right corner. The color code is the same as in
+panels A and B.}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_6.eps}}
-    \caption{
-    Removing biases from Hi-C on aberrant karyotypes (SCC). A and B. Average changes
-    compared to the raw data on the T47D and K562 data sets. The bars represent
-    95\% confidence intervals centered on the mean difference of the
-    stratum-adjusted correlation coefficient (SCC) between a given correction method and the raw data. C
-    and D. ROC curves on the T47D and K562 sets. The areas under the curve are
-    indicated in the bottom right corner. The color code is the same as in panels A and B.}
+  \caption{Removing biases from Hi-C on aberrant karyotypes (SCC). A and
+B. Average changes compared to the raw data on the T47D and K562 data
+sets. The bars represent 95\% confidence intervals centered on the mean
+difference of the stratum-adjusted correlation coefficient (SCC) between a
+given correction method and the raw data. C and D. ROC curves on the T47D
+and K562 sets. The areas under the curve are indicated in the bottom right
+corner. The color code is the same as in panels A and B.}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_7.eps}}
-    \caption{
-    Distribution of the pair-wise comparisons between the samples included in
-    the mm10 set. X axis: similarity metric. Y axis: correction
-    method. Each panel groups pairs of samples with the corresponding
-    characteristics (in terms of cell type and protocol). A. Pearson correlation
-    between observed over expected counts. B. Spearman correlation between
-    observed counts. C. Stratum-adjusted correlation coefficient (SCC) between
-    observed counts. D. Reproducibility score of observed counts.}
+  \caption{Distribution of the pair-wise comparisons between the samples
+included in the mm10 data set. X axis: similarity metric. Y axis:
+correction method. Each panel groups pairs of samples with the
+corresponding characteristics (in terms of cell type and protocol). A.
+Pearson correlation between observed over expected counts. B. Spearman
+correlation between observed counts. C. Stratum-adjusted correlation
+coefficient (SCC) between observed counts. D. Reproducibility score of
+observed counts.}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_8.eps}}
-    \caption{
-    Removing biases from Hi-C on normal karyotypes (Spearman correlation). A. Average changes
-    compared to the raw data on the mm10 data set. The bars represent
-    95\% confidence intervals centered on the mean difference of the
-    Spearman correlation between a given correction method and the raw
-    data. B. ROC curves on the mm10 set. The areas under the curve are indicated
-    in the bottom right corner. The color code is the same as in panel A.}
+  \caption{Removing biases from Hi-C on normal karyotypes (Spearman
+correlation). A. Average changes compared to the raw data on the mm10 data
+set. The bars represent 95\% confidence intervals centered on the mean
+difference of the Spearman correlation between a given correction method
+and the raw data. B. ROC curves on the mm10 data set. The areas under the
+curve are indicated in the bottom right corner. The color code is the same
+as in panel A.}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_9.eps}}
-    \caption{
-    Removing biases from Hi-C on normal karyotypes (Pearson correlation). A. Average changes
-    compared to the raw data on the mm10 data set. The bars represent
-    95\% confidence intervals centered on the mean difference of the
-    Pearson correlation of the observed over expected correlation between a given correction method and the raw
-    data. B. ROC curves on the mm10 set. The areas under the curve are indicated
-    in the bottom right corner. The color code is the same as in panel A.}
+  \caption{Removing biases from Hi-C on normal karyotypes (Pearson
+correlation). A. Average changes compared to the raw data on the mm10 data
+set. The bars represent 95\% confidence intervals centered on the mean
+difference of the Pearson correlation of the observed over expected counts
+between a given correction method and the raw data. B. ROC curves on the
+mm10 data set. The areas under the curve are indicated in the bottom right
+corner. The color code is the same as in panel A.}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_10.eps}}
-    \caption{
-    Removing biases from Hi-C on normal karyotypes (SCC). A. Average changes
-    compared to the raw data on the mm10 data set. The bars represent
-    95\% confidence intervals centered on the mean difference of the
-    stratum-adjusted correlation coefficient (SCC) between a given correction method and the raw
-    data. B. ROC curves on the mm10 set. The areas under the curve are indicated
-    in the bottom right corner. The color code is the same as in panel A.}
+  \caption{Removing biases from Hi-C on normal karyotypes (SCC). A.
+Average changes compared to the raw data on the mm10 data set. The bars
+represent 95\% confidence intervals centered on the mean difference of the
+stratum-adjusted correlation coefficient (SCC) between a given correction
+method and the raw data. B. ROC curves on the mm10 data set. The areas
+under the curve are indicated in the bottom right corner. The color code
+is the same as in panel A.}
 \end{figure}

### 2017-11-24 manuscript_nar.tex
Adding supp tables in excel with legend + last changes

--- 2017-11-22 manuscript_nar.tex
@@
 safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
-in terms of computing resources for resolutions up to 5~kbp.
+in terms of computing resources for resolutions up to 5~Kbp.
 %\textit{OneD} is implemented as an R package
@@
 normalization methods. Finally, the implementation is as fast as
-\textit{ICE} and it scales up to 5~kbp resolution with reasonable
+\textit{ICE} and it scales up to 5~Kbp resolution with reasonable
 computing resources.
@@
 $\{f_k(\cdot)\}$ are estimated jointly with the negative binomial
-dispersion parameter $\theta$ using the \texttt{mgcv} package
-\citep{wood2011fast} of the R software \citep{coreteam2014r}.
+dispersion parameter $\theta$ using the the default arguments of the
+\texttt{mgcv::s} and \texttt{mgcv::nb} functions \citep{wood2011fast} of the R
+software \citep{coreteam2014r}.
 
@@
 used to fit the model are the local G+C content, the read mappability and
-the content and number of restriction sites. The model and the
-implementation can be modifed or extended with any user-provided genomic
+the number of restriction sites. The model and the
+implementation can be modified or extended with any user-provided genomic
 features.
@@
 \textit{caICB} corrections respectively. All the results are based on a
-resolution of 100~kbp, but we found no major differences at different
+resolution of 100~Kbp, but we found no major differences at different
 resolutions (not shown).
@@
 Simulations were based on altering four experimental Hi-C samples obtained
-from diploid mouse cell lines (either B cell or ES cell, and from either
-CRG or UCSC). The simulation strategy was the following: First we selected
+from diploid mouse cell lines (b7fa2d8db\_bfac48760, fc3e8b36a\_7bf1bf374,
+GSM987817 and GSM862720), either B cell or ES cell, and from either
+CRG or UCSC. The simulation strategy was the following: First we selected
 uniformly at random the amount of copy number break points (from 3 to 10).
@@
 fixed effect was the normalization method and the random effect was the
-chromosome. Receiver operating characteristic (ROC) curves were computed
-using the \texttt{ROCR} package \citep{sing2005rocr} for the three first sets
-(those including samples of different cell types).
+chromosome. To test the ability of the different methods to separate samples
+from different cell origin we generated receiver operating characteristic (ROC) curves
+using the similarity metric (e.g. reproducibility score) as the classifier score
+and the relative cell type (i.e. same or different cell type) as the binary
+classification. ROC curves were computed using the \texttt{ROCR} package
+\citep{sing2005rocr} for the three first sets (those including samples of
+different cell types).
 
@@
 (\ref{eq:cnvnorm}), \textit{i.e.} it is divided by the square root of the
-inferred copy number. In essence, this approach flatens the contact
+inferred copy number. In essence, this approach flattens the contact
 profile so that it is no longer proportional to the local copy number. The
@@
 both with aberrant karyotypes. The first pool, referred to as the T47D data
-set contained six T47D samples and two K562 samples, the second, refered
+set contained six T47D samples and two K562 samples, the second, referred
 to as the K562 data set contained eight K562 samples and two T47D samples.
@@
 the same cell type). The results are summarized in
-Figure~\ref{fig:aberrant}.
+Figure~\ref{fig:aberrant} and Supplementary Table~3.
 
@@
 cells, both with a normal karyotype. The cell types were pooled in almost
-equal proportions (see Supplementary Table~1) and we performe the same
+equal proportions (see Supplementary Table~1) and we performed the same
 tests as above using the same metrics (distributions shown in
@@
 the methods were able to remove some technical biases. \textit{ICE} and
-\textit{KR} showed the stongest increase in reproducibility, but
+\textit{KR} showed the strongest increase in reproducibility, but
 \textit{OneD}, \textit{OneD+CN}, and \textit{caICB} had competitive
@@
 it is equivalent to no normalization on diploid cells. Other metrics gave
-similar results (Supplementary Figures~8, 9 and 10).
+similar results (Supplementary Figures~8, 9 and 10, Supplementary Table~3).
 
@@
 \textit{ICE}. The performance of \textit{OneD+CN} was the same as that of
-\textit{OneD} because copy numer correction has no effect on diploid cell
+\textit{OneD} because copy number correction has no effect on diploid cell
 lines. Using other metrics to compare matrices gave similar results
@@
 methods were minor and \textit{OneD} showed the greatest improvement in
-reporducibility (Supplementary Figure~11).
+reproducibility (Supplementary Figure~11).
 
@@
 To test this idea, we collected a Hi-C data set from breast cancer cell
-lines and measured the reproducibility score bewteen identical cell types
+lines and measured the reproducibility score between identical cell types
 after normalization with the different tools. In this case, we found that
@@
 
+Besides matrix correction, copy number estimation could have a potential stand-alone
+utility. For instance, we applied \textit{OneD+CN} to a T47D sample
+(dc3a1e069\_51720e9cf) at 10 Kbp and compared the correlation with the
+array-based copy number. The correlation increased from .80 for the raw totals,
+to .88 for the \textit{OneD} corrected totals to .92 for the estimated CN. We
+provide the table of the estimated start and end points as Supplementary Table~
+2.
+
 In summary, \textit{OneD+CN} can improve the reproducibility of samples
@@
 most of the genome is present in zero copy. We thus evaluated the
-performance of \textit{OneD} on a Capture Hi-C data set at 5~kbp
+performance of \textit{OneD} on a Capture Hi-C data set at 5~Kbp
 resolution centered on a 6~Mbp domain of chromosome 6 in T47D cells. We
@@
 slightly lower than the raw data, suggesting that it barely disturbs the
-biological signal. Other tools were not included beause they either
+biological signal. Other tools were not included beaus they either
 crashed (\textit{caICB}) or had prohibitive run time (\textit{LGF}).
@@
 normalized (capt), or the data were normalized before removing the signal
-oustside the domain on chromosome 6 (full). Four matrices per method were
+outside the domain on chromosome 6 (full). Four matrices per method were
 generated and the reproducibility score between the pairs from different
@@
 the data (figure~\ref{fig:chic}B). Also, in this case, \textit{OneD}
-enhanced the reproducibility bewteen replicates compared to the raw data.
+enhanced the reproducibility between replicates compared to the raw data.
 This confirms the view that capture methods are a natural framework for
@@
 Figure~\ref{fig:times}A shows the running times of the different methods
-on all the samples shown in Supplementary Figure~1 at 100~kbp resolution.
+on all the samples shown in Supplementary Figure~1 at 100~Kbp resolution.
 The fastest method was \textit{vanilla} and the slowest \textit{LGF}, with
@@
 \caption{Computing time of the bias correction methods. A. Total time to
-process samples listed in Supplementary Table~1 at 100~kbp resolution.
+process samples listed in Supplementary Table~1 at 100~Kbp resolution.
 Each dot corresponds to a sample. B. Total time to process sample HIC070
@@
 test, we included only the methods competing with \textit{OneD} in terms
-of speed, \textit{i.e.} the matrix-balacing methods.
+of speed, \textit{i.e.} the matrix-balancing methods.
 Figure~\ref{fig:times}B shows that the running time of \textit{OneD} does
@@
 and \textit{OneD} was usually faster than \textit{ICE} and \textit{KR}
-at high but not at low sequencing depth. At low resolution (100~kbp) the
+at high but not at low sequencing depth. At low resolution (100~Kbp) the
 speed advantage of \textit{OneD} appeared from low sequencing depth, but
-at high resolution reslution (5~kbp) it appeared only at high sequening
+at high resolution resolution (5~Kbp) it appeared only at high sequencing
 depth.

### 2017-11-24 manuscript_nar.tex

@@
 slightly lower than the raw data, suggesting that it barely disturbs the
-biological signal. Other tools were not included beaus they either
-crashed (\textit{caICB}) or had prohibitive run time (\textit{LGF}).
+biological signal. Other tools were not included beause they had
+prohibitive run time.
 

### 2017-11-24 manuscript_nar.tex
Changing to NAR references

@@
 To test the correction of biases, we gathered a set of
-published \citep{ledily2014distinct, encode2012integrated, rao20143d,
-stadhouders2017transcription, lin2012global, dixon2012topological} and
+published \citep[][BioRxiv: https://doi.org/10.1101/132456]{ledily2014distinct, encode2012integrated, rao20143d,
+lin2012global, dixon2012topological} and
 unpublished Hi-C data of different cell types and organisms
@@
 \texttt{random-breaks}. For more details, see the methods section of
-\cite{stadhouders2017transcription}. In addition, the pipeline is
+(BioRxiv: https://doi.org/10.1101/132456). In addition, the pipeline is
 available from the Supplementary Material published by
@@
 
-\bibliographystyle{natbib}
+\bibliographystyle{nar}
 \bibliography{references}

### 2017-11-24 manuscript_nar.tex
Last edits up to results (still need to do last sections).

@@
 content. This approach is used in the method of
-\cite{yaffe2011probabilistic} and in Hicnorm by \cite{hu2012hicnorm}. The
-second strategy is to implicitly correct unknown biases by enforcing some
-regularity condition on the data. This approach is used in the
+\citet{yaffe2011probabilistic} and in HiCNorm by \citet{hu2012hicnorm}.
+The second strategy is to implicitly correct unknown biases by enforcing
+some regularity condition on the data. This approach is used in the
 Iterative Correction and Eigenvector decomposition method (\textit{ICE})
-of \cite{imakaev2012iterative}, whereby the total amount of contacts of
+of \citet{imakaev2012iterative}, whereby the total amount of contacts of
 every bin is imposed to be the same. \textit{ICE} is currently the most
@@
 aberrations have so much influence on the outcome that they can be used as
-signature to re-assemble the target genome \citep{korbel2013genome}. An
+signatures to re-assemble the target genome \citep{korbel2013genome}. An
 additional complication is that karyotypic aberrations are not
@@
 Iterative Correction Bias method (\textit{caICB}) of
-\cite{wu2016computational}. However, \textit{caICB} applies a uniform
+\citet{wu2016computational}. However, \textit{caICB} applies a uniform
 chromosome-wide copy number correction, effectively excluding the numerous
@@
 
-Our approach is to model the tally of contacts for each bin, thus reducing
-the matrix to a one dimension score (hence the name \textit{OneD}). We
-assume that the total number of contacts per bin ($t_{i}$) can be
-approximated by a negative binomial distribution. This choice is sensible
-because the amount of contacts is a discrete variable and because the
-negative binomial distribution allows for overdispersion. We further
-assume that the explicit sources of bias have independent contributions to
-the mean of the distribution for a given bin ($\lambda_i$).
+Our approach is to model the total number of contacts for each bin, thus
+reducing the matrix to a one-dimension score (hence the name
+\textit{OneD}) referred to as the `contact profile'. We assume that the
+total number of contacts per bin $t_{i}$ can be approximated by a
+negative binomial distribution. This choice is sensible because the amount
+of contacts is a discrete variable and because the negative binomial
+distribution allows for overdispersion. We further assume that the
+explicit sources of bias have independent contributions to the mean of the
+distribution for a given bin $\lambda_i$.
 
 Given that this relationship might not be linear (see for instance
-Figure~\ref{fig:totals}A), we allowed a smooth representation
-using thin plate penalized regression splines \citep{wood2003thin} in a
-generalized additive model \citep{wood2011fast}. The model can be
+Figure~\ref{fig:totals}A), we allowed a smooth representation using thin
+plate penalized regression splines \citep{wood2003thin} in a generalized
+additive model \citep{wood2011fast}. Thin plate regression splines are
+`isotropic' because rotation of the covariate co-ordinate system does not
+change the result of smoothing and `low rank' because they have far fewer
+coefficients than there are data points to smooth. The model can be
 parametrized as
+\begin{gather*}
+t_i = \sum_{j=1}^n{x_{ij}} \sim  NB(\lambda_i, \theta) \text{ and} \\
+\log(\lambda_i) \propto \sum_{k}{f_k(z_{k,i})},
+\end{gather*}
+where $x_{ij}$ is the raw number of contacts between bins $i$ and $j$, and
+$z_{k,i}$ is the additive bias of genomic feature $k$ in bin $i$.
 
-\begin{align*}
-t_i = \sum_j^n{x_{ij}} &\sim  NB(\lambda_i, \theta) \\
-\log(\lambda_i) &\propto \sum_{k}{f_k(z_k)}
-\end{align*}
-
-\noindent
-where $x_{ij}$ is the raw number of contacts between bins $i$ and $j$, and
-$z_k$ is the additive bias of genomic feature $k$. The smooth functions
-$\{f_k(\cdot)\}$ are estimated jointly with the negative binomial
-dispersion parameter $\theta$ using the the default arguments of the
-\texttt{mgcv::s} and \texttt{mgcv::nb} functions \citep{wood2011fast} of the R
-software \citep{coreteam2014r}.
+The smooth functions $\{f_k(\cdot)\}$ are estimated through the thin plate
+spline penalty, jointly with the negative binomial dispersion parameter
+$\theta$ using the the default arguments of the \texttt{mgcv::s} and
+\texttt{mgcv::nb} functions \citep{wood2011fast} of the R software
+\citep{coreteam2014r}.
 
@@
 $\{\lambda_i'\}$ that can be used to compute the corrected counts
-($\hat{x}_{ij}$).
-
+$\hat{x}_{ij}$.
 \begin{align}
@@
 In line with previous methods
-\citep{yaffe2011probabilistic,hu2012hicnorm}, the default features
-used to fit the model are the local G+C content, the read mappability and
-the number of restriction sites. The model and the
-implementation can be modified or extended with any user-provided genomic
-features.
+\citep{yaffe2011probabilistic,hu2012hicnorm}, the default features used to
+fit the model are the local G+C content, the read mappability and the
+number of restriction sites. The model and the implementation can be
+modified or extended with any user-provided genomic features.
 
@@
 $t$ variable is fitted on the corrected total amount of contacts per bin
-($\hat{t}_i = \sum_j^n{\hat{x}_{ij}}$). The model consists of 8 states that correspond to 1, 2, 3,
-4, 5, 6, 7, and `8 or more' copies of the target, for a total of 4
-emission parameters (a single position parameter for states 1-7, a
-position parameter for state `8 or more', a single standard deviation for
-all the states and a single degree of freedom for all the states) and 56
-transition parameters.
+$\hat{t}_i = \sum_{j=1}^n{\hat{x}_{ij}}$. The model consists of 8
+states that correspond to 1, 2, 3, 4, 5, 6, 7, and `8 or more' copies of
+the target bin, for a total of 4 emission parameters (a single position
+parameter for states 1-7, a position parameter for state `8 or more', a
+single standard deviation for all the states and a single degree of
+freedom for all the states) and 56 transition parameters.
 
@@
 \citep{filion2010systematic}.  The Viterbi path is then computed and
-corresponds to the inferred copy number of each bin ($c_i$).
+corresponds to the inferred copy number of each bin $c_i$.
 
 A correction equal to the square root of the copy number is then applied
-to the whole matrix. More specifically, each cell is updated to
-
+to the whole matrix. More specifically, the entry at position $(i,j)$ is
+updated to
 \begin{equation}
@@
 
-To test the correction of biases, we gathered a set of
-published \citep[][BioRxiv: https://doi.org/10.1101/132456]{ledily2014distinct, encode2012integrated, rao20143d,
-lin2012global, dixon2012topological} and
-unpublished Hi-C data of different cell types and organisms
-(Supplementary Table~1).
+We gathered a set of published Hi-C data \citep[][BioRxiv:
+https://doi.org/10.1101/132456]{ledily2014distinct, encode2012integrated,
+rao20143d, lin2012global, dixon2012topological} and unpublished data of
+different cell types and organisms (Supplementary Table~1).
 
-We used several experiments comprising T47D breast cancer cell lines (6
-samples), K562 leukemia cell lines (8 samples), both with aberrant
-karyotypes; mouse primary B cells (6 samples) and ES cells (7
-samples), both with normal diploid karyotypes; GM12878 B-lymphoblastoid cells
-(58 samples), with diploid karyotype; BT474, MCF10, MCF-7 and SKBR3 breast cell
-lines (1 sample each). The experiments were
-carried out in different laboratories, following either the original Hi-C
-protocol \citep{lieberman2009comprehensive} or the newer \textit{in situ}
-version \citep{rao20143d}, and using different restriction enzymes (DpnII,
-HindIII, MspI, MboI and NcoI). We also used array-based copy-number
-segmentation of the two cell lines obtained from the COSMIC database
-\citep{forbes2010cosmic}.
+The Hi-C experiments were performed in T47D breast cancer cell lines
+(aberrant karyotype, six samples), K562 leukemia cell lines (aberrant
+karyotype, eight samples), mouse primary B cells (normal karyotyp, six
+samples) ES cells (normal karotype, seven samples), GM12878
+B-lymphoblastoid cells (normal karyotype, fifty eight samples), BT474
+breast cancer cell line (aberrant karyotype, one sample), MCF10 breast
+cancer cell line (aberrant karyotype, one sample), MCF-7 cancer cell line
+(aberrant karyotype, one sample) and SKBR3 breast cell lines (aberrant
+karyotype, one sample).
 
-%\begin{landscape}
-%
-%\begin{table*}
-%\label{tab:samples}
-%{\begin{tabular}{ccccccc}
-%  \hline
-%  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
-%  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set(s)} &
-%  \textbf{Source} \\
-%  \hline
-%dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
-%  DpnII & CRG & T47D, K562 & NA \\
-%b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
-%  DpnII & CRG & T47D & NA \\
-%dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
-%  DpnII & CRG & T47D, K562 & NA \\
-%HindIII\_T0 & T47D & dilution Hi-C &
-%  HindIII & CRG & T47D & SRR1054341 \\
-%NcoI\_T0    & T47D & dilution Hi-C &
-%  NcoI      & CRG & T47D & SRR1054343 \\
-%ENCLB758KFU & T47D & dilution Hi-C &
-%  HindIII   & UMass & T47D & ENCLB758KFU \\
-%ENCLB183QHG & T47D & dilution Hi-C &
-%  HindIII   & UMass & T47D & ENCLB183QHG \\
-%HIC069  & K562 & \textit{in situ} Hi-C &
-%  MboI & Baylor & T47D, K562 &    SRR1658693 \\
-%HIC070  & K562 & \textit{in situ} Hi-C &
-%  MboI & Baylor & T47D, K562 &    SRR1658694 \\
-%HIC071  & K562 & \textit{in situ} Hi-C &
-%  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
-%HIC074  & K562 & \textit{in situ} Hi-C &
-%  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
-%b7fa2d8db\_bfac48760 & B cell  & \textit{in situ} Hi-C &
-%  DpnII & CRG & mm10 & GSE96611 \\
-%fc3e8b36a\_7bf1bf374 & ES cell & \textit{in situ} Hi-C &
-%  DpnII & CRG & mm10 & GSE96611 \\
-%b7fa2d8db\_7284b867a & B cell  & \textit{in situ} Hi-C &
-%  DpnII & CRG & mm10 & GSE96611 \\
-%fc3e8b36a\_38bfd1b33 & ES cell & \textit{in situ} Hi-C &
-%  DpnII & CRG & mm10 & GSE96611 \\
-%b7fa2d8db\_58e812fc2 & B cell  & \textit{in situ} Hi-C &
-%  DpnII & CRG & mm10 & GSE96611 \\
-%fc3e8b36a\_c990a254e & ES cell & \textit{in situ} Hi-C &
-%  DpnII & CRG & mm10 & GSE96611 \\
-%b7fa2d8db\_73f11d923 & B cell  & \textit{in situ} Hi-C &
-%  DpnII & CRG & mm10 & GSE96611 \\
-%fc3e8b36a\_4bf044f18 & ES cell & \textit{in situ} Hi-C &
-%  DpnII & CRG & mm10 & GSE96611 \\
-%GSM987817 & B cell  & dilution Hi-C &
-%  HindIII & UCSD & mm10 & SRR543428-SRR543431 \\
-%GSM987818 & B cell  & dilution Hi-C &
-%  HindIII & UCSD & mm10 & SRR543432-SRR543442 \\
-%GSM862720 & ES cell & dilution Hi-C &
-%  HindIII & LICR & mm10 & SRR443883-SRR443885 \\
-%GSM862721 & ES cell & dilution Hi-C &
-%  HindIII & LICR & mm10 & SRR400251-SRR400256 \\
-%GSM862722 & ES cell & dilution Hi-C &
-%  NcoI    & LICR & mm10 & SRR443886-SRR443888 \\
-%  \hline
-%\end{tabular}}{}
-%\caption{Hi-C data sets used in this study. Sample ID: Unique sample
-%identifier in the respective projects. Cell type: source of the biological
-%sample. Application: Hi-C protocol. RE: Restriction enzyme used for the
-%digestion. Sequencing core: Laboratory performing the experiment (CRG:
-%Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor:
-%Baylor College of Medicine; UCSD: University of California San Diego;
-%LICR: Ludwig Institue for Cancer Research). Set(s): Sets in which the
-%sample was included (see text for detail).  Source: SRA, GEO or ENCODE raw
-%data identifier.}
-%\end{table*}
-%
-%\end{landscape}
+The experiments were carried out in different laboratories, following
+either the original Hi-C protocol \citep{lieberman2009comprehensive} or
+the newer \textit{in situ} version \citep{rao20143d}, and using different
+restriction enzymes (DpnII, HindIII, MspI, MboI and NcoI). In the figures,
+``same protocol'' means same laboratory, same Hi-C protocol and same
+restriction enzyme, and ``different protocol'' means that any of the three
+is different.
 
-%\subsection{Data processing}
+We also used array-based copy-number segmentation of the two cell lines
+obtained from the COSMIC database \citep{forbes2010cosmic} as an external
+reference for validation.
 
@@
 FASTQ files, paired-end reads were mapped to the corresponding reference
-genome (hg38 and mm10) taking into account the restriction enzyme site.
+genome (hg38 or mm10) taking into account the restriction enzyme site.
 Non-informative contacts were removed applying the following TADbit
@@
 \texttt{extra-dangling-ends}, \texttt{duplicated} and
-\texttt{random-breaks}. For more details, see the methods section of
-(BioRxiv: https://doi.org/10.1101/132456). In addition, the pipeline is
-available from the Supplementary Material published by
-\cite{quilez2017managing}.
+\texttt{random-breaks} (for more details see BioRxiv:
+https://doi.org/10.1101/132456). In addition, the pipeline is available
+from the Supplementary Material published by \citet{quilez2017managing}.
 
@@
 \citep{wu2016computational} were used to carry out the \textit{LGF} and
-\textit{caICB} corrections respectively. All the results are based on a
-resolution of 100~Kbp, but we found no major differences at different
-resolutions (not shown).
+\textit{caICB} corrections respectively. When not specified, the
+resolution of the analysis is 100~Kbp.
 
@@
 from diploid mouse cell lines (b7fa2d8db\_bfac48760, fc3e8b36a\_7bf1bf374,
-GSM987817 and GSM862720), either B cell or ES cell, and from either
-CRG or UCSC. The simulation strategy was the following: First we selected
-uniformly at random the amount of copy number break points (from 3 to 10).
-We then placed the break points along chromosomes 18 and 19 uniformly at
-random. Next, we assigned a copy number (2, 3, 4 or 10 with equal
-probability) to each segment delimited by the break points. We computed
-the outer product of this simulated copy number profile and multiplied it
-element-wise with the original contact matrices of chromosomes 18 and 19.
-The resulting matrices were used as input for correction methods. For each
-simulation (100 in total), we measured the pairwise reproducibility score
-defined by \citet{yan2017hicspector} and computed the average for pairs
-from the same cell type minus the average for pairs from different cell
-types.
+GSM987817 and GSM862720, see Supplementary Table~1), that were performed
+in either B cells or ES cells, and from either CRG or UCSC. The simulation
+strategy was the following: First we selected uniformly at random the
+amount of copy number break points (from 3 to 10). We then placed the
+break points along chromosomes 18 and 19 uniformly at random. Next, we
+assigned a copy number (2, 3, 4 or 10 with equal probability) to each
+segment delimited by the break points. We computed the outer product of
+this simulated copy number profile and multiplied it element-wise with the
+original contact matrices of chromosomes 18 and 19. The resulting matrices
+were used as input for correction methods. For each simulation (100 in
+total), we measured the pairwise reproducibility score defined by
+\citet{yan2017hicspector} and computed the average for pairs from the same
+cell type minus the average for pairs from different cell types.
 
@@
 obtain a stratum-adjusted correlation coefficient (SCC) as defined by
-\cite{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
+\citet{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
 \citet{yan2017hicspector} sums the distances between the leading twenty
@@
 from K562, the second contained the samples from K562 plus two samples
-from T47D, the third contained all the mouse samples, the fourth contained all
-GM12878 samples, the fifth contained one sample of each breast cells (see Supplementary
-Table~1 for details). Given a set of experiments and a metric, we first
-computed all pairwise combinations between experiments and then classified
-the comparisons according to the characteristics of each pair (cell type,
-protocol, batch and treatment).
+from T47D, the third contained all the mouse samples, the fourth contained
+all the GM12878 samples and the fifth contained one sample of each breast
+cancer cells (see Supplementary Table~1 for details). Given a set of
+experiments and a metric, we first computed all pairwise combinations
+between experiments and then classified the comparisons according to the
+characteristics of each pair (cell type, protocol, batch and treatment).
 
@@
 fixed effect was the normalization method and the random effect was the
-chromosome. To test the ability of the different methods to separate samples
-from different cell origin we generated receiver operating characteristic (ROC) curves
-using the similarity metric (e.g. reproducibility score) as the classifier score
-and the relative cell type (i.e. same or different cell type) as the binary
-classification. ROC curves were computed using the \texttt{ROCR} package
-\citep{sing2005rocr} for the three first sets (those including samples of
-different cell types).
+chromosome. To test the ability of the different methods to separate
+samples from different cell origin we generated receiver operating
+characteristic (ROC) curves using the similarity metric (e.g.
+reproducibility score) as the classifier score and the relative cell type
+(i.e. same \textit{vs} different) as the binary classification. ROC curves
+were computed using the \texttt{ROCR} package \citep{sing2005rocr}.
 

### 2017-11-25 manuscript_nar.tex
Minor changes.

@@
 
-The experimental biases are estimated genome-wide and each cell of the
+The experimental biases are estimated genome-wide and each entry of the
 Hi-C matrix is then corrected using equation (\ref{eq:xhat}). Note that
@@
 (Supplementary Figure~1). Figure~\ref{fig:totals}B shows the corrected
-contact profile along chromosome 17 of T47D. \textit{OneD} greatly reduces
-the wiggling of the contact profile (dark blue line).
+contact profile along chromosome 17 of T47D, where \textit{OneD} greatly
+reduces the fluctuations (dark blue line).
 
@@
 corrected contact profile in order to infer the local copy number. Each
-cell of the Hi-C matrix is then further corrected using equation
+entry of the Hi-C matrix is then further corrected using equation
 (\ref{eq:cnvnorm}), \textit{i.e.} it is divided by the square root of the

### 2018-01-04 manuscript_nar.tex
first changes for R2

@@
 negative binomial distribution. This choice is sensible because the amount
-of contacts is a discrete variable and because the negative binomial
-distribution allows for overdispersion. We further assume that the
-explicit sources of bias have independent contributions to the mean of the
-distribution for a given bin $\lambda_i$.
+of contacts is a discrete variable that often presents a variance higher
+than the mean \citep{carty2017integrated} (a phenomenon known as
+overdispersion), and because the negative binomial distribution allows
+for overdispersion \citep{love2014moderated}. We further assume that
+the explicit sources of bias have independent contributions to the mean
+of the distribution for a given bin $\lambda_i$.
 
@@
 sparse representations of contact matrices and further apply
-\textit{vanilla}, \textit{ICE}, \textit{KR} and \textit{oneD} corrections.
+\textit{vanilla} (a single iteration of ICE), \textit{ICE}, \textit{KR} and \textit{oneD} corrections.
 \texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}

### 2018-01-08 manuscript_nar.tex
Answering R2 reviewer's comments

@@
 content. This approach is used in the method of
-\citet{yaffe2011probabilistic} and in HiCNorm by \citet{hu2012hicnorm}.
+\cite{yaffe2011probabilistic} and in HiCNorm by \cite{hu2012hicnorm}.
 The second strategy is to implicitly correct unknown biases by enforcing
@@
 Iterative Correction and Eigenvector decomposition method (\textit{ICE})
-of \citet{imakaev2012iterative}, whereby the total amount of contacts of
+of \cite{imakaev2012iterative}, whereby the total amount of contacts of
 every bin is imposed to be the same. \textit{ICE} is currently the most
-popular method, due in part to its speed.
+popular method, due in part to its speed and simplicity.
 
@@
 Iterative Correction Bias method (\textit{caICB}) of
-\citet{wu2016computational}. However, \textit{caICB} applies a uniform
+\cite{wu2016computational}. However, \textit{caICB} applies a uniform
 chromosome-wide copy number correction, effectively excluding the numerous
@@
 bins. The values are stored in the cells of the contact matrix ($x_{ij}$),
-quantifying the interaction between the two loci at positions $i$ and
-$j$.
+quantifying the interaction frequency between the two loci at positions
+$i$ and $j$.
 
@@
 (aberrant karyotype, six samples), K562 leukemia cell lines (aberrant
-karyotype, eight samples), mouse primary B cells (normal karyotyp, six
+karyotype, eight samples), mouse primary B cells (normal karyotype, six
 samples) ES cells (normal karotype, seven samples), GM12878
@@
 https://doi.org/10.1101/132456). In addition, the pipeline is available
-from the Supplementary Material published by \citet{quilez2017managing}.
+from the Supplementary Material published by \cite{quilez2017managing}.
 
@@
 GSM987817 and GSM862720, see Supplementary Table~1), that were performed
-in either B cells or ES cells, and from either CRG or UCSC. The simulation
+in either B cells or ES cells, and from either CRG or UCSD. The simulation
 strategy was the following: First we selected uniformly at random the
@@
 total), we measured the pairwise reproducibility score defined by
-\citet{yan2017hicspector} and computed the average for pairs from the same
+\cite{yan2017hicspector} and computed the average for pairs from the same
 cell type minus the average for pairs from different cell types.
@@
 obtain a stratum-adjusted correlation coefficient (SCC) as defined by
-\citet{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
-\citet{yan2017hicspector} sums the distances between the leading twenty
+\cite{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
+\cite{yan2017hicspector} sums the distances between the leading twenty
 eigenvectors of the Laplacian of the Hi-C matrix. This approach borrows
@@
 four copies, explaining that the signal is about twice higher than for
-the short arm.}
+the short arm. Both figures present data at 100~Kbp resolution}
 \label{fig:totals}
@@
 For consistency and concision, the results based on the reproducibility
-score of \citet{yan2017hicspector} are shown in the main figures, and the
+score of \cite{yan2017hicspector} are shown in the main figures, and the
 results based on the other metrics are shown in the Supplementary
@@
 these cells have diverged, so that the copy number correction performed by
-\textit{OneD} increased the reproducibility between samples from the same
+\textit{OneD+CN} increased the reproducibility between samples from the same
 cell type.
@@
 \textit{LGF} strengthened the central region and the diagonal.
-\textit{OneD} reduced the level of the central portion by a factor 2
+\textit{OneD+CN} reduced the level of the central portion by a factor 2
 approximately, but did not otherwise distort the main features of the
@@
 normalization with \textit{OneD} than with matrix-balancing methods
-(figure~\ref{fig:chic}A). The performance of \textit{OneD} was only
-slightly lower than the raw data, suggesting that it barely disturbs the
-biological signal. Other tools were not included beause they had
-prohibitive run time.
+(figure~\ref{fig:chic}A and Supplementary Figure~13). The performance
+of \textit{OneD} was only slightly lower than the raw data, suggesting
+that it barely disturbs the biological signal. Other tools were not
+included beause they had prohibitive run time.
 
@@
 Figure~\ref{fig:times}A shows the running times of the different methods
-on all the samples shown in Supplementary Figure~1 at 100~Kbp resolution.
+on all the samples shown in Supplementary Table~1 at 100~Kbp resolution.
 The fastest method was \textit{vanilla} and the slowest \textit{LGF}, with
@@
 (HIC070, see Supplementary Table~1) at different bin sizes while
-downsampling the contact matrices to mimic lower resolutions. On this
+downsampling the contact matrices to mimic lower depths. On this
 test, we included only the methods competing with \textit{OneD} in terms
@@
 nearly identical results at very different sequencing depths
-(Supplementary Figure~13). \textit{vanilla} was the overall fastest method
+(Supplementary Figure~14). \textit{vanilla} was the overall fastest method
 and \textit{OneD} was usually faster than \textit{ICE} and \textit{KR}

### 2018-01-08 supplementary_figures_nar.tex
Answering R2 reviewer's comments

@@
 	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_13.eps}}
+    \caption{Bias removal on Capture Hi-C data A and C. Performance on the Capture
+    Hi-C data. Two replicates centered on a 6~Mbp domain on chromosome 6 in
+    T47D cells were normalized by the indicated tools and their
+    Pearson (A) or Spearman (C) correlation computed. B and D. Performance on virtual
+    Capture Hi-C  data. Two Hi-C experiments were post-processed to produce an output
+    similar to the Capture Hi-C experiment depicted in panels A and B. The signal
+    outside the domain on chromosome 6 was removed and the data were
+    normalized (capt), or the data were normalized before removing the signal
+    outside the domain on chromosome 6 (full). Four matrices per method were
+    generated and the Pearson (A) or Spearman (C) correlation computed between the
+    pairs from different replicates were computed. Note that the scores are identical
+    for the raw data because in this case the matrices of the same replicate are
+    identical.}
+\end{figure}
+
+\begin{figure}
+	\centerline{\includegraphics[width=\textwidth]{nar_figures/supp_figure_14.eps}}
     \caption{Correlation between subsamples at different resolutions and

### 2018-01-12 manuscript_nar.tex
More changes in R2

@@
 % Enter dates of publication
-\copyrightyear{2017}
+\copyrightyear{2018}
 \pubdate{** ***** ****}
-\pubyear{2017}
+\pubyear{2018}
 \jvolume{**}
@@
 in terms of computing resources for resolutions up to 5~Kbp.
-%\textit{OneD} is implemented as an R package
-%available at http://www.github.com/qenvio/dryhic.
+\textit{OneD} is implemented as an R package available at
+http://www.github.com/qenvio/dryhic.
 \end{abstract}
@@
 \texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
-\citep{wu2016computational} were used to carry out the \textit{LGF} and
+\citep{wu2016computational} were used to carry out the \textit{HiCNorm} and
 \textit{caICB} corrections respectively. When not specified, the
@@
 of contacts per bin in T47D. B. Total number of contacts per bin on
-chromosome 17 of T47D. The purple line represents the raw signal, the dark blue
+chromosome 9 of T47D. The purple line represents the raw signal, the dark blue
 line represents the signal after OneD bias correction, the light blue line
 represents the signal after OneD bias and copy number (CN) correction and
-the orange line represents the signal after ICE bias correction. The long arm
-of chromosome 17 (the region corresponding to 20-80 Mbp) is present in
-four copies, explaining that the signal is about twice higher than for
-the short arm. Both figures present data at 100~Kbp resolution}
+the orange line represents the signal after ICE bias correction. The grey
+line represents the copy number estimation by independent CNV array and it
+has been scaled to match the figure ranges. Some regions of the long arm of
+chromosome 9 (the region corresponding to 60-140 Mbp) are present in four
+copies, explaining that the signal is about twice higher than for the short
+arm. Both figures present data at 100~Kbp resolution}
 \label{fig:totals}
@@
 \citep{knight2013fast}, \textit{caICB} \citep{wu2016computational} and the
-Local Genomic Features method \citep[\textit{LGF},][]{hu2012hicnorm,
+Local Genomic Features method \citep[\textit{HiCNorm},][]{hu2012hicnorm,
 servant2012hitc}. The first four methods correct biases implicitly,
@@
 \textit{KR} gave inferior results, while \textit{raw+CN} (copy number
-correction alone), \textit{LGF} and \textit{vanilla} were not competitive.
+correction alone), \textit{HiCNorm} and \textit{vanilla} were not competitive.
 The conclusions were unchanged when using other metrics to compare Hi-C
@@
 \textit{OneD}, \textit{OneD+CN}, and \textit{caICB} had competitive
-performance. \textit{vanilla} and \textit{LGF} were less competitive, and
+performance. \textit{vanilla} and \textit{HiCNorm} were less competitive, and
 \textit{raw+CN} (copy number correction alone) performed poorly, because
@@
 capacity of the correction methods to distinguish different cell types
-(Supplementary Figure~12). On this test, \textit{LGF} and \textit{OneD}
+(Supplementary Figure~12). On this test, \textit{HiCNorm} and \textit{OneD}
 performed best. In comparison, \textit{OneD+CN} performed less well,
@@
 has an increased copy number, which affects normalization. \textit{ICE}
-fades it away, \textit{LGF} enhances it and \textit{OneD} reduces the
+fades it away, \textit{HiCNorm} enhances it and \textit{OneD} reduces the
 signal by about half.}
@@
 structure that was not visible in the raw data. On the contrary,
-\textit{LGF} strengthened the central region and the diagonal.
+\textit{HiCNorm} strengthened the central region and the diagonal.
 \textit{OneD+CN} reduced the level of the central portion by a factor 2
@@
 on all the samples shown in Supplementary Table~1 at 100~Kbp resolution.
-The fastest method was \textit{vanilla} and the slowest \textit{LGF}, with
+The fastest method was \textit{vanilla} and the slowest \textit{HiCNorm}, with
 an over 100-fold span between the two.

### 2018-01-14 manuscript_nar.tex
More changes due to R2

@@
 capacity of the correction methods to distinguish different cell types
-(Supplementary Figure~12). On this test, \textit{HiCNorm} and \textit{OneD}
+(Supplementary Figures~12 and 13). On this test, \textit{HiCNorm} and \textit{OneD}
 performed best. In comparison, \textit{OneD+CN} performed less well,
@@
 normalization with \textit{OneD} than with matrix-balancing methods
-(figure~\ref{fig:chic}A and Supplementary Figure~13). The performance
+(figure~\ref{fig:chic}A and Supplementary Figure~14). The performance
 of \textit{OneD} was only slightly lower than the raw data, suggesting
@@
 nearly identical results at very different sequencing depths
-(Supplementary Figure~14). \textit{vanilla} was the overall fastest method
+(Supplementary Figure~15). \textit{vanilla} was the overall fastest method
 and \textit{OneD} was usually faster than \textit{ICE} and \textit{KR}

### 2018-01-16 manuscript_nar.tex
Minor edits finalizing revision R2.

@@
 in terms of computing resources for resolutions up to 5~Kbp.
-\textit{OneD} is implemented as an R package available at
-http://www.github.com/qenvio/dryhic.
+The \textit{OneD} R package and the data used here are
+available at http://www.github.com/qenvio/dryhic.
 \end{abstract}
@@
 GSM987817 and GSM862720, see Supplementary Table~1), that were performed
-in either B cells or ES cells, and from either CRG or UCSD. The simulation
+in either B cells or ES cells, and from either CRG (Center for Genomic
+Regulation) or UCSD (University of California San Diego). The simulation
 strategy was the following: First we selected uniformly at random the
@@
 increase the reproducibility (Figure~\ref{fig:cnv_correction}A).
+If the purpose is to remove the effect of copy number, then OneD+CN
+outperforms a karyotype correction method such as caICB because it allows
+for more flexible types of aberrations. On the other hand, if the purpose
+is to improve the reproducibility among samples of the same type,
+correcting the copy number removes part of the signal, and therefore blurs
+the distinction between experiments (Figures~\ref{fig:aberrant} and
+\ref{fig:normal}).
 

### 2018-01-16 manuscript_nar.tex
Updating references and sources

@@
 
-We gathered a set of published Hi-C data \citep[][BioRxiv:
-https://doi.org/10.1101/132456]{ledily2014distinct, encode2012integrated,
-rao20143d, lin2012global, dixon2012topological} and unpublished data of
-different cell types and organisms (Supplementary Table~1).
+We gathered a set of published Hi-C data
+\citep{stadhouders2018transcription,ledily2014distinct,encode2012integrated,rao20143d,lin2012global,dixon2012topological}
+and unpublished data of different cell types and organisms (Supplementary
+Table~1).
 
@@
 All data were processed through a pipeline based on TADbit
-\citep{serra2016structural}. Briefly, after controlling the quality of
-FASTQ files, paired-end reads were mapped to the corresponding reference
-genome (hg38 or mm10) taking into account the restriction enzyme site.
-Non-informative contacts were removed applying the following TADbit
-filters: \texttt{self-circle}, \texttt{dangling-ends}, \texttt{error},
-\texttt{extra-dangling-ends}, \texttt{duplicated} and
-\texttt{random-breaks} (for more details see BioRxiv:
-https://doi.org/10.1101/132456). In addition, the pipeline is available
-from the Supplementary Material published by \cite{quilez2017managing}.
+\citep{serra2016structural}. Briefly, after controlling the quality of FASTQ
+files, paired-end reads were mapped to the corresponding reference genome (hg38
+or mm10) taking into account the restriction enzyme site.  Non-informative
+contacts were removed applying the following TADbit filters:
+\texttt{self-circle}, \texttt{dangling-ends}, \texttt{error},
+\texttt{extra-dangling-ends}, \texttt{duplicated} and \texttt{random-breaks}
+(for more details see \cite{stadhouders2018transcription}). In addition, the
+pipeline is available from the Supplementary Material published by
+\cite{quilez2017managing}.
 
